
# Community composition

## Gut microbiota

```{r gut_load_data_tax}
load("data/data.Rdata")
```

### Taxonomy overview 

#### Phylum level

```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
          strip.text = element_text(size = 12, lineheight = 0.6,face="bold"),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

**Number of MAGs**
```{r mag, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nmags <- nrow(genome_counts)
cat(nmags)
```

**Number of bacteria phyla**

```{r phyla, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Number of Archaea phyla**

```{r arch, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

***Phylum relative abundances***

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,broad_environment, sex) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary_envir, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = sd(relabun * 100, na.rm = TRUE)
  ) %>%
  mutate(Total = str_c(round(total_mean, 3), "±", round(total_sd, 3))) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, Total)

```

```{r taxonomy_jitterplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    left_join(genome_metadata %>% select(phylum,phylum) %>% unique(),by=join_by(phylum==phylum)) %>%
#    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    filter(phylum %in% phylum_arrange[1:20]) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) +
        theme_minimal() + 
        labs(y="phylum", x="Relative abundance", color="Phylum")
```
**Bacteria phyla in individuals from females**

```{r phyla_nat_female, comment="", echo=FALSE, message=FALSE, warning=FALSE}

female_sample <- sample_metadata %>% 
  filter(sex=="Female") %>% 
  dplyr::select(sample) %>% 
  pull()

female_genomes <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  select(all_of(female_sample)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% female_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

```{r phyla_nat_male, comment="", echo=FALSE, message=FALSE, warning=FALSE}

male_sample <- sample_metadata %>% 
  filter(sex=="Male") %>% 
  dplyr::select(sample) %>% 
  pull()

male_genomes <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  select(all_of(male_sample)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% male_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```



**Bacteria phyla in individuals from Temperate woodland**

```{r phyla_nat, comment="", echo=FALSE, message=FALSE, warning=FALSE}

Temperate_woodland_samples <- sample_metadata %>% 
  filter(broad_environment=="1000221 - Temperate woodland") %>% 
  dplyr::select(sample) %>% 
  pull()

Temperate_woodland_genomes <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  select(all_of(Temperate_woodland_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% Temperate_woodland_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```


**Bacteria phyla in individuals from Xeric shrubland**

```{r phyla_cap_xeric, comment="", echo=FALSE, message=FALSE, warning=FALSE}

xeric_shrubland_samples <- sample_metadata %>% 
  filter(broad_environment=="1000218 - Xeric shrubland") %>% 
  dplyr::select(sample) %>% 
  pull()

xeric_shrubland_genomes <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  select(all_of(xeric_shrubland_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% xeric_shrubland_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```
**Bacteria phyla in individuals from mixed_forest**
```{r phyla_cap_forest, comment="", echo=FALSE, message=FALSE, warning=FALSE}

mixed_forest_samples <- sample_metadata %>% 
  filter(broad_environment=="1000198 - Mixed forest") %>% 
  dplyr::select(sample) %>% 
  pull()

mixed_forest_genomes <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  select(all_of(mixed_forest_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% mixed_forest_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Bacteria phyla in individuals from cropland**
```{r phyla_cap_cropland, comment="", echo=FALSE, message=FALSE, warning=FALSE}

cropland_samples <- sample_metadata %>% 
  filter(broad_environment=="1000245 - Cropland") %>% 
  dplyr::select(sample) %>% 
  pull()

cropland_genomes <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  select(all_of(cropland_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% cropland_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

```{r taxonomy_phylum_summary_broad_environment, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = sd(relabun * 100, na.rm = TRUE),
    le_mean = mean(ifelse(broad_environment == "1000245 - Cropland", relabun * 100, NA), na.rm = TRUE),
    le_sd = sd(ifelse(broad_environment == "1000245 - Cropland", relabun * 100, NA), na.rm = TRUE),
    ha_mean = mean(ifelse(broad_environment == "1000198 - Mixed forest", relabun * 100, NA), na.rm = TRUE),
    ha_sd = sd(ifelse(broad_environment == "1000198 - Mixed forest", relabun * 100, NA), na.rm = TRUE),
    er_mean = mean(ifelse(broad_environment == "1000218 - Xeric shrubland", relabun * 100, NA), na.rm = TRUE),
    er_sd = sd(ifelse(broad_environment == "1000218 - Xeric shrubland", relabun * 100, NA), na.rm = TRUE),
    go_mean = mean(ifelse(broad_environment == "1000221 - Temperate woodland", relabun * 100, NA), na.rm = TRUE),
    go_sd = sd(ifelse(broad_environment == "1000221 - Temperate woodland", relabun * 100, NA), na.rm = TRUE)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Cropland = str_c(round(le_mean, 3), "±", round(le_sd, 3)),
    Mixed_forest = str_c(round(ha_mean, 3), "±", round(ha_sd, 3)),
    Xeric_shrubland = str_c(round(er_mean, 3), "±", round(er_sd, 3)),
    Temperate_woodland = str_c(round(go_mean, 3), "±", round(go_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  select(phylum, Total, Cropland, Temperate_woodland, Mixed_forest, Xeric_shrubland)

```

```{r taxonomy_phylum_summary_sex, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = sd(relabun * 100, na.rm = TRUE),
    le_mean = mean(ifelse(sex == "Female", relabun * 100, NA), na.rm = TRUE),
    le_sd = sd(ifelse(sex == "Female", relabun * 100, NA), na.rm = TRUE),
    ha_mean = mean(ifelse( sex== "Male", relabun * 100, NA), na.rm = TRUE),
    ha_sd = sd(ifelse(sex == "Male", relabun * 100, NA), na.rm = TRUE)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Female = str_c(round(le_mean, 3), "±", round(le_sd, 3)),
    Male = str_c(round(ha_mean, 3), "±", round(ha_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  select(phylum, Total, Female, Male)

```

#### Family level

***Percentange of families in each group of environment***

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family, broad_environment, sex) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_family_summary_origin_environment, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
    summarise(
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = sd(relabun * 100, na.rm = TRUE),
    le_mean = mean(ifelse(broad_environment == "1000245 - Cropland", relabun * 100, NA), na.rm = TRUE),
    le_sd = sd(ifelse(broad_environment == "1000245 - Cropland", relabun * 100, NA), na.rm = TRUE),
    ha_mean = mean(ifelse(broad_environment == "1000198 - Mixed forest", relabun * 100, NA), na.rm = TRUE),
    ha_sd = sd(ifelse(broad_environment == "1000198 - Mixed forest", relabun * 100, NA), na.rm = TRUE),
    er_mean = mean(ifelse(broad_environment == "1000218 - Xeric shrubland", relabun * 100, NA), na.rm = TRUE),
    er_sd = sd(ifelse(broad_environment == "1000218 - Xeric shrubland", relabun * 100, NA), na.rm = TRUE),
    go_mean = mean(ifelse(broad_environment == "1000221 - Temperate woodland", relabun * 100, NA), na.rm = TRUE),
    go_sd = sd(ifelse(broad_environment == "1000221 - Temperate woodland", relabun * 100, NA), na.rm = TRUE)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Cropland = str_c(round(le_mean, 3), "±", round(le_sd, 3)),
    Mixed_forest = str_c(round(ha_mean, 3), "±", round(ha_sd, 3)),
    Xeric_shrubland = str_c(round(er_mean, 3), "±", round(er_sd, 3)),
    Temperate_woodland = str_c(round(go_mean, 3), "±", round(go_sd, 3))
  ) %>%
    dplyr::select(family,Total,Cropland, Temperate_woodland, Mixed_forest, Xeric_shrubland)
```

```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    pull()

# Per environment
family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~broad_environment)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")

```

```{r taxonomy_family_summary_broad_environment, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              le_mean=mean(relabun[broad_environment=="1000245 - Cropland"]*100, na.rm=T),
             le_sd=sd(relabun[broad_environment=="1000245 - Cropland"]*100, na.rm=T),
              ha_mean=mean(relabun[broad_environment=="1000198 - Mixed forest"]*100, na.rm=T),
              ha_sd=sd(relabun[broad_environment=="1000198 - Mixed forest"]*100, na.rm=T),
              er_mean=mean(relabun[broad_environment=="1000218 - Xeric shrubland"]*100, na.rm=T),
              er_sd=sd(relabun[broad_environment=="1000218 - Xeric shrubland"]*100, na.rm=T),
              go_mean=mean(relabun[broad_environment=="1000221 - Temperate woodland"]*100, na.rm=T),
              go_sd=sd(relabun[broad_environment=="1000221 - Temperate woodland"]*100, na.rm=T)) %>%
    mutate(Total=str_c(round(total_mean,3),"±",round(total_sd,3)),
          Cropland=str_c(round(le_mean,3),"±",round(le_sd,3)),
          Mixed_forest=str_c(round(ha_mean,3),"±",round(ha_sd,3)),
          Xeric_shrubland=str_c(round(er_mean,3),"±",round(er_sd,3)),
          Temperate_woodland=str_c(round(go_mean,3),"±",round(go_sd,3))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(family,Total,Cropland,Temperate_woodland, Mixed_forest,Xeric_shrubland)
```
***Family summary grouped by sex***

```{r taxonomy_family_summary_origin_sex, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
   summarise(
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = sd(relabun * 100, na.rm = TRUE),
    le_mean = mean(ifelse(sex == "Female", relabun * 100, NA), na.rm = TRUE),
    le_sd = sd(ifelse(sex == "Female", relabun * 100, NA), na.rm = TRUE),
    ha_mean = mean(ifelse( sex== "Male", relabun * 100, NA), na.rm = TRUE),
    ha_sd = sd(ifelse(sex == "Male", relabun * 100, NA), na.rm = TRUE)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Female = str_c(round(le_mean, 3), "±", round(le_sd, 3)),
    Male = str_c(round(ha_mean, 3), "±", round(ha_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  select(family, Total, Female, Male)
```
```{r taxonomy_jitterplot_family_sex, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    pull()

# Per environment
family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~sex)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")

```

```{r taxonomy_family_summary_sex, warning=FALSE, comments="", message=FALSE}
family_summary %>%
    group_by(family) %>%
    summarise(
    total_mean = mean(relabun * 100, na.rm = TRUE),
    total_sd = sd(relabun * 100, na.rm = TRUE),
    le_mean = mean(ifelse(sex == "Female", relabun * 100, NA), na.rm = TRUE),
    le_sd = sd(ifelse(sex == "Female", relabun * 100, NA), na.rm = TRUE),
    ha_mean = mean(ifelse( sex== "Male", relabun * 100, NA), na.rm = TRUE),
    ha_sd = sd(ifelse(sex == "Male", relabun * 100, NA), na.rm = TRUE)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Female = str_c(round(le_mean, 3), "±", round(le_sd, 3)),
    Male = str_c(round(ha_mean, 3), "±", round(ha_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
    dplyr::select(family,Total,Female, Male)
```
#### Genus level

*** Percetange of genera in each group of environment***

```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,phylum,genus,broad_environment) %>%
  summarise(relabun=sum(count)) 
# %>%
#   filter(genus != "g__") %>%
#   mutate(genus= sub("^g__", "", genus))

genus_summary %>%
    group_by(genus) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              le_mean=mean(relabun[broad_environment=="1000245 - Cropland"]*100, na.rm=T),
             le_sd=sd(relabun[broad_environment=="1000245 - Cropland"]*100, na.rm=T),
              ha_mean=mean(relabun[broad_environment=="1000198 - Mixed forest"]*100, na.rm=T),
              ha_sd=sd(relabun[broad_environment=="1000198 - Mixed forest"]*100, na.rm=T),
              er_mean=mean(relabun[broad_environment=="1000218 - Xeric shrubland"]*100, na.rm=T),
              er_sd=sd(relabun[broad_environment=="1000218 - Xeric shrubland"]*100, na.rm=T),
              go_mean=mean(relabun[broad_environment=="1000221 - Temperate woodland"]*100, na.rm=T),
              go_sd=sd(relabun[broad_environment=="1000221 - Temperate woodland"]*100, na.rm=T)) %>%
    mutate(Total=str_c(round(total_mean,3),"±",round(total_sd,3)),
          Cropland=str_c(round(le_mean,3),"±",round(le_sd,3)),
          Mixed_forest=str_c(round(ha_mean,3),"±",round(ha_sd,3)),
          Xeric_shrubland=str_c(round(er_mean,3),"±",round(er_sd,3)),
          Temperate_woodland=str_c(round(go_mean,3),"±",round(go_sd,3))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(genus,Total, Cropland,Temperate_woodland, Mixed_forest,Xeric_shrubland) 
```

```{r taxonomy_jitterplot_genus, fig.height=14, fig.width=10, fig.fullwidth=TRUE}
genus_summary_sort <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) 

genus_summary %>%
  mutate(genus=factor(genus, levels=rev(genus_summary_sort %>% pull(genus)))) %>%
  filter(relabun > 0) %>%
  ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
  scale_color_manual(values=phylum_colors) +
  geom_jitter(alpha=0.5) + 
  facet_grid(.~broad_environment)+
  theme_minimal() + 
  theme(axis.text.y = element_text(size=6))+
  labs(y="Family", x="Relative abundance", color="Phylum")

```
*** Percetange of genera in each group of sex***

```{r taxonomy_genus_summary_sex, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,phylum,genus,sex) %>%
  summarise(relabun=sum(count)) 
# %>%
#   filter(genus != "g__") %>%
#   mutate(genus= sub("^g__", "", genus))

genus_summary %>%
    group_by(genus) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              le_mean=mean(relabun[sex=="Female"]*100, na.rm=T),
             le_sd=sd(relabun[sex=="Female"]*100, na.rm=T),
              ha_mean=mean(relabun[sex=="Male"]*100, na.rm=T),
              ha_sd=sd(relabun[sex=="Male"]*100, na.rm=T)) %>%
    mutate(Total=str_c(round(total_mean,3),"±",round(total_sd,3)),
          Female=str_c(round(le_mean,3),"±",round(le_sd,3)),
          Male=str_c(round(ha_mean,3),"±",round(ha_sd,3))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(genus,Total,Male,Female) 
```

```{r taxonomy_jitterplot_genus_sex, fig.height=14, fig.width=10, fig.fullwidth=TRUE}
genus_summary_sort <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) 

genus_summary %>%
  mutate(genus=factor(genus, levels=rev(genus_summary_sort %>% pull(genus)))) %>%
  filter(relabun > 0) %>%
  ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
  scale_color_manual(values=phylum_colors) +
  geom_jitter(alpha=0.5) + 
  facet_grid(.~sex)+
  theme_minimal() + 
  theme(axis.text.y = element_text(size=6))+
  labs(y="Family", x="Relative abundance", color="Phylum")

```

**Number of MAGs without genera-level annotation**
```{r nongenera_gut, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(genus == "g__") %>%
  nrow() %>% 
  cat()
```

```{r nongene_phylum_gut, comment="", echo=FALSE, message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(genus == "g__") %>%
  group_by(phylum) %>%
  summarize(count_nogene = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage=100*count_nogene/count_total) %>% 
  tt()
```

**Percentage of MAGs without genus-level annotation**
```{r gen_percet, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
  filter(genus == "g__") %>%
  summarize(Mag_nogenera = n()) %>% 
  pull()
perct <- nongenera*100/nmags
cat(perct)
```


**Number of MAGs without species-level annotation**
```{r nonspe, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "s__") %>%
  summarize(Mag_nospecies = n())

```

```{r nonspe_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "s__") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(species_annotated=count_total-count_nospecies) %>% 
  mutate(percentage=100*count_nospecies/count_total) %>% 
  mutate(percentage_species=100-100*count_nospecies/count_total)%>% 
  tt()
```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nonspecies <- genome_metadata %>%
  filter(species == "s__") %>%
  summarize(Mag_nospecies = n()) %>% 
  pull()
perct <- nonspecies*100/nmags
cat(perct)
```
