
# Alpha diversity

We quantify within-sample diversity using Hill numbers: richness (q0), neutral/abundance-weighted (q1), phylogenetic (q1 + tree), and functional (q1 + traits), and compare these across environments and sex.

## microbiota

**Note:** Load all required data objects including genome counts, metadata, and color schemes. Set factor levels for broad_environment variable to ensure consistent ordering.

```{r load_data_alpha, message=FALSE, warning=FALSE}
load("data/data.Rdata")
treatment_colors <- c("#f56042","#429ef5", "#42f58d", "#b142f5", "#FFA500")
sample_metadata$broad_environment <- factor(sample_metadata$broad_environment, levels=c("1000198 - Mixed forest", "1000221 - Temperate woodland", "1000218 - Xeric shrubland", "1000215 - Temperate shrubland", "1000245 - Cropland"))
```

**Note:** Filter out sample EHI01340 and filter genome counts to only include valid samples.

```{r exclude_sample3, message=FALSE, warning=FALSE}

sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
valid_samples <- sample_metadata$sample

genome_counts_filt <- genome_counts_filt %>%
  dplyr::select(genome, tidyselect::any_of(valid_samples))

```

**Note:** Prune the phylogenetic tree to match genomes present in the filtered count data. Identifies genomes that exist in both the count data and the tree, then removes tree tips that don't have corresponding count data.

```{r pruning_tree, message=FALSE, warning=FALSE}
# 1) Figure out the common labels
otus  <- genome_counts_filt$genome
tips  <- genome_tree$tip.label
common <- intersect(tips, otus)

# 2) Prune the tree to only those tips
genome_tree_pruned <- drop.tip(genome_tree, setdiff(tips, common))
```

### Hill numbers
Richness of microbiotic species in samples

**Note:** Calculate four types of Hill number diversity metrics for each sample: richness (q=0, species count unweighted by abundance), neutral (q=1, Shannon diversity abundance-weighted), phylogenetic (q=1 + tree, incorporates phylogenetic relationships), and functional (q=1 + traits, incorporates functional trait distances). Then merge all metrics into a single alpha_div dataframe.

```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree_pruned) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% rownames(dist)) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```

**Note:** Create a summary statistics table showing mean ± SD for each alpha diversity metric. Calculates overall statistics and statistics by environment type, formats results as "mean±sd" strings, and displays results in a formatted table.

```{r alpha_div_diets_summary_all, comment="",echo=FALSE, message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  inner_join(sample_metadata, by = join_by(sample == sample)) %>%
    group_by(alpha)%>%
    summarise(total_mean=mean(value, na.rm=T),
              total_sd=sd(value, na.rm=T),
              Xeric_shrubland_mean=mean(value[broad_environment=="1000218 - Xeric shrubland"], na.rm=T),
             Xeric_shrubland_sd=sd(value[broad_environment=="1000218 - Xeric shrubland"], na.rm=T),
             Cropland_mean=mean(value[broad_environment=="1000245 - Cropland"], na.rm=T),
             Cropland_sd=sd(value[broad_environment=="1000245 - Cropland"], na.rm=T),
              Mixed_forest_mean=mean(value[broad_environment=="1000198 - Mixed forest"], na.rm=T),
             Mixed_forest_sd=sd(value[broad_environment=="1000198 - Mixed forest"], na.rm=T),
              Temperate_woodland_mean=mean(value[broad_environment=="1000221 - Temperate woodland"], na.rm=T),
              Temperate_woodland_sd=sd(value[broad_environment=="1000221 - Temperate woodland"], na.rm=T),
              Temperate_shrubland_mean=mean(value[broad_environment=="1000215 - Temperate shrubland"], na.rm=T),
              Temperate_shrubland_sd=sd(value[broad_environment=="1000215 - Temperate shrubland"], na.rm=T)) %>%
    mutate(Total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Xeric_shrubland=str_c(round(Xeric_shrubland_mean,2),"±",round(Xeric_shrubland_sd,2)),
           Cropland=str_c(round(Cropland_mean,2),"±",round(Cropland_sd,2)),
           Mixed_forest=str_c(round(Mixed_forest_mean,2),"±",round(Mixed_forest_sd,2)),
           Temperate_woodland=str_c(round(Temperate_woodland_mean,2),"±",round(Temperate_woodland_sd,2)),
           Temperate_shrubland=str_c(round(Temperate_shrubland_mean,2),"±",round(Temperate_shrubland_sd,2))) %>% 
    arrange(-total_mean) %>% 
    dplyr::select(alpha,Total,Xeric_shrubland,Temperate_woodland,Temperate_shrubland,Cropland,Mixed_forest) %>% 
  tt()
```

Alpha-diversity across environments: richness (q0), neutral (q1), phylogenetic (q1 + tree), and functional (q1 + traits); jittered points show samples, boxes show group distributions; Wilcoxon p-values annotated.

Richness of microbiotic species - environment clusters

**Note:** Create boxplots comparing alpha diversity metrics across environments. Defines environment labels (short and long forms) and color mappings, creates a reusable plotting function for each metric, generates four plots (richness, neutral, phylogenetic, functional), combines plots into a 2x2 grid with shared legend, and includes Wilcoxon test p-values on each plot.

```{r alpha_div_boxplot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=16, fig.fullwidth=TRUE}
#Richness

env_limits <- c(
  "1000198 - Mixed forest",
  "1000221 - Temperate woodland",
  "1000218 - Xeric shrubland",
  "1000215 - Temperate shrubland",
  "1000245 - Cropland"
)

env_labels_short <- c("MF","TW","XS","TS","C")
env_labels_long <- c(
  "1000198 - Mixed forest"        = "Mixed forest (MF)",
  "1000221 - Temperate woodland"  = "Temperate woodland (TW)",
  "1000218 - Xeric shrubland"     = "Xeric shrubland (XS)",
  "1000215 - Temperate shrubland" = "Temperate shrubland (TS)",
  "1000245 - Cropland"            = "Cropland (C)"
)

make_plot <- function(target_metric, ylab){
  alpha_div %>%
    pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
    inner_join(sample_metadata, by = dplyr::join_by(sample == sample)) %>%
    filter(metric == target_metric) %>%
    ggplot(aes(x = broad_environment, y = value)) +
    geom_boxplot(aes(fill = broad_environment),
                 width = 0.65, alpha = 0.75, outlier.shape = NA,
                 show.legend = TRUE, linewidth = 0.6) +
    geom_jitter(aes(color = broad_environment),
                width = 0.3, size = 1.8, alpha = 0.65,
                show.legend = FALSE) +
    scale_fill_manual(values = treatment_colors, labels = env_labels_long,
                      breaks = env_limits, drop = FALSE) +
    scale_color_manual(values = treatment_colors, guide = "none") +
    scale_x_discrete(limits = env_limits, labels = env_labels_short, drop = FALSE) +
    # p-value: white label at the very top of the panel
    stat_compare_means(method = "wilcox.test", label = "p.format",
                       geom = "label", fill = "white", label.size = 0.4,
                       label.padding = unit(0.2, "lines"),
                       label.r = unit(0.15, "lines"),
                       color = "black",
                       label.y.npc = 0.98, label.x = 3, size = 4, fontface = "bold") +
    scale_y_continuous(expand = expansion(mult = c(0.08, 0.12))) +
    coord_cartesian(clip = "off") +
    theme_classic() +
    theme(
      legend.title = element_blank(),
      legend.text = element_text(size = 11),
      legend.position = "top",
      legend.margin = margin(b = 6, t = 0),
      legend.key.size = unit(0.8, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.box.margin = margin(b = 5),
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, 
                                 margin = margin(t = 8)),
      axis.text.y = element_text(size = 12),
      axis.title.y = element_text(size = 13, margin = margin(r = 20), face = "bold"),
      plot.margin = margin(t = 8, r = 12, b = 10, l = 12),
      panel.spacing = unit(1.2, "lines")
    ) +
    labs(y = ylab)
}

plot1 <- make_plot("richness",      "Richness")
plot2 <- make_plot("neutral",       "Neutral alpha diversity")
plot3 <- make_plot("phylogenetic",  "Phylogenetic alpha diversity")
plot4 <- make_plot("functional",    "Functional alpha diversity")

# >>> Plots without legends <<<
# Remove legends from individual plots
plot1_no_legend <- plot1 + theme(legend.position = "none")
plot2_no_legend <- plot2 + theme(legend.position = "none")
plot3_no_legend <- plot3 + theme(legend.position = "none")
plot4_no_legend <- plot4 + theme(legend.position = "none")

# Combine plots without legends
plots_combined <- (plot1_no_legend + plot2_no_legend + plot3_no_legend + plot4_no_legend) +
  plot_layout(ncol = 2, widths = c(1, 1), heights = c(1, 1))

# Create the legend
legend_data <- data.frame(
  environment = factor(env_limits, levels = env_limits),
  x = 1:5,
  y = 1
)

alpha_div_legend <- ggplot(legend_data, aes(x = x, y = y, fill = environment)) +
  geom_point(shape = 22, size = 8, show.legend = TRUE) +
  scale_fill_manual(
    values = treatment_colors,
    labels = env_labels_long,
    breaks = env_limits,
    drop = FALSE,
    name = NULL
  ) +
  theme_void() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "center",
    legend.text = element_text(size = 13),
    legend.key.size = unit(1.5, "lines"),
    legend.spacing.x = unit(0.8, "cm"),
    legend.margin = margin(b = 15, t = 10),
    plot.margin = margin(t = 10, b = 5, l = 0, r = 0)
  ) +
  guides(fill = guide_legend(override.aes = list(size = 5)))

# Combine legend on top with plots below
final_plot <- cowplot::plot_grid(
  alpha_div_legend,
  plots_combined,
  ncol = 1,
  rel_heights = c(0.15, 1),
  align = "v"
)

final_plot
```


Richness of microbiotic species - sex clusters

**Note:** Create boxplots comparing alpha diversity metrics by sex, faceted by environment. Defines color scheme for sex categories, creates a plotting function that facets by environment, generates four plots (richness, neutral, phylogenetic, functional), combines plots into a 2x2 grid with shared legend, and includes Wilcoxon test p-values comparing Female vs Male within each environment.

```{r alpha_div_boxplot_sex, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
# colors (keep or replace)
sex_colors <- c("Female" = "#E41A1C", "Male" = "#377EB8", "Unknown" = "grey50")

make_plot_sex <- function(target_metric, ylab) {
  alpha_div %>%
    pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
    inner_join(sample_metadata, by = dplyr::join_by(sample == sample)) %>%
    filter(metric == target_metric) %>%
    mutate(
      # ensure order + map short ticks
      sex = factor(sex, levels = c("Female","Male","Unknown"))
    ) %>%
    ggplot(aes(x = sex, y = value, fill = sex)) +
    geom_boxplot(width = 0.6, alpha = 0.7, outlier.shape = NA, 
                 show.legend = TRUE, linewidth = 0.5) +
    geom_jitter(aes(color = sex), width = 0.2, size = 1.3, alpha = 0.7, 
                show.legend = FALSE) +
    facet_grid(
      ~ broad_environment,
      labeller = labeller(broad_environment = c(
        "1000198 - Mixed forest"       = "MF",
        "1000221 - Temperate woodland" = "TW",
        "1000218 - Xeric shrubland"    = "XS",
        "1000215 - Temperate shrubland" = "TS",
        "1000245 - Cropland"           = "C"
      )),
      scales = "fixed", space = "free_x"
    ) +
    # p-values: readable white labels at the top of each facet
    ggpubr::stat_compare_means(
      method = "wilcox.test",
      comparisons = list(c("Female","Male")),
      label = "p.format",
      geom = "label", fill = "white", color = "black",
      label.size = 0.3, label.padding = grid::unit(0.15, "lines"),
      label.r = grid::unit(0.1, "lines"),
      label.y.npc = 0.98, label.x.npc = 0.5, size = 3.5, fontface = "bold"
    ) +
    # SHORT x-ticks + legend with full names
    scale_x_discrete(labels = c(Female = "F", Male = "M", Unknown = "U")) +
    scale_fill_manual(
      values = sex_colors,
      labels = c("Female (F)", "Male (M)", "Unknown (U)")
    ) +
    scale_color_manual(values = sex_colors, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    coord_cartesian(clip = "off") +
    theme_classic() +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 11),
      legend.margin = margin(b = 8),
      strip.background = element_rect(fill = "grey95", color = "white"),
      strip.text = element_text(size = 12, face = "bold", margin = margin(4, 0, 4, 0)),
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 12, margin = margin(r = 15)),
      plot.margin = margin(t = 10, r = 8, b = 8, l = 8),
      plot.caption = element_text(size = 9, hjust = 0, margin = margin(t = 5))
    ) +
    labs(
      y = ylab,
      caption = "Axis codes: F=Female; M=Male; U=Unknown."
    )
}

plot1 <- make_plot_sex("richness",      "Richness")
plot2 <- make_plot_sex("neutral",       "Neutral alpha diversity")
plot3 <- make_plot_sex("phylogenetic",  "Phylogenetic alpha diversity")
plot4 <- make_plot_sex("functional",    "Functional alpha diversity")

# Remove legends from individual plots
plot1_no_legend <- plot1 + theme(legend.position = "none")
plot2_no_legend <- plot2 + theme(legend.position = "none")
plot3_no_legend <- plot3 + theme(legend.position = "none")
plot4_no_legend <- plot4 + theme(legend.position = "none")

# Combine plots without legends
plots_combined_sex <- (plot1_no_legend + plot2_no_legend + plot3_no_legend + plot4_no_legend) +
  plot_layout(ncol = 2, widths = c(1, 1), heights = c(1, 1))

# Create the legend for sex categories
legend_data_sex <- data.frame(
  sex = factor(c("Female", "Male", "Unknown"), levels = c("Female", "Male", "Unknown")),
  x = 1:3,
  y = 1
)

alpha_div_legend_sex <- ggplot(legend_data_sex, aes(x = x, y = y, fill = sex)) +
  geom_point(shape = 22, size = 8, show.legend = TRUE) +
  scale_fill_manual(
    values = sex_colors,
    labels = c("Female (F)", "Male (M)", "Unknown (U)"),
    breaks = c("Female", "Male", "Unknown"),
    name = NULL
  ) +
  theme_void() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "center",
    legend.text = element_text(size = 13),
    legend.key.size = unit(1.5, "lines"),
    legend.spacing.x = unit(0.8, "cm"),
    legend.margin = margin(b = 15, t = 10),
    plot.margin = margin(t = 10, b = 5, l = 0, r = 0)
  ) +
  guides(fill = guide_legend(override.aes = list(size = 5)))

# Combine legend on top with plots below
final_plot_sex <- cowplot::plot_grid(
  alpha_div_legend_sex,
  plots_combined_sex,
  ncol = 1,
  rel_heights = c(0.15, 1),
  align = "v"
)

final_plot_sex

```

### Mixed models

**Note:** Fit mixed-effects and linear models for richness metric. Merges alpha diversity data with sample metadata, filters to complete cases (removes NAs), fits three models: environment with sex as random effect, environment+sex as fixed effects, and sex with environment as random effect. Performs pairwise comparisons using emmeans.

```{r rich_MM, comment="", message=FALSE, warning=FALSE}

alpha_div_meta <- alpha_div %>%
  inner_join(sample_metadata, by = join_by(sample == sample))

alpha_div_meta_clean <- alpha_div_meta %>%
  filter(!is.na(richness), !is.na(broad_environment), !is.na(sex))

alpha_div_meta_filtered_out <- anti_join(alpha_div_meta, alpha_div_meta_clean)

Model_richness_random <- lme(fixed = richness ~ broad_environment, data = alpha_div_meta_clean,
               random = ~ 1 | sex)
summary(Model_richness_random)

Model_richness<- lm(formula = richness ~ broad_environment+sex, data = alpha_div_meta_clean) 
summary(Model_richness)

Model_richness_sex_random <- lme(fixed = richness ~ sex, data = alpha_div_meta_clean,
               random = ~ 1 | broad_environment)
summary(Model_richness_sex_random)
emmeans(Model_richness_sex_random, pairwise ~ sex)
```

**Note:** Fit mixed-effects and linear models for neutral diversity metric. Uses the same model structure as richness models but with neutral as response variable. Fits three models: environment with sex as random effect, environment+sex as fixed effects, and sex with environment as random effect. Performs pairwise comparisons using emmeans.

```{r neutral_MM, comment="", message=FALSE, warning=FALSE}
Model_neutral_random <- lme(fixed = neutral ~ broad_environment, data = alpha_div_meta_clean,
               random = ~ 1 | sex)
summary(Model_neutral_random)

Model_neutral<- lm(formula = neutral ~ broad_environment+sex, data = alpha_div_meta_clean) 
summary(Model_neutral)

Model_neutral_sex_random <- lme(fixed = neutral ~ sex, data = alpha_div_meta_clean,
               random = ~ 1 | broad_environment)
summary(Model_neutral_sex_random)
emmeans(Model_neutral_sex_random, pairwise ~ sex)
```

**Note:** Fit mixed-effects and linear models for phylogenetic diversity metric. Uses the same model structure as richness models but with phylogenetic as response variable. Fits three models: environment with sex as random effect, environment+sex as fixed effects, and sex with environment as random effect. Performs pairwise comparisons using emmeans.

```{r phylo_MM, comment="", message=FALSE, warning=FALSE}

Model_phylogenetic_random <- lme(fixed = phylogenetic ~ broad_environment, data = alpha_div_meta_clean,
               random = ~ 1 | sex)
summary(Model_phylogenetic_random)

Model_phylogenetic<- lm(formula = phylogenetic ~ broad_environment+sex, data = alpha_div_meta_clean) 
summary(Model_phylogenetic)

Model_phylogenetic_sex_random <- lme(fixed = phylogenetic ~ sex, data = alpha_div_meta_clean,
               random = ~ 1 | broad_environment)
summary(Model_phylogenetic_sex_random)
emmeans(Model_phylogenetic_sex_random, pairwise ~ sex)

```

**Note:** Fit mixed-effects and linear models for functional diversity metric. Uses the same model structure as richness models but with functional as response variable. Fits three models: environment with sex as random effect, environment+sex as fixed effects, and sex with environment as random effect. Performs pairwise comparisons using emmeans.

```{r funct_MM, comment="", message=FALSE, warning=FALSE}
Model_functional_random <- lme(fixed = functional ~ broad_environment, data = alpha_div_meta_clean,
               random = ~ 1 | sex)
summary(Model_functional_random)

Model_functional<- lm(formula = functional ~ broad_environment+sex, data = alpha_div_meta_clean) 
summary(Model_functional)

Model_functional_sex_random <- lme(fixed = functional ~ sex, data = alpha_div_meta_clean,
               random = ~ 1 | broad_environment)
summary(Model_functional_sex_random)
emmeans(Model_functional_sex_random, pairwise ~ sex)
```
