
# Environment types

Environmental context around each sample: we extract land-cover within 100/200/400 m buffers using 200m as the primary buffer zone for the main analysis (using the QML palette to map classes), group them into interpretable categories, normalise per sample, and plot stacked compositions; selected bioclim layers are prepared for downstream use.

## Landcover percentage
**Generate Bufferzones**
```{r extract-buffer, message=FALSE, warning=FALSE}
# PROJ database (required before terra reads CRS on macOS)
if (!nzchar(Sys.getenv("PROJ_LIB"))) {
  sf_pkg <- tryCatch(find.package("sf", quiet = TRUE), error = function(e) character())
  if (length(sf_pkg)) {
    proj_lib <- file.path(sf_pkg, "proj")
    if (file.exists(file.path(proj_lib, "proj.db"))) {
      Sys.setenv(PROJ_LIB = proj_lib)
    }
  }
}

# Read once, filter once
sample_points <- read.csv("sample_points.csv", stringsAsFactors = FALSE) |>
  dplyr::filter(id != "EHI01340")

# WGS84 points (source CRS)
points_sf <- st_as_sf(sample_points, coords = c("longitude", "latitude"), crs = 4326)

# --- Build a virtual mosaic (VRT) to save RAM ---
tile_files <- list.files("Landcover/Tif", pattern = "\\.tif$", full.names = TRUE)
stopifnot(length(tile_files) > 0)

mosaic_vrt <- terra::vrt(tile_files)
lc_crs     <- terra::crs(mosaic_vrt, proj = TRUE)
if (!nzchar(lc_crs)) {
  # GA landcover tiles: GDA94 / Australian Albers (EPSG:3577)
  lc_crs <- "+proj=aea +lat_0=0 +lon_0=132 +lat_1=-18 +lat_2=-36 +x_0=0 +y_0=0 +ellps=GRS80 +units=m +no_defs"
}

# Meter-accurate buffers in the landcover tile CRS (GDA94 / Australian Albers)
points_lc <- st_transform(points_sf, lc_crs)

# Buffers at 100/200/400 m; 200 m = primary
radii_m <- c(100, 200, 400)
buffers_list <- setNames(lapply(radii_m, \(r) st_buffer(points_lc, dist = r)), as.character(radii_m))
buffers <- buffers_list[["200"]]  # primary buffer

# Extract landcover data for the primary buffer (200m)
extracted <- exactextractr::exact_extract(mosaic_vrt, buffers)
```

```{r style, message=FALSE, warning=FALSE}

summary_list <- lapply(seq_along(extracted), function(i) {
  df  <- extracted[[i]]
  agg <- aggregate(df$coverage_fraction,
                   by   = list(class = round(df$value)),
                   FUN  = sum)
  colnames(agg) <- c("land_cover_class", "fraction")
  agg$percent <- round(100 * agg$fraction / sum(agg$fraction), 2)
  agg$id      <- sample_points$id[i]
  agg
})
result_df <- bind_rows(summary_list)

# 1) Read & strip ns
qml_path <- "Landcover/landcover_class_cyear_4_style.qml"
qml  <- xml_ns_strip(read_xml(qml_path))

# 2) Grab the palette entries
entries <- xml_find_all(qml, ".//paletteEntry")
stopifnot(length(entries) > 0)  # confirm you actually found some

# 3) Build your style table from those entries
style_df <- tibble(
  land_cover_class       = as.integer(xml_attr(entries, "value")),
  land_cover_label       = xml_attr(entries, "label"),
  color                  = xml_attr(entries, "color")
) %>%
  mutate(
    form = case_when(
      str_detect(land_cover_label, "Cultivated Terrestrial Vegetated: Herbaceous")    ~ "Cultivated Herbaceous
      ",
      str_detect(land_cover_label, "Cultivated Terrestrial Vegetated: Woody")    ~ "Cultivated woody",
      str_detect(land_cover_label, "Natural Terrestrial Vegetated: Woody")     ~ "Natural - Woody",
      str_detect(land_cover_label, "Natural Terrestrial Vegetated: Herbaceous")~ "Natural - Herbaceous",
      str_detect(land_cover_label, "Aquatic Vegetated")                        ~ "Aquatic",
      str_detect(land_cover_label, "Natural Surface")                          ~ "Bare Surface",
      str_detect(land_cover_label, "Artificial Surface")                       ~ "Urban/Artificial",
      TRUE                                                                      ~ NA_character_
    ),
    density = case_when(
      str_detect(land_cover_label, "Closed \\(greater than 65") ~ 80,
      str_detect(land_cover_label, "Open \\(40 to 65")         ~ 52.5,
      str_detect(land_cover_label, "Open \\(15 to 40")         ~ 27.5,
      str_detect(land_cover_label, "Sparse \\(4 to 15")        ~ 9.5,
      TRUE                                                     ~ NA_real_
    )
  ) %>%
  select(land_cover_class, form, density, everything())

```

```{r pivot, message=FALSE, warning=FALSE}

# 1. Join in the density & labels
result_df <- result_df %>%
  left_join(
    style_df %>% select(land_cover_class, density, land_cover_label),
    by = "land_cover_class"
  )

# 2. Drop rows that can’t be assigned a class or label
result_df_clean <- result_df %>%
  filter(!is.na(land_cover_class), !is.na(land_cover_label))

# 3. Compute class‐mean densities and impute missing
class_means <- result_df_clean %>%
  group_by(land_cover_label) %>%
  summarize(mean_veg_density = mean(density, na.rm = TRUE))

result_df_imputed <- result_df_clean %>%
  left_join(class_means, by = "land_cover_label") %>%
  mutate(
    # if density is already there, keep it; otherwise use the class mean
    density = coalesce(density, mean_veg_density)
  ) %>%
  select(-mean_veg_density)  # drop the helper column

# 4. Pivot, fill, and summarize
landcover_wide <- result_df_imputed %>%
  pivot_wider(
    id_cols     = id,
    names_from  = land_cover_label,
    values_from = c(percent, density),
    values_fill = list(percent = 0, density = 0)
  ) %>%
  mutate(
    Cultivated_Total = rowSums(
      select(., starts_with("percent_Cultivated Terrestrial")),
      na.rm = TRUE
    ),
    Native_Total     = rowSums(
      select(
        .,
        starts_with("percent_Natural Terrestrial"),
        starts_with("percent_Natural Aquatic Vegetated")
      ),
      na.rm = TRUE
    ),
    Water            = rowSums(
      select(., starts_with("percent_Water:")),
      na.rm = TRUE
    ),
    # split your former Artificial_Bare into two pieces:
    Bare_Surface      = rowSums(
      select(., starts_with("percent_Natural Surface:")),
      na.rm = TRUE
    ),
    Urban_Artificial  = rowSums(
      select(., starts_with("percent_Artificial Surface:")),
      na.rm = TRUE
    )
  ) %>%
  rowwise() %>%
  mutate(
    Overall_Density = sum(
      c_across(starts_with("percent_")) / 100 *
      c_across(starts_with("density_")),
      na.rm = TRUE
    )
  ) %>%
  ungroup()

# Export for publication cache (ch07/ch15): bare land-cover class names as columns
landcover_wide %>%
  dplyr::select(id, dplyr::starts_with("percent_")) %>%
  dplyr::rename_with(~ sub("^percent_", "", .x), dplyr::starts_with("percent_")) %>%
  readr::write_csv("landcover_wide.csv")

```

```{r landcover_multiscale, message=FALSE, warning=FALSE}

extract_landcover_at_radius <- function(buffers_at_r) {
  ex <- exact_extract(mosaic_vrt, buffers_at_r)
  summary_list <- lapply(seq_along(ex), function(i) {
    df  <- ex[[i]]
    agg <- aggregate(df$coverage_fraction, by = list(class = round(df$value)), FUN = sum)
    colnames(agg) <- c("land_cover_class", "fraction")
    agg$percent <- round(100 * agg$fraction / sum(agg$fraction), 2)
    agg$id      <- sample_points$id[i]
    agg
  })
  bind_rows(summary_list)
}

lc_ms_long <- purrr::imap_dfr(buffers_list, function(buf, rlab) {
  out <- extract_landcover_at_radius(buf)
  out$radius_m <- as.integer(rlab)
  out
})

# labels/densities (reusing your style_df)
lc_ms_long <- lc_ms_long %>%
  left_join(style_df %>% select(land_cover_class, density, land_cover_label), by = "land_cover_class") %>%
  filter(!is.na(land_cover_class), !is.na(land_cover_label))

# impute density by class mean
class_means_ms <- lc_ms_long %>%
  group_by(land_cover_label) %>%
  summarize(mean_veg_density = mean(density, na.rm = TRUE), .groups = "drop")

lc_ms_imputed <- lc_ms_long %>%
  left_join(class_means_ms, by = "land_cover_label") %>%
  mutate(density = coalesce(density, mean_veg_density)) %>%
  select(-mean_veg_density)

# wide + Overall_Density per radius
landcover_wide_ms <- lc_ms_imputed %>%
  pivot_wider(
    id_cols    = c(id, radius_m),
    names_from = land_cover_label,
    values_from = c(percent, density),
    values_fill = list(percent = 0, density = 0)
  ) %>%
  rowwise() %>%
  mutate(
    Overall_Density = sum(
      c_across(starts_with("percent_")) / 100 *
      c_across(starts_with("density_")),
      na.rm = TRUE
    )
  ) %>%
  ungroup()

# grouped (broad) land-cover, like your plot_df
lc_ms_groups <- lc_ms_imputed %>%
  transmute(
    id, radius_m, land_cover_label, percent,
    group = case_when(
      str_detect(land_cover_label, "^Cultivated Terrestrial Vegetated: Woody")      ~ "Cultivated Vegetation - Woody",
      str_detect(land_cover_label, "^Cultivated Terrestrial Vegetated: Herbaceous") ~ "Cultivated Vegetation - Herbaceous",
      str_detect(land_cover_label, "^Natural Terrestrial Vegetated: Woody")         ~ "Natural Vegetation - Woody",
      str_detect(land_cover_label, "^Natural Terrestrial Vegetated: Herbaceous")    ~ "Natural Vegetation - Herbaceous",
      str_detect(land_cover_label, "^Natural Terrestrial Vegetated:")               ~ "Natural Vegetation - General",
      str_detect(land_cover_label, "Aquatic Vegetated")                             ~ "Aquatic Vegetation",
      str_detect(land_cover_label, "Scattered|Sparse|Natural Surface:")             ~ "Bare Surface",
      str_detect(land_cover_label, "^Artificial Surface:")                          ~ "Urban/Artificial",
      TRUE                                                                          ~ "Other"
    )
  ) %>%
  filter(group != "Water")

plot_df_ms <- lc_ms_groups %>%
  group_by(id, radius_m, group) %>%
  summarise(pct = sum(percent, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(pct = pct / sum(pct) * 100) %>%
  ungroup()

plot_df_wide_ms <- plot_df_ms %>%
  tidyr::pivot_wider(names_from = group, values_from = pct)
```

```{r group_colors_definition, message=FALSE, warning=FALSE}
# Define color scheme for landcover groups
group_colors <- c(
  "Cultivated Vegetation - Woody"      = "#6a51a3",
  "Cultivated Vegetation - Herbaceous" = "#9e9ac8",
  "Natural Vegetation - Woody"         = "#006400",
  "Natural Vegetation - Herbaceous"    = "#1b9e77",
  "Natural Vegetation - General"       = "#b2df8a",
  "Aquatic Vegetation"                 = "#41b6c4",
  "Bare Surface"                       = "#e6ab02",
  "Urban/Artificial"                   = "#757575",
  "Other"                              = "#cccccc"
)
```

```{r multiscale plot, message=FALSE, warning=FALSE}
ggplot(plot_df_ms, aes(x = factor(id), y = pct, fill = group)) +
  geom_col(width = 0.8) +
  facet_wrap(~ radius_m, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = group_colors) +
  labs(title = "Landcover per Sample at 100/200/400 m",
       x = "Sample ID", y = "Percent Cover", fill = "Category") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Stacked percentages of land-cover categories within a 200 m buffer around each sample (water removed, re-scaled to 100%)

```{r barplot visualisation data, message=FALSE, warning=FALSE}


# 1) Define your exact colors
group_colors <- c(
  "Cultivated Vegetation - Woody"      = "#6a51a3",
  "Cultivated Vegetation - Herbaceous" = "#9e9ac8",
  "Natural Vegetation - Woody"         = "#006400",
  "Natural Vegetation - Herbaceous"    = "#1b9e77",
  "Natural Vegetation - General"       = "#b2df8a",
  "Aquatic Vegetation"                 = "#41b6c4",
  "Bare Surface"                       = "#e6ab02",
  "Urban/Artificial"                   = "#757575",
  "Other"                              = "#cccccc"
)

# 2) Start from your long, imputed data
long_df <- result_df_imputed %>%
  select(id, land_cover_label, percent)

# 3) Bucket into exactly the groups you want
long_df <- long_df %>%
  mutate(
    group = case_when(
      str_detect(land_cover_label, "^Cultivated Terrestrial Vegetated: Woody")      ~ "Cultivated Vegetation - Woody",
      str_detect(land_cover_label, "^Cultivated Terrestrial Vegetated: Herbaceous") ~ "Cultivated Vegetation - Herbaceous",
      str_detect(land_cover_label, "^Natural Terrestrial Vegetated: Woody")         ~ "Natural Vegetation - Woody",
      str_detect(land_cover_label, "^Natural Terrestrial Vegetated: Herbaceous")    ~ "Natural Vegetation - Herbaceous",
      # anything else “Natural Terrestrial Vegetated” → general
      str_detect(land_cover_label, "^Natural Terrestrial Vegetated:")               ~ "Natural Vegetation - General",
      str_detect(land_cover_label, "Aquatic Vegetated")                             ~ "Aquatic Vegetation",
      # Sparse/Scattered or bare surface labels
      str_detect(land_cover_label, "Scattered|Sparse|Natural Surface:")             ~ "Bare Surface",
      str_detect(land_cover_label, "^Artificial Surface:")                          ~ "Urban/Artificial",
      TRUE                                                                          ~ "Other"
    ),
    group = factor(group, levels = names(group_colors))
  )

# 4) Drop NAs & Water, sum by id×group, then re-scale so each id sums to 100%
plot_df <- long_df %>%
  filter(!is.na(land_cover_label), group != "Water") %>%
  group_by(id, group) %>%
summarise(pct = sum(percent, na.rm = TRUE), .groups = "drop") %>%
group_by(id) %>%
mutate(pct = pct / sum(pct) * 100) %>%
ungroup()

# 5) Plot
ggplot(plot_df, aes(x = factor(id), y = pct, fill = group)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = group_colors) +
  labs(
    title = "Landcover Breakdown per Sample (No Water, Normalized)",
    x     = "Sample ID",
    y     = "Percent Cover",
    fill  = "Category"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x        = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.title       = element_text(face = "bold"),
    legend.text        = element_text(size = 10),
    plot.title         = element_text(face = "bold", size = 16, hjust = 0.5),
    panel.grid.major.x = element_blank()
  )

plot_df_wide <- plot_df %>%
  pivot_wider(names_from = group, values_from = pct)
```
```{r download devil tif, message=FALSE, warning=FALSE}

article_id <- 13728220
api_base   <- "https://api.figshare.com/v2"
ua         <- user_agent("R-thesis-script/1.0 (https://figshare.com)")

pred_path <- "predictionStack_devils_1985to2035.tif"
sd_path   <- "sdStack_devils_1985to2035.tif"

safe_get <- function(url, ...) {
  RETRY("GET", url, ua, timeout(120), times = 5, pause_min = 1,
        terminate_on = c(400,401,403,404), ...)
}

download_to <- function(url, dest) {
  resp <- safe_get(url, write_disk(dest, overwrite = TRUE), progress())
  if (http_error(resp) || !file.exists(dest) || is.na(file.info(dest)$size) || file.info(dest)$size == 0) {
    message("httr failed; trying curl::curl_download() …")
    curl::curl_download(url, destfile = dest, mode = "wb", quiet = FALSE)
  }
  stopifnot(file.exists(dest), file.info(dest)$size > 0)
}

# 1) List files in the article
files_url <- sprintf("%s/articles/%s/files", api_base, article_id)
res <- safe_get(files_url)
stop_for_status(res)
files <- fromJSON(content(res, as = "text", encoding = "UTF-8"))

# helpful log:
message("Figshare files:\n", paste0("  - ", files$name, collapse = "\n"))

# 2) Pick files by name (be tolerant to naming)
pick_file <- function(candidates_regex) {
  hit <- files[grep(candidates_regex, files$name, ignore.case = TRUE), ]
  if (nrow(hit) == 0) return(NULL)
  hit[1, ]
}

pred_file <- pick_file("prediction.*\\.tif$|predict.*\\.tif$")
sd_file   <- pick_file("sd.*\\.tif$|uncert.*\\.tif$|std.*\\.tif$")

if (!file.exists(pred_path)) {
  if (!is.null(pred_file$download_url)) {
    message("Downloading prediction stack …")
    download_to(pred_file$download_url, pred_path)
  } else if (!is.null(pred_file$id)) {
    download_to(sprintf("%s/file/download/%s", api_base, pred_file$id), pred_path)
  } else stop("Could not find prediction .tif in the Figshare record.")
} else message("Prediction raster already exists: ", pred_path)

if (!file.exists(sd_path)) {
  if (!is.null(sd_file$download_url)) {
    message("Downloading SD/uncertainty stack …")
    download_to(sd_file$download_url, sd_path)
  } else if (!is.null(sd_file$id)) {
    download_to(sprintf("%s/file/download/%s", api_base, sd_file$id), sd_path)
  } else stop("Could not find SD .tif in the Figshare record.")
} else message("SD raster already exists: ", sd_path)
```

```{r devils density - load & slice, message=FALSE, warning=FALSE}

pred_path <- "predictionStack_devils_1985to2035.tif"
sd_path   <- "sdStack_devils_1985to2035.tif"
stopifnot(file.exists(pred_path), file.exists(sd_path))

dev_all    <- terra::rast(pred_path)    # predictions stack
dev_sd_all <- terra::rast(sd_path)      # SD/uncertainty stack

# Slice years 2021–2023 (index = year - 1984)
all_years <- 1985:2035
wanted    <- 2021:2023
idx       <- match(wanted, all_years)

dev_2123 <- dev_all[[idx]]; names(dev_2123) <- paste0("y", wanted)
sd_2123  <- dev_sd_all[[idx]]; names(sd_2123) <- paste0("y", wanted, "_sd")
```

```{r devils density - mask & extract, message=FALSE, warning=FALSE}
# Ensure 'buffers' exists (from your earlier extract-buffer chunk)
stopifnot(exists("buffers"))

# 1) Mask (outside-domain detection)
mask_layer    <- dev_2123[[1]]
mask_bin      <- !is.na(mask_layer)
buffers_dev   <- sf::st_transform(buffers, terra::crs(mask_bin, proj = TRUE))
cov_valid     <- exactextractr::exact_extract(mask_bin, buffers_dev, fun = "mean")
outside_model <- is.na(cov_valid) | cov_valid == 0

# 2) Area-weighted means
pred_means <- exactextractr::exact_extract(dev_2123, buffers_dev, fun = "mean") |> tibble::as_tibble()
sd_means   <- exactextractr::exact_extract(sd_2123,  buffers_dev, fun = "mean") |> tibble::as_tibble()

# Fix column names if exact_extract returned V1,V2,...
if (!identical(names(pred_means), names(dev_2123))) names(pred_means) <- names(dev_2123)
if (!identical(names(sd_means),   names(sd_2123)))  names(sd_means)   <- names(sd_2123)

pred_means <- dplyr::rename_with(pred_means, ~ paste0("within_buffer_", .x))
sd_means   <- dplyr::rename_with(sd_means,   ~ paste0("within_buffer_", .x))

n_years   <- length(wanted)
pred_cols <- paste0("within_buffer_y", wanted)
sd_cols   <- paste0("within_buffer_y", wanted, "_sd")

dev_baseline <- tibble::tibble(id = sample_points$id) |>
  dplyr::bind_cols(pred_means, sd_means) |>
  dplyr::mutate(
    devil_density_baseline = rowMeans(dplyr::across(dplyr::all_of(pred_cols)), na.rm = TRUE),
    # variance propagation for the mean of n years
    devil_var_baseline     = rowSums(dplyr::across(dplyr::all_of(sd_cols), ~ .x^2), na.rm = TRUE) / (n_years^2),
    devil_sd_baseline      = sqrt(devil_var_baseline),
    # impute 0 outside model domain
    outside_model              = outside_model,
    devil_density_baseline_imp = ifelse(outside_model, 0, devil_density_baseline),
    devil_sd_baseline_imp      = ifelse(outside_model, 0, devil_sd_baseline)
  ) |>
  dplyr::select(-devil_var_baseline)


```

```{r devils density_multiscale, message=FALSE, warning=FALSE}

# Reproject each radius' buffers to the raster CRS
buffers_dev_list <- lapply(
  buffers_list,
  sf::st_transform,
  crs = terra::crs(dev_2123, proj = TRUE)
)

dev_ms <- purrr::imap_dfr(buffers_dev_list, function(buf, rlab) {
  # Extract predictions and SDs for this radius
  m_pred <- exactextractr::exact_extract(dev_2123, buf, fun = "mean") %>%
    tibble::as_tibble()                 # columns per year (may be V1..V3)
  m_sd   <- exactextractr::exact_extract(sd_2123, buf, fun = "mean") %>%
    tibble::as_tibble()

  # Ensure numeric matrix and take row means
  mean_pred <- rowMeans(as.matrix(m_pred), na.rm = TRUE)
  mean_sd   <- rowMeans(as.matrix(m_sd),   na.rm = TRUE)

  tibble::tibble(
    id        = sample_points$id,
    radius_m  = as.integer(rlab),
    devil_density_baseline = mean_pred,
    devil_sd_baseline      = mean_sd
  )
})

readr::write_csv(dev_ms, "data/devil_density_baseline_ms.csv")

```

```{r climate variable data compression, message=FALSE, warning=FALSE}
#download & load bioclim for AUS

wc_aus <- worldclim_country(
  country = "AUS",       # ISO code
  var     = "bio",       # bioclim variables
  res     = 0.5,         # 0.5° (~10 arc-minutes)
  path    = "data/aus"   # where to store the download
)

#rename & subset to bio1 & bio12
orig <- names(wc_aus)
new  <- sub(".*bio[_~\\.](\\d+)$", "bio\\1", orig)
names(wc_aus) <- new
wc_aus <- wc_aus[[c("bio1", "bio12")]]

#  write uncompressed, compress & tile via GDAL
final_dir <- "data/aus"
tmp_tif    <- file.path(final_dir, "wc_aus_tmp2bands.tif")
comp_tif   <- file.path(final_dir, "wc_aus_compressed.tif")
final_tif  <- file.path(final_dir, "wc_aus.tif")

# write 2-band GeoTIFF
terra::writeRaster(
  wc_aus,
  filename  = tmp_tif,
  filetype  = "GTiff",
  overwrite = TRUE
)

# apply DEFLATE compression + tiling
sf::gdal_utils(
  util        = "translate",
  source      = tmp_tif,
  destination = comp_tif,
  options     = c(
    "-co", "COMPRESS=DEFLATE",
    "-co", "PREDICTOR=2",
    "-co", "TILED=YES"
  )
)

# cleanup & rename
file.remove(tmp_tif)
raws <- setdiff(list.files(final_dir, "\\.tif$", full.names = TRUE), comp_tif)
if (length(raws)) file.remove(raws)
file.rename(comp_tif, final_tif)

# reload compressed
wc_aus <- rast(final_tif)

```


```{r climate variables data, message=FALSE, warning=FALSE}
# a) reproject buffers into raster CRS
buffers_ll <- st_transform(buffers, st_crs(wc_aus))

# b) codes & names
bio_codes <- c("bio1", "bio12")
bio_names <- c(
  annual_mean_temp     = "bio1",
  annual_precipitation = "bio12"
)

# c) extract weighted means
means_list <- purrr::map(bio_codes, function(code) {
  exact_extract(
    wc_aus[[code]],
    buffers_ll,
    fun = function(vals, coverage_fraction) {
      weighted.mean(vals, coverage_fraction, na.rm = TRUE)
    }
  )
})

# d) assemble tibble & rename
clim_df <- tibble(id = sample_points$id) %>%
  bind_cols(as_tibble(set_names(means_list, bio_codes))) %>%
  rename(!!!bio_names)

# e) fill NAs with sensible defaults
clim_df <- clim_df %>%
  replace_na(list(
    annual_mean_temp     = 13.7,  # aus ½(16.7 + 10.6)
    annual_precipitation = 675.3  # langjähriger Mittelwert Greens Beach
  ))

# final result
print(clim_df)

```

#mean temperature and precipation per annual are used for samples with NA climate daat 

```{r climate variable data multiscale, message=FALSE, warning=FALSE}
buffers_ll_list <- lapply(buffers_list, st_transform, crs = st_crs(wc_aus))

clim_ms <- purrr::imap_dfr(buffers_ll_list, function(buf, rlab) {
  means_list <- purrr::map(bio_codes, function(code) {
    exact_extract(
      wc_aus[[code]],
      buf,
      fun = function(vals, coverage_fraction) weighted.mean(vals, coverage_fraction, na.rm = TRUE)
    )
  })
  tibble(
    id  = sample_points$id,
    radius_m = as.integer(rlab),
    annual_mean_temp     = means_list[[which(bio_codes == "bio1")]],
    annual_precipitation = means_list[[which(bio_codes == "bio12")]]
  )
}) %>%
  tidyr::replace_na(list(
    annual_mean_temp     = 13.7,
    annual_precipitation = 675.3
  ))

readr::write_csv(clim_ms, "data/climate_ms.csv")
```

```{r population variable data, message=FALSE, warning=FALSE}
# 1. Download & load the 1 km WorldPop density for Australia -------------
out1km <- "data/pop1km/aus_pd_2020_1km.tif"
dir.create(dirname(out1km), recursive = TRUE, showWarnings = FALSE)

download.file(
  "https://data.worldpop.org/GIS/Population_Density/Global_2000_2020_1km/2020/AUS/aus_pd_2020_1km.tif",
  out1km, mode = "wb"
)

# compress the downloaded GeoTIFF losslessly and tile it 
tmp1km <- "data/pop1km/aus_pd_2020_1km_tmp.tif"
sf::gdal_utils(
  util        = "translate",
  source      = out1km,
  destination = tmp1km,
  options     = c(
    "-co", "COMPRESS=DEFLATE",
    "-co", "PREDICTOR=2",
    "-co", "TILED=YES"
  )
)
# replace the original with the compressed version
file.remove(out1km)
file.rename(tmp1km, out1km)

pop1km_rast <- rast(out1km)

# 2. Reproject your 150 m buffers into the raster’s CRS (WGS84 lon/lat)
buffers_ll <- st_transform(buffers, st_crs(pop1km_rast))

# 3. For each buffer, compute the mean density (people / km²)
pop1km_mean <- exact_extract(pop1km_rast, buffers_ll, fun = "mean")

# 4. Assemble into a data‐frame
pop1km_df <- tibble(
  id              = sample_points$id,
  pop_density_1km = pop1km_mean
)

print(pop1km_df)
```

```{r population variable data multiscale, message=FALSE, warning=FALSE}
buffers_ll_pop <- lapply(buffers_list, st_transform, crs = st_crs(pop1km_rast))

pop_ms <- purrr::imap_dfr(buffers_ll_pop, function(buf, rlab) {
  vals <- exactextractr::exact_extract(pop1km_rast, buf, fun = "mean")
  tibble(id = sample_points$id, radius_m = as.integer(rlab), pop_density_1km = vals)
})

readr::write_csv(pop_ms, "data/pop_density_ms.csv")
```

**finegrade vs broad enviroment diversity**
```{r landcover_proportion matrix, message=FALSE, warning=FALSE}
# 1) Build proportions from landcover_wide
landcover_prop <- landcover_wide %>%
  select(id, starts_with("percent_")) %>%
  mutate(across(-id, ~ . / 100))

prop_matrix <- landcover_prop %>%
  select(-id) %>%
  as.matrix()

prop_matrix[is.na(prop_matrix)] <- 0

# 2) Compute Hill numbers
h0 <- rowSums(prop_matrix > 0)
H  <- -rowSums(ifelse(prop_matrix > 0, prop_matrix * log(prop_matrix), 0))
h1 <- exp(H)
h2 <- 1 / rowSums(prop_matrix^2)

# 3) Bind back to the matching ids
hill_df_full <- data.frame(
  id = landcover_prop$id,
  h0 = h0,
  h1 = h1,
  h2 = h2 
)


hill_long <- hill_df_full %>%
  tidyr::pivot_longer(
    cols      = h0:h2,
    names_to  = "q",        # must be "q" for chapter 10
    values_to = "value"     # must be "value" for chapter 10
  ) %>%
  dplyr::arrange(id, q) %>%
  dplyr::select(id, q, value)  

dir.create("data", showWarnings = FALSE, recursive = TRUE)
save(hill_long, file = "data/hill.Rdata")


broad_prop <- plot_df_wide %>%
  mutate(across(-id, ~ . / 100))    # convert % → proportions

broad_mat <- broad_prop %>%
  select(-id) %>%
  as.matrix()
broad_mat[is.na(broad_mat)] <- 0    # just in case

# 2) Function to compute Hill numbers
compute_hills <- function(mat) {
  h0 <- rowSums(mat > 0)
  H1 <- -rowSums(ifelse(mat > 0, mat * log(mat), 0), na.rm = TRUE)
  h1 <- exp(H1)
  h2 <- 1 / rowSums(mat^2)
  tibble(h0 = h0, h1 = h1, h2 = h2)
}

# 3) Compute broad‐scale hills
hills_broad <- compute_hills(broad_mat)

# 4) Assemble into a data.frame with IDs
hill_broad_df <- tibble(
  id      = broad_prop$id,
  h0_broad = hills_broad$h0,
  h1_broad = hills_broad$h1,
  h2_broad = hills_broad$h2
)

# 5) Combine with your existing fine‐scale hills
hill_compare <- hill_df_full %>%
  left_join(hill_broad_df, by = "id")

# 6) Inspect & correlate
print(hill_compare)

# correlation between fine‐ and broad‐scale diversities
cor_mat <- hill_compare %>%
  select(h0, h0_broad, h1, h1_broad, h2, h2_broad) %>%
  cor()
print(cor_mat)

# 1) Numerical correlations with significance
tests <- list(
  h0  = cor.test(hill_compare$h0,  hill_compare$h0_broad),
  h1  = cor.test(hill_compare$h1,  hill_compare$h1_broad),
  h2  = cor.test(hill_compare$h2,  hill_compare$h2_broad)
)

# extract the r and p‐value neatly
corr_stats <- tibble(
  order    = c("h0","h1","h2"),
  r        = c(tests$h0$estimate,
               tests$h1$estimate,
               tests$h2$estimate),
  p_value  = c(tests$h0$p.value,
               tests$h1$p.value,
               tests$h2$p.value)
)
print(corr_stats)


# Inspect
print(hill_df_full)
```

```{r hill_multiscale, message=FALSE, warning=FALSE}
landcover_prop_ms <- landcover_wide_ms %>%
  select(id, radius_m, starts_with("percent_")) %>%
  mutate(across(-c(id, radius_m), ~ . / 100))

hills_ms <- landcover_prop_ms %>%
  group_by(radius_m) %>%
  group_modify(~{
    # .x has only rows for this radius; .y holds the key (radius_m)
    mat <- .x %>% select(-id) %>% as.matrix()
    mat[is.na(mat)] <- 0
    h0 <- rowSums(mat > 0)
    H  <- -rowSums(ifelse(mat > 0, mat * log(mat), 0))
    h1 <- exp(H)
    h2 <- 1 / rowSums(mat^2)
    tibble(id = .x$id, h0 = h0, h1 = h1, h2 = h2)  # <- do NOT add radius_m here
  }) %>%
  ungroup()

```

**hill visualisation**
```{r hill_visualisation, message=FALSE, warning=FALSE}

# 1) Compute a distance matrix on the 3 Hill numbers
hill_mat <- hill_df_full %>% select(h0, h1, h2) %>% as.matrix()
d       <- dist(hill_mat)          # Euclidean by default

# 2) Hierarchical clustering
hc      <- hclust(d, method = "ward.D2")

# 3) Cut the tree into k clusters (change k to taste)
k       <- 4
clusters <- cutree(hc, k = k)

# 4) Attach cluster labels
hill_df_cl <- hill_df_full %>%
  mutate(cluster = factor(clusters))

# 5) Plot colored by cluster
ggplot(hill_df_cl %>% pivot_longer(h0:h2, names_to="q", values_to="value"),
       aes(x=q, y=value, group=id, color=cluster)) +
  geom_line(alpha=0.6) +
  geom_point() +
  labs(
    title = paste("Hill Curves, clustered into", k, "groups"),
    x     = "Hill order (q)",
    y     = "Effective diversity",
    color = "Cluster"
  ) +
  theme_minimal()

plot(hc, labels = hill_df_full$id, main = "Dendrogram of Samples by Hill Numbers")
rect.hclust(hc, k = k, border = 2:5)  # draws boxes around your k clusters


```

```{r variables fusion, warning=FALSE, message=FALSE}
# add 'devil_density_baseline' to variables_df and selected_variables_df
variables_df <- hill_df_full %>%
  dplyr::left_join(pop1km_df,      by = "id") %>%
  dplyr::left_join(clim_df,        by = "id") %>%
  dplyr::left_join(landcover_wide, by = "id") %>%
  dplyr::left_join(
    dev_baseline %>% dplyr::select(id, devil_density_baseline),
    by = "id"
  )

selected_variables_df <- variables_df %>%
  dplyr::select(
    id,
    h1,
    pop_density_1km,
    annual_mean_temp,
    annual_precipitation,
    Overall_Density,
    devil_density_baseline  
  )

```

```{r variable fusion multiscale, message=FALSE, warning=FALSE}
variables_ms <- hills_ms %>%
  left_join(pop_ms,      by = c("id","radius_m")) %>%
  left_join(clim_ms,     by = c("id","radius_m")) %>%
  left_join(landcover_wide_ms %>% select(id, radius_m, Overall_Density), by = c("id","radius_m")) %>%
  left_join(dev_ms,      by = c("id","radius_m"))

selected_variables_ms <- variables_ms %>%
  select(id, radius_m, h1, pop_density_1km, annual_mean_temp, annual_precipitation,
         Overall_Density, devil_density_baseline)

dir.create("data", showWarnings = FALSE, recursive = TRUE)
save(selected_variables_ms, file = "data/enviromental_var_ms.Rdata")
write.csv(selected_variables_ms, "data/enviromental_var_ms.csv", row.names = FALSE)
```

```{r selected variables, message=FALSE, warning=FALSE}
save(selected_variables_df, 
     file = "data/enviromental_var.Rdata")

# 1) CSV (comma-separated values)
write.csv(
  selected_variables_df,
  file       = "data/enviromental_var.csv",
  row.names  = FALSE
)

# 2) TSV (tab-separated values)
write.table(
  selected_variables_df,
  file       = "data/enviromental_var.tsv",
  sep        = "\t",
  row.names  = FALSE,
  quote      = FALSE
)
```

**visualisation**

```{r visualisations1 - dominance, message=FALSE, warning=FALSE}
# 2) Sum percent cover by 'form' per sample
result_with_form <- result_df_imputed %>%
  left_join(
    style_df %>% select(land_cover_class, form),
    by = "land_cover_class"
  ) %>%
  filter(!is.na(form))   # drop anything that didn’t map

form_cover <- result_with_form %>%
  group_by(id, form) %>%
  summarize(percent_cover = sum(percent), .groups = "drop")

form_wide <- form_cover %>%
  pivot_wider(
    id_cols     = id,
    names_from  = form,
    values_from = percent_cover,
    values_fill = 0
  )

dominant_form <- form_cover %>%
  group_by(id) %>%
  slice_max(percent_cover, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(id, dominant_form = form)
```

```{r visualisations2 - hill long, message=FALSE, warning=FALSE}
# 2) Pivot Hill numbers to long form and join dominant cover (viz-only schema)
hill_long_viz <- hill_df_full %>%
  pivot_longer(
    cols      = h0:h2,
    names_to  = "hill_order",
    values_to = "diversity"
  ) %>%
  left_join(dominant_form, by = "id")
```

```{r visualisations3, message=FALSE, warning=FALSE}
# --- Visualization 1: Hill curves colored by dominant landcover ---
ggplot(hill_long_viz, aes(x = diversity, fill = dominant_form)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 25) +
  facet_wrap(~ hill_order, scales = "free_x") +
  labs(
    title = "Histograms of Hill Orders by Dominant Landcover Form",
    x     = "Effective Diversity",
    y     = "Count",
    fill  = "Dominant Form"
  ) +
  theme_minimal() +
  theme(
    strip.text       = element_text(face = "bold"),
    axis.text.x      = element_text(size = 10),
    legend.position  = "bottom"
  )

```

```{r visualisations4, message=FALSE, warning=FALSE}
# --- Visualization 2: Boxplots of each Hill metric by dominant cover ---
hill1 <- filter(hill_long_viz, hill_order == "h1")

ggplot(hill1, aes(x = diversity, color = dominant_form)) +
  geom_density(size = 1) +
  labs(
    title = "Density of Hill-1 Diversity by Dominant Landcover Form",
    x     = "Hill-1 (exp-Shannon)",
    y     = "Density",
    color = "Form"
  ) +
  theme_minimal()

```
```{r visualisations5, message=FALSE, warning=FALSE}
for (ord in c("h0","h1","h2")) {
  df_ord <- filter(hill_long_viz, hill_order == ord)
  p <- ggplot(df_ord, aes(x = dominant_form, y = diversity, fill = dominant_form)) +
    geom_boxplot(alpha = 0.8, outlier.size = 1) +
    labs(
      title = paste("Distribution of", ord, "by Landcover Form"),
      x     = "Dominant Form",
      y     = paste("Diversity (", ord, ")", sep = "")
    ) +
    theme_minimal() +
    theme(
      axis.text.x     = element_text(angle = 30, hjust = 1),
      legend.position = "none"
    )
  print(p)
}
```

```{r visualisations6, message=FALSE, warning=FALSE}
ggplot(hill_long_viz, aes(x = diversity)) +
  geom_histogram(bins = 30, fill = "grey80", color = "white") +
  facet_wrap(~ hill_order, scales = "free_x") +
  labs(
    title = "Overall Distribution of Hill Numbers",
    x     = "Effective Diversity",
    y     = "Count"
  ) +
  theme_minimal()
```
  
```{r visualisations7, message=FALSE, warning=FALSE}
ggplot(filter(hill_long_viz, hill_order == "h1"), 
       aes(x = diversity, fill = dominant_form)) +
  geom_histogram(position = "dodge", alpha = 0.7, bins = 20) +
  labs(
    title = "Histogram of Hill-1 by Dominant Landcover Form",
    x     = "Hill-1 (exp‐Shannon)",
    y     = "Count",
    fill  = "Dominant Form"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
  
```{r visualisations8, message=FALSE, warning=FALSE}
ggplot(filter(hill_long_viz, hill_order == "h1"), 
       aes(x = diversity, fill = dominant_form)) +
  geom_histogram(position = "dodge", alpha = 0.7, bins = 20) +
  labs(
    title = "Histogram of Hill-1 by Dominant Landcover Form",
    x     = "Hill-1 (exp‐Shannon)",
    y     = "Count",
    fill  = "Dominant Form"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r multiscale-compare boxplot, message=FALSE, warning=FALSE}

df_ms <- selected_variables_ms %>%
  pivot_longer(
    cols = -c(id, radius_m),
    names_to = "variable",
    values_to = "value"
  )

# Plot distributions across radii
ggplot(df_ms, aes(x = factor(radius_m), y = value, fill = factor(radius_m))) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Comparison of environmental variables across scales",
    x = "Buffer radius (m)",
    y = "Value",
    fill = "Radius"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold")
  )
```

```{r multiscale-lines, message=FALSE, warning=FALSE}
ggplot(df_ms, aes(x = radius_m, y = value, group = id, color = id)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 1.5) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Within-sample variation across scales",
    x = "Buffer radius (m)",
    y = "Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")


```

```{r multiscale-cor, message=FALSE, warning=FALSE}
scale_cor <- df_ms %>%
  pivot_wider(names_from = radius_m, values_from = value) %>%
  group_by(variable) %>%
  summarise(correlation_100_200 = cor(`100`, `200`, use="pairwise"),
            correlation_200_400 = cor(`200`, `400`, use="pairwise"),
            correlation_100_400 = cor(`100`, `400`, use="pairwise"))

print(scale_cor)
```

```{r hsmc variabels, message=FALSE, warning=FALSE}

env_sites <- readr::read_csv("data/sample_sites.csv", show_col_types = FALSE) %>%
  dplyr::rename(Site = site)

# 1) Replace NA devil densities with 0 to prevent hsmc crashing.
selected_variables_ms_fixed <- selected_variables_ms %>%
  mutate(devil_density_baseline = ifelse(is.na(devil_density_baseline), 0, devil_density_baseline))

# 2) Split into three per-radius datasets
r100 <- selected_variables_ms_fixed %>%
  filter(radius_m == 100) %>%
  select(-radius_m)

r200 <- selected_variables_ms_fixed %>%
  filter(radius_m == 200) %>%
  select(-radius_m)

r400 <- selected_variables_ms_fixed %>%
  filter(radius_m == 400) %>%
  select(-radius_m)


# 3) (optional) Save for HSMC runs
dir.create("data/hsmc_scales", showWarnings = FALSE, recursive = TRUE)
write.csv(r100, "data/hsmc_scales/env_vars_r100.csv", row.names = FALSE)
write.csv(r200, "data/hsmc_scales/env_vars_r200.csv", row.names = FALSE)
write.csv(r400, "data/hsmc_scales/env_vars_r400.csv", row.names = FALSE)

# Check the dimensions quickly
message("Rows at 100m: ", nrow(r100))
message("Rows at 200m: ", nrow(r200))
message("Rows at 400m: ", nrow(r400))
```

```{r make-hsmc-ready-h1, message=FALSE, warning=FALSE}

stopifnot(exists("selected_variables_ms"))

# Helper to standardize to the target schema
to_hsmc_schema <- function(df_ms, hill_metric = "h1") {
  stopifnot(hill_metric %in% c("h0","h1","h2"))
  
  # 1) Ensure a site column exists
  if (!"site" %in% names(df_ms)) {
    df_ms <- df_ms %>% mutate(site = id)
  }
  
  # 2) Choose which devil column to use
  devil_candidates <- c("devil_density_baseline_imp",
                        "devil_density_baseline",
                        "devil_density")
  devil_col <- intersect(devil_candidates, names(df_ms))
  devil_col <- if (length(devil_col)) devil_col[1] else NA_character_
  
  # 3) Build output with required names + order
  out <- df_ms %>%
    mutate(
      devil = if (!is.na(devil_col)) .data[[devil_col]] else NA_real_,
      devil = ifelse(is.na(devil), 0, devil)   # NA → 0 for HSMC robustness
    ) %>%
    transmute(
      id,
      diversity     = .data[[hill_metric]],      
      pop_density   = pop_density_1km,
      temperature   = annual_mean_temp,
      precipitation = annual_precipitation,
      density       = Overall_Density,
      devil,
      radius_m
    )
  
  out
}

# Build standardized multiscale table (still includes radius_m so we can split)
hsmc_ms <- to_hsmc_schema(selected_variables_ms, hill_metric = "h1")

# Split into three per-radius data frames for 3 HSMC runs
r100_h1 <- hsmc_ms %>% filter(radius_m == 100) %>% select(-radius_m)
r200_h1 <- hsmc_ms %>% filter(radius_m == 200) %>% select(-radius_m)
r400_h1 <- hsmc_ms %>% filter(radius_m == 400) %>% select(-radius_m)

r200_h1 <- r200_h1 |>
  left_join(env_sites, by = c("id" = "id"))


```

```{r load_sample_metadata_hsmc, warning=FALSE, comments="", message=FALSE}
load("data/data.Rdata")

sample_metadata <- sample_metadata %>%
  dplyr::filter(sample != "EHI01340")

sample_metadata_tasmania <- sample_metadata %>%
  dplyr::filter(island == "Tasmania")
```

```{r tasmania_australia_split_r200, warning=FALSE, comments="", message=FALSE}

r200_h1 <- r200_h1 %>%
  left_join(
    sample_metadata %>% select(sample, island),
    by = c("id" = "sample")
  )

r200_h1_tas <- r200_h1 %>% filter(island == "Tasmania")
r200_h1_aus <- r200_h1 %>% filter(island == "Australia")


```

```{r tasmania_australia_split_save, warning=FALSE, comments="", message=FALSE}
# (Optional) save
dir.create("data/hsmc_scales", showWarnings = FALSE, recursive = TRUE)
write.csv(r100_h1, "data/hsmc_scales/env_vars_r100_h1.csv", row.names = FALSE)
write.csv(r200_h1_tas, "data/hsmc_scales/env_vars_r200_h1_tas.csv", row.names = FALSE)
write.csv(r400_h1, "data/hsmc_scales/env_vars_r400_h1.csv", row.names = FALSE)

# Sanity check
message("Columns: ", paste(names(r200_h1), collapse = ", "))
message("Rows (100/200/400): ", nrow(r100_h1), "/", nrow(r200_h1), "/", nrow(r400_h1))

```

```{r tasmania_animal_var, warning=FALSE, comments="", message=FALSE}
animal_vars_tas <- sample_metadata_tasmania |>
  select(sample, sex, Development, `Weight (g)`) |>
  filter(
    !is.na(sex),
    sex != "Unknown",
    !is.na(`Weight (g)`)
  )

write.csv(animal_vars_tas, "data/hsmc_scales/animal_vars_tas.csv", row.names = FALSE)
```

```{r tasmania_env_short, warning=FALSE, comments="", message=FALSE}

tasmania_env_short <- r200_h1_tas |>
  dplyr::filter(id %in% animal_vars_tas$sample)

write.csv(tasmania_env_short, "data/hsmc_scales/tasmania_env_short.csv", row.names = FALSE)
```

```{r export_hmsc_input_csvs, warning=FALSE, comments="", message=FALSE}
# Canonical HMSC inputs for chapter 13 (regenerated on each knit of this chapter).
readr::write_csv(
  sample_metadata %>%
    dplyr::select(id = sample, sex, development = Development, weight = `Weight (g)`),
  "data/animal_var.csv"
)

readr::write_csv(
  r200_h1 %>%
    dplyr::select(id, site = Site, diversity, pop_density, temperature, precipitation, density, devil),
  "data/environment_var_r200.csv"
)
```