
# Beta diversity

Between-sample dissimilarity using Hill-based distances (richness, neutral, phylogenetic, functional), PERMANOVA tests for broad environment/island effects, and NMDS ordinations to visualise separation in community structure.

**Note:** Load all required data objects including genome counts, metadata, and color schemes.

```{r load_data_beta, message=FALSE, warning=FALSE}
load("data/data.Rdata")

treatment_colors <- c("#f56042","#429ef5", "#42f58d", "#b142f5", "#f5e642")
```

**Note:** Filter out sample EHI01340, identify valid samples, remove samples with missing broad_environment values, and filter genome counts to only include valid samples.

```{r filter_NAs, message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
valid_samples <- sample_metadata$sample

sample_metadata_without_NA <- sample_metadata %>%
  dplyr::filter(!is.na(broad_environment))

genome_counts_filt <- genome_counts_filt %>%
  dplyr::select(genome, tidyselect::any_of(valid_samples))
```

***beta_div***

**Note:** Calculate four types of beta diversity distance matrices using Hill-based pairwise distances: richness-based (q=0, presence/absence), neutral/abundance-weighted (q=1), phylogenetic (q=1 with phylogenetic tree), and functional (q=1 with functional trait distances). Filters out genomes and samples with zero counts before calculation.

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval = FALSE}
#Presence/Absence
beta_q0n <- genome_counts_filt %>%
  tibble::column_to_rownames(., "genome") %>%
  dplyr::filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  dplyr::select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)
#Abundance-sensitive
beta_q1n <- genome_counts_filt %>%
  tibble::column_to_rownames(., "genome") %>%
  dplyr::filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  dplyr::select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)
#Abundance + Phylogeny
genome_counts_filt_beta <- genome_counts_filt %>%
  tibble::column_to_rownames(., "genome") %>%
  dplyr::filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  dplyr::select_if(~!all(. == 0)) %>%
  tibble::rownames_to_column(., "genome")
#Abundance + Trait Distance
genome_tree <- keep.tip(genome_tree, tip = genome_counts_filt_beta$genome)
beta_q1p <- genome_counts_filt %>%
  tibble::column_to_rownames(., "genome") %>%
  dplyr::filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  dplyr::select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)
```

**Note:** Calculate functional beta diversity (beta_q1f) by matching genomes between count data and functional trait data (genome_gifts). Converts GIFT traits to elements, calculates Gower distance matrix, filters and aligns both datasets to common genomes, then computes Hill-based pairwise distances with functional trait distances.

```{r beta_div2, comment="", message=FALSE, warning=FALSE, eval=TRUE}
# Create genome_counts_filt_beta if it doesn't exist (from beta_div chunk)
if (!exists("genome_counts_filt_beta")) {
  genome_counts_filt_beta <- genome_counts_filt %>%
    tibble::column_to_rownames(., "genome") %>%
    dplyr::filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
    dplyr::select_if(~!all(. == 0)) %>%
    tibble::rownames_to_column(., "genome")
}

# 1. Only use genomes that also occur in genome_gifts
valid_genomes <- genome_counts_filt_beta$genome[genome_counts_filt_beta$genome %in% rownames(genome_gifts)]

# 2. filter genome_gifts on these genomes
genome_gifts1 <- genome_gifts[valid_genomes, ]
genome_gifts1 <- genome_gifts1[, colSums(genome_gifts1 != 0) > 0]

# 3. Calculate functional distance matrix
dist <- genome_gifts1 %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

# 4. Determine genomes that exist in both genome_counts_filt_beta and dist
common_genomes <- intersect(
  genome_counts_filt_beta$genome,
  rownames(dist)
)

# 5. genome_counts_filt to filter and sort the shared genomes
genome_counts_filt_matched <- genome_counts_filt %>%
  tibble::column_to_rownames("genome") %>%
  dplyr::filter(rownames(.) %in% common_genomes) %>%
  dplyr::select_if(~!all(. == 0))  # optional: remove empty columns

# 6. Limit and sort dist matrix to common genomes
dist_matched <- dist[common_genomes, common_genomes]

beta_q1f <- genome_counts_filt_matched %>%
  hillpair(q = 1, dist = dist_matched)
```

**Note:** Save all four beta diversity distance matrices (richness, neutral, phylogenetic, functional) to an R data file for later use.

```{r save_beta, comment="", message=FALSE,echo=FALSE,warning=FALSE, eval=FALSE}
save(beta_q0n, 
     beta_q1n, 
     beta_q1p, 
     beta_q1f, 
     file = "data/beta.Rdata")
```

**Note:** Set random seed for reproducibility of permutation-based statistical tests.

```{r seed, comment="", message=FALSE,echo=FALSE,warning=FALSE, eval=FALSE}

set.seed(2025)
```

## Permanova

**Note:** Define a reusable function `run_permanova` that performs PERMANOVA analysis on beta diversity matrices. The function matches samples between distance matrix and metadata, filters out NAs, tests beta dispersion (homogeneity of variances), runs PERMANOVA with adonis2, optionally performs pairwise comparisons, and returns both dispersion and PERMANOVA results.

```{r Funktion für PERMANOVA Analysen, comment="", message=FALSE, warning=FALSE}
run_permanova <- function(beta_matrix,
                          metadata,
                          sample_col = "sample",
                          group_col,
                          group_vars,
                          pairwise = FALSE) {
  
  # 1. Sample labels from Beta-Matrix (dist object)
  sample_labels <- labels(beta_matrix$S)
  
  # 2. Metadata filtered and sorted
  metadata_matched <- metadata %>%
    dplyr::filter(!!rlang::sym(sample_col) %in% sample_labels) %>%
    dplyr::arrange(match(!!rlang::sym(sample_col), sample_labels))
  
  # 3. Synchronize beta_matrix$S to metadata (subset via matrix, then back to dist)
  common_samples <- intersect(sample_labels, metadata_matched[[sample_col]])
  
  S_mat <- as.matrix(beta_matrix$S)
  S_mat <- S_mat[common_samples, common_samples, drop = FALSE]
  beta_matrix$S <- as.dist(S_mat)
  
  metadata_matched <- metadata_matched %>%
    dplyr::filter(!!rlang::sym(sample_col) %in% common_samples) %>%
    dplyr::arrange(match(!!rlang::sym(sample_col), common_samples))
  
  # 4. Remove samples with NA in group_col (again via matrix → dist)
  valid_idx <- !is.na(metadata_matched[[group_col]])
  
  S_mat <- as.matrix(beta_matrix$S)
  S_mat <- S_mat[valid_idx, valid_idx, drop = FALSE]
  beta_matrix$S <- as.dist(S_mat)
  
  metadata_matched <- metadata_matched[valid_idx, ]
  
  # 5. Betadisper and Permutest
  cat("\n===== Beta Dispersion Test =====\n")
  dispersion_result <- betadisper(beta_matrix$S, metadata_matched[[group_col]])
  print(permutest(dispersion_result))
  
  # 6. Dynamic formula building
  formula_str <- paste0("beta_matrix$S ~ ", paste(group_vars, collapse = " + "))
  cat("\n===== PERMANOVA Test (adonis2) =====\n")
  permanova_result <- adonis2(
    as.formula(formula_str),
    data = metadata_matched,
    permutations = 999,
    by = "terms"
  )
  print(broom::tidy(permanova_result))
  
  # 7. (Optional) Pairwise Adonis
  if (pairwise) {
    cat("\n===== Pairwise PERMANOVA =====\n")
    print(pairwise.adonis(beta_matrix$S, metadata_matched[[group_col]], perm = 999))
  }
  
  # 8. Return outputs
  return(list(
    dispersion = dispersion_result,
    permanova  = permanova_result
  ))
}

```

### Richness

**Note:** Run PERMANOVA analysis on richness-based beta diversity (beta_q0n) to test for differences in community composition across broad environment types and islands, including interaction effects. Tests whether presence/absence community structure differs between environments.

```{r permanova_richness, comment="", message=FALSE, warning=FALSE}
run_permanova(
  beta_matrix = beta_q0n,
  metadata = sample_metadata_without_NA,
  sample_col = "sample",
  group_col = "broad_environment",
  group_vars = c("broad_environment", "island", "broad_environment:island"),
  pairwise = FALSE
)

```

### Neutral diversity

**Note:** Run PERMANOVA analysis on abundance-weighted beta diversity (beta_q1n) to test for differences in community composition across broad environment types and islands, including interaction effects and pairwise comparisons. Tests whether abundance-weighted community structure differs between environments.

```{r permanova_neutral, comment="", message=FALSE, warning=FALSE}
run_permanova(
  beta_matrix = beta_q1n,
  metadata = sample_metadata_without_NA,
  sample_col = "sample",
  group_col = "broad_environment",
  group_vars = c("broad_environment", "island", "broad_environment:island"),
  pairwise = TRUE
)
```

### Phylogenetic diversity (beta_q1p)

**Note:** Run PERMANOVA analysis on phylogenetic beta diversity (beta_q1p) to test for differences in community composition across broad environment types and islands, including interaction effects. Tests whether phylogenetically-weighted community structure differs between environments.

```{r permanova_phylo, comment="", message=FALSE, warning=FALSE}
run_permanova(
  beta_matrix = beta_q1p,
  metadata = sample_metadata_without_NA,
  sample_col = "sample",
  group_col = "broad_environment",
  group_vars = c("broad_environment", "island", "broad_environment:island"),
  pairwise = FALSE  # no pairwise comparisons here
)
```

### Functional diversity (beta_q1f)

**Note:** Run PERMANOVA analysis on functional beta diversity (beta_q1f) to test for differences in community composition across broad environment types and islands, including interaction effects and pairwise comparisons. Tests whether functionally-weighted community structure differs between environments.

```{r permanova_func, comment="", message=FALSE, warning=FALSE}
run_permanova(
  beta_matrix = beta_q1f,
  metadata = sample_metadata_without_NA,
  sample_col = "sample",
  group_col = "broad_environment",
  group_vars = c("broad_environment", "island", "broad_environment:island"),
  pairwise = TRUE  
)
```

## NMDS 

NMDS of β-diversity (richness/neutral/phylogenetic/functional). Points are samples, coloured by broad environment and shaped by island; segments draw each point to its group centroid.

**Note:** Define a reusable function to create NMDS ordination plots. The function performs metaMDS on the distance matrix, extracts NMDS scores, joins with sample metadata, calculates group centroids, and visualizes samples with optional labels.

```{r nmds_plot_function, comment="", message=FALSE, warning=FALSE}
# Function to calculate overall axis limits from all beta matrices
calculate_nmds_limits <- function(beta_matrices) {
  all_scores <- purrr::map_dfr(beta_matrices, function(beta_mat) {
    beta_mat$S %>%
      vegan::metaMDS(trymax = 500, k = 2, trace = 0) %>%
      vegan::scores() %>%
      tibble::as_tibble(rownames = "sample")
  })
  
  x_range <- range(all_scores$NMDS1, na.rm = TRUE)
  y_range <- range(all_scores$NMDS2, na.rm = TRUE)
  
  # Add some padding (5% on each side)
  x_padding <- diff(x_range) * 0.05
  y_padding <- diff(y_range) * 0.05
  
  list(
    xlim = c(x_range[1] - x_padding, x_range[2] + x_padding),
    ylim = c(y_range[1] - y_padding, y_range[2] + y_padding)
  )
}

create_nmds_plot <- function(beta_matrix, 
                             show_labels = FALSE,
                             point_size = 3,
                             segment_alpha = 0.3,
                             show_ellipses = FALSE,
                             show_centroids = TRUE,
                             xlim = NULL,
                             ylim = NULL) {
  # Access global variables explicitly with fallbacks
  if (exists("sample_metadata", envir = .GlobalEnv)) {
    sample_metadata <- get("sample_metadata", envir = .GlobalEnv)
  } else {
    stop("sample_metadata not found in global environment")
  }
  
  if (exists("treatment_colors", envir = .GlobalEnv)) {
    treatment_colors <- get("treatment_colors", envir = .GlobalEnv)
  } else {
    # Fallback: use default colors
    treatment_colors <- c("#f56042", "#429ef5", "#42f58d", "#b142f5", "#f5e642", 
                          "#FFA500", "#FF69B4", "#00CED1", "#9370DB", "#FFD700")
  }
  
  # Calculate NMDS and get stress value
  nmds_obj <- beta_matrix$S %>%
    vegan::metaMDS(trymax = 500, k = 2, trace = 0)
  
  stress_value <- nmds_obj$stress
  
  # Extract scores and join with metadata
  nmds_scores <- nmds_obj %>%
    vegan::scores() %>%
    tibble::as_tibble(rownames = "sample") %>%
    dplyr::inner_join(sample_metadata, by = join_by(sample == sample)) %>%
    group_by(broad_environment) %>%
    mutate(
      x_cen = mean(NMDS1, na.rm = TRUE),
      y_cen = mean(NMDS2, na.rm = TRUE)
    ) %>%
    ungroup()
  
  # Determine all factor levels - check both the joined data and original metadata
  # First, check if broad_environment is a factor in the joined data
  if (is.factor(nmds_scores$broad_environment)) {
    # Get levels from the factor (this includes all levels, even if not in data)
    all_env_levels <- levels(nmds_scores$broad_environment)
  } else if (is.factor(sample_metadata$broad_environment)) {
    # If not a factor in joined data, check original metadata
    all_env_levels <- levels(sample_metadata$broad_environment)
    # Convert to factor with these levels
    nmds_scores$broad_environment <- factor(nmds_scores$broad_environment, levels = all_env_levels)
  } else {
    # If not a factor anywhere, get unique values
    all_env_levels <- unique(nmds_scores$broad_environment)
    all_env_levels <- all_env_levels[!is.na(all_env_levels)]
    nmds_scores$broad_environment <- factor(nmds_scores$broad_environment, levels = all_env_levels)
  }
  
  # Remove NA from levels if present
  all_env_levels <- all_env_levels[!is.na(all_env_levels)]
  n_levels <- length(all_env_levels)
  
  # Ensure we have enough colors - ALWAYS recycle to match number of levels
  # This is critical: we need colors for ALL factor levels
  # Check if treatment_colors has values
  if (length(treatment_colors) == 0) {
    # Fallback: create default colors if treatment_colors is empty
    treatment_colors <- grDevices::rainbow(n_levels)
  }
  
  # Always create color vector with exactly n_levels colors (recycling if needed)
  # This ensures we have exactly the right number of colors for all factor levels
  color_values <- rep(treatment_colors, length.out = n_levels)
  
  # Debug: print if we're recycling (optional, can remove later)
  if (n_levels > length(treatment_colors)) {
    message(paste0("Recycling ", length(treatment_colors), " colors to create ", 
                   n_levels, " colors for ", n_levels, " factor levels"))
  }
  
  # Create named vector - names must match ALL factor level values exactly
  # This ensures ggplot2 can match each level to its color
  names(color_values) <- as.character(all_env_levels)
  
  # Ensure broad_environment is a factor with all levels (even if not all appear in data)
  # This is critical - ggplot2 needs colors for ALL factor levels, not just those in the data
  nmds_scores$broad_environment <- factor(nmds_scores$broad_environment, levels = all_env_levels)
  
  # Final verification: ensure color vector has the right length
  stopifnot(length(color_values) == n_levels)
  stopifnot(all(names(color_values) %in% all_env_levels))
  
  # Create base plot
  p <- nmds_scores %>%
    ggplot(aes(x = NMDS1, y = NMDS2, 
               color = broad_environment, 
               shape = as.factor(island))) +
    # Add segments from centroids to points (lighter and thinner)
    geom_segment(
      aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2),
      alpha = segment_alpha,
      linewidth = 0.3,
      color = "gray60",
      show.legend = FALSE
    ) +
    # Add sample points
    geom_point(size = point_size, alpha = 0.8) +
    # Color scale with proper mapping - use named vector for automatic matching
    scale_color_manual(
      values = color_values
    ) +
    # Shape scale
    scale_shape_discrete(name = "Island")
  
  # Add confidence ellipses if requested
  if (show_ellipses) {
    p <- p + stat_ellipse(
      aes(color = broad_environment),
      type = "norm",
      level = 0.95,
      linetype = 2,
      alpha = 0.5,
      linewidth = 0.5,
      show.legend = FALSE
    )
  }
  
  # Add centroid markers if requested
  if (show_centroids) {
    p <- p + geom_point(
      data = nmds_scores %>% 
        dplyr::distinct(broad_environment, x_cen, y_cen),
      aes(x = x_cen, y = y_cen, color = broad_environment),
      shape = 21,
      size = 6,
      stroke = 2,
      fill = "white",
      show.legend = FALSE
    )
  }
  
  # Add theme and labels
  p <- p +
    # Theme
    theme(
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(size = 10),
      panel.background = element_blank(),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
      panel.grid.minor = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 10),
      legend.title = element_text(size = 11, face = "bold"),
      legend.position = "right", 
      legend.box = "vertical",
      legend.spacing = unit(0.5, "cm"),
      legend.key.size = unit(1, "lines")
    ) +
    # Labels with stress value in title
    labs(
      title = paste0("NMDS (Stress = ", round(stress_value, 3), ")"),
      x = "NMDS1",
      y = "NMDS2",
      color = "Environmental context",
      shape = "Island"
    ) +
    # Legend guides
    guides(
      color = guide_legend(order = 1, override.aes = list(size = 3)),
      shape = guide_legend(order = 2, override.aes = list(size = 3))
    )
  
  # Add coordinate system with fixed limits if provided
  if (!is.null(xlim) && !is.null(ylim)) {
    p <- p + coord_fixed(ratio = 1, xlim = xlim, ylim = ylim, expand = TRUE)
  } else {
    p <- p + coord_fixed(ratio = 1, expand = TRUE)
  }
  
  # Text labels removed - no longer adding sample IDs
  
  return(p)
}
```

**Note:** Calculate overall axis limits from all beta diversity matrices to ensure consistent scaling across all NMDS plots.

```{r calculate_nmds_limits, comment="", message=FALSE, warning=FALSE}
# Calculate limits once from all beta matrices
nmds_limits <- calculate_nmds_limits(list(beta_q0n, beta_q1n, beta_q1p, beta_q1f))
```

###Richness

**Note:** Create NMDS ordination plot for richness-based beta diversity. Performs metaMDS on the distance matrix, extracts NMDS scores, joins with sample metadata, calculates group centroids for each broad environment, and visualizes samples colored by environment and shaped by island, with segments connecting samples to their group centroids.

```{r beta_div_nmds_richness_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
create_nmds_plot(beta_q0n, 
                 show_labels = FALSE,
                 xlim = nmds_limits$xlim,
                 ylim = nmds_limits$ylim)
```

### Neutral diversity

**Note:** Create NMDS ordination plot for abundance-weighted beta diversity. Performs metaMDS on the distance matrix, extracts NMDS scores, joins with sample metadata, calculates group centroids for each broad environment, and visualizes samples colored by environment and shaped by island, with segments connecting samples to their group centroids.

```{r beta_div_nmds_neutral_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
create_nmds_plot(beta_q1n, 
                 show_labels = FALSE,
                 xlim = nmds_limits$xlim,
                 ylim = nmds_limits$ylim)
```

### Phylogenetic diversity

**Note:** Create NMDS ordination plot for phylogenetic beta diversity. Performs metaMDS on the distance matrix, extracts NMDS scores, joins with sample metadata, calculates group centroids for each broad environment, and visualizes samples colored by environment and shaped by island, with segments connecting samples to their group centroids.

```{r beta_div_nmds_phylo_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
create_nmds_plot(beta_q1p, 
                 show_labels = FALSE,
                 xlim = nmds_limits$xlim,
                 ylim = nmds_limits$ylim)
```


[Simply the graph to make it easier explained - community gifts-nmds sample vs gifts]
### Functional diversity

**Note:** Create NMDS ordination plot for functional beta diversity. Performs metaMDS on the distance matrix, extracts NMDS scores, joins with sample metadata, calculates group centroids for each broad environment, and visualizes samples colored by environment and shaped by island, with segments connecting samples to their group centroids.

```{r beta_div_nmds_funct_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
create_nmds_plot(beta_q1f, 
                 show_labels = FALSE,
                 xlim = nmds_limits$xlim,
                 ylim = nmds_limits$ylim)
```

**Note:** Create a zoomed-in version of the functional diversity NMDS plot with axes fitted to the data range for better visualization of the functional diversity patterns. The y-axis is stretched with additional padding for improved visibility.

```{r beta_div_nmds_funct_plot_zoomed, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
# Calculate functional diversity specific limits with more y-axis padding
func_nmds_obj <- beta_q1f$S %>%
  vegan::metaMDS(trymax = 500, k = 2, trace = 0)

func_scores <- func_nmds_obj %>%
  vegan::scores() %>%
  tibble::as_tibble(rownames = "sample")

x_range_func <- range(func_scores$NMDS1, na.rm = TRUE)
y_range_func <- range(func_scores$NMDS2, na.rm = TRUE)

# Add padding: 5% for x-axis, 15% for y-axis (more stretch on y)
x_padding_func <- diff(x_range_func) * 0.05
y_padding_func <- diff(y_range_func) * 0.50

func_limits <- list(
  xlim = c(x_range_func[1] - x_padding_func, x_range_func[2] + x_padding_func),
  ylim = c(y_range_func[1] - y_padding_func, y_range_func[2] + y_padding_func)
)

create_nmds_plot(beta_q1f, 
                 show_labels = FALSE,
                 xlim = func_limits$xlim,
                 ylim = func_limits$ylim)
```

**Note:** Create a zoomed-in version of the NMDS plot using community-weighted function-level GIFTs (GIFTs_functions_community). Calculates Euclidean distance matrix from the function-level community data and visualizes with axes fitted to the data range for better visualization.

```{r beta_div_nmds_gift_functions_community_plot_zoomed, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
# Calculate distance matrix from GIFTs_functions_community
# Ensure samples are in rows and functions in columns
gift_functions_dist <- GIFTs_functions_community %>%
  as.data.frame() %>%
  stats::dist(method = "euclidean")

# Create a structure similar to beta_q1f (with $S component) for compatibility with create_nmds_plot
gift_functions_beta <- list(S = gift_functions_dist)

# Calculate function-level community specific limits with more y-axis padding
gift_func_nmds_obj <- gift_functions_beta$S %>%
  vegan::metaMDS(trymax = 500, k = 2, trace = 0)

gift_func_scores <- gift_func_nmds_obj %>%
  vegan::scores() %>%
  tibble::as_tibble(rownames = "sample")

x_range_gift_func <- range(gift_func_scores$NMDS1, na.rm = TRUE)
y_range_gift_func <- range(gift_func_scores$NMDS2, na.rm = TRUE)

# Add padding: 5% for x-axis, 50% for y-axis (more stretch on y)
x_padding_gift_func <- diff(x_range_gift_func) * 0.05
y_padding_gift_func <- diff(y_range_gift_func) * 0.50

gift_func_limits <- list(
  xlim = c(x_range_gift_func[1] - x_padding_gift_func, x_range_gift_func[2] + x_padding_gift_func),
  ylim = c(y_range_gift_func[1] - y_padding_gift_func, y_range_gift_func[2] + y_padding_gift_func)
)

create_nmds_plot(gift_functions_beta, 
                 show_labels = FALSE,
                 xlim = gift_func_limits$xlim,
                 ylim = gift_func_limits$ylim)
```

