
# Differential abundance analysis

ANCOM-BC2-style models test which taxa change with sex, with and without random effects and while accounting for structural zeros. Results are summarised at MAG, phylum, family, and genus levels.

**Note:** Load all required data objects including genome counts, metadata, and color schemes.

```{r load_data_mag_filtdamr_diffe_sex, comment="", echo=FALSE, message=FALSE, warning=FALSE}

load("data/data.Rdata")
```

**Note:** Filter out sample EHI01340 and filter genome counts to only include valid samples.

```{r exclude_Sample7_sex, message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
valid_samples <- sample_metadata$sample

genome_counts_filt <- genome_counts_filt %>%
  dplyr::select(genome, tidyselect::any_of(valid_samples))
```

**Note:** Remove genomes and samples with zero counts across all samples to clean the data.

```{r genom_count_filt_sex, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))
```

**Note:** Create a phyloseq object combining count data, sample metadata, taxonomy table, and phylogenetic tree for downstream analyses.

```{r phyloseq, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
#Phyloseq object
count_phy <- genome_counts_filt %>%
  column_to_rownames(var="genome") %>%
  otu_table(., taxa_are_rows=T)

sample_info_tab_phy <- sample_metadata %>%
  column_to_rownames(var="sample") %>%
  sample_data()

TAX <- genome_metadata %>%
  column_to_rownames(var="genome") %>%
  select(1:7) %>%
  as.matrix() %>%
  tax_table()
tree <- phy_tree(genome_tree)

physeq_all = phyloseq(count_phy, TAX, sample_info_tab_phy, tree)
```

## Structural zeros

**Note:** Identify structural zeros - genomes that are completely absent in one sex group but present in the other. Calculate average abundances for genomes present only in one group and attach phylum-level taxonomy.

```{r struct_zero, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# groups
Female_samples <- sample_metadata %>% filter(sex == "Female") %>% pull(sample)
Male_samples   <- sample_metadata %>% filter(sex == "Male")   %>% pull(sample)

# keep only samples that exist in the counts
existing_samples <- colnames(genome_counts_filt)
Female_samples <- intersect(Female_samples, existing_samples)
Male_samples   <- intersect(Male_samples,   existing_samples)

# compute structural zeros row-wise
structural_zeros <- genome_counts_filt %>% 
  rowwise() %>%
  mutate(
    all_zeros_Female = all(c_across(all_of(Female_samples)) == 0),
    all_zeros_Male   = all(c_across(all_of(Male_samples))   == 0),
    average_Female   = mean(c_across(all_of(Female_samples)), na.rm = TRUE),
    average_Male     = mean(c_across(all_of(Male_samples)),   na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(all_zeros_Female | all_zeros_Male) %>%
  mutate(
    # present only in the *other* group (the one that is not all zeros)
    present = dplyr::case_when(
      all_zeros_Female & !all_zeros_Male ~ "Male",
      all_zeros_Male   & !all_zeros_Female ~ "Female",
      TRUE ~ NA_character_
    ),
    average = dplyr::case_when(
      present == "Female" ~ average_Female,
      present == "Male"   ~ average_Male,
      TRUE ~ NA_real_
    )
  )

# attach taxonomy
structural_zeros <- structural_zeros %>%
  dplyr::left_join(genome_metadata %>% dplyr::select(genome, phylum), by = "genome")

# normalize possible taxonomic prefixes on both sides
structural_zeros <- structural_zeros %>%
  dplyr::mutate(phylum = stringr::str_replace(phylum, "^[dpcofgs]__", ""))

```

**Note:** Create a bar plot showing the top 30 genomes with the strongest "present-only-in" patterns, displaying average counts for genomes found exclusively in either Female or Male samples.

```{r structu_Zero_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#top 30 strongest "present-only-in" genomes
structural_zeros_top <- structural_zeros %>%
  dplyr::mutate(abs_avg = abs(average)) %>%
  dplyr::arrange(dplyr::desc(abs_avg)) %>%
  dplyr::slice_head(n = 30)

xmax <- max(structural_zeros_top$abs_avg, na.rm = TRUE) * 1.1

ggplot(structural_zeros_top,
       aes(x = average,
           y = forcats::fct_reorder(genome, average),
           fill = present)) +
  geom_col(width = 0.8) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c(Female = "#C33C54", Male = "#2A9D8F"),
                    name = "Present only in") +
  coord_cartesian(xlim = c(-xmax, xmax), expand = FALSE) +
  labs(x = "Genome counts (negative = Male only, positive = Female only)",
       y = "MAG (top 30 by |difference|)") +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 6),
        plot.margin = margin(6, 6, 6, 6))

```

**Note:** Create a phylum-level summary bar chart showing the number of MAGs present only in each sex group, organized by phylum.

```{r chart2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#phylum overview of "present only in"
phylum_summary_sz <- structural_zeros %>%
  dplyr::filter(!is.na(phylum), !is.na(present)) %>%
  dplyr::count(phylum, present, name = "n") %>%
  tidyr::complete(phylum, present = c("Female","Male"), fill = list(n = 0)) %>%
  dplyr::group_by(phylum) %>%
  dplyr::mutate(total = sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(dplyr::desc(total))

ggplot(phylum_summary_sz,
       aes(x = reorder(phylum, total), y = n, fill = present)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c(Female = "#C33C54", Male = "#2A9D8F"),
                    name = "Present only in") +
  labs(x = "Phylum", y = "Number of MAGs present only in group") +
  theme_classic() +
  theme(legend.position = "top")
```

## MAGs in different locations and shared among locations

**Note:** Create an UpSet plot showing the distribution of MAGs across different broad environments (Mixed forest, Temperate woodland, Xeric shrubland, Cropland), highlighting unique and shared MAGs.

```{r chart1, comment="", echo=FALSE, message=FALSE, warning=FALSE}
locationcolors=c('#408892','#c04062')

genome_counts_rel <- genome_counts %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  column_to_rownames(., "genome")

genome_counts_rel_fil <- genome_counts_filt %>%
    select(one_of(c("genome",sample_metadata$sample))) %>%
  column_to_rownames(., "genome")
  
genome_counts_rel_pa = 1*(genome_counts_rel_fil>0)
#MAGrel_pa[1:6,1:6]
table_upset_analysis_cont = t(aggregate(t(genome_counts_rel_pa),by=list(sample_metadata$sex),FUN=sum)[,-1])
colnames(table_upset_analysis_cont) = levels(as.factor(sample_metadata$sex))
table_upset_analysis = (table_upset_analysis_cont>0)*1
table_upset_analysis = data.frame(table_upset_analysis)
table_upset_analysis = apply(table_upset_analysis,2,as.integer)
rownames(table_upset_analysis) <- rownames(genome_counts_rel_pa)

# === UpSet with Unique vs Shared colors (ComplexUpset) ===

# 1) Presence/absence: genomes (rows) x samples (cols)
pa <- genome_counts_filt %>%
  tibble::column_to_rownames("genome") %>%
  dplyr::mutate(across(everything(), ~ as.integer(. > 0)))

# 2) Align sample order in metadata to PA columns, then aggregate by environment
env_by_sample <- setNames(sample_metadata$broad_environment, sample_metadata$sample)
env_vec <- unname(env_by_sample[colnames(pa)])  # reorder
agg <- aggregate(t(pa), by = list(env_vec), FUN = function(x) as.integer(sum(x) > 0))
mat <- t(as.matrix(agg[, -1]))
colnames(mat) <- agg$Group.1

# 3) Long → short labels
env_map <- c(
  "1000198 - Mixed forest"       = "MF",
  "1000221 - Temperate woodland" = "TW",
  "1000218 - Xeric shrubland"    = "XS",
  "1000245 - Cropland"           = "C"
)
colnames(mat) <- ifelse(colnames(mat) %in% names(env_map),
                        unname(env_map[colnames(mat)]),
                        colnames(mat))

# 4) Final boolean data frame for ComplexUpset
table_upset_analysis <- as.data.frame((mat > 0) * 1)

# 5) Sets & colors
wanted_sets <- c("MF","TW","XS","C")
set_colors  <- c(MF="#f56042", TW="#429ef5", XS="#42f58d", C="#b142f5")
uniqshare_colors <- c(Unique = "#F39C12", Shared = "#7F8C8D")  # tweak if you like

# 6) Plot:

# `table_upset_analysis` is your 0/1 dataframe with columns MF,TW,XS,C
UpSetR::upset(
  table_upset_analysis,
  sets = c("MF","TW","XS","C"),
  keep.order = TRUE,
  sets.bar.color = c("#f56042", "#429ef5", "#42f58d", "#b142f5"),
  order.by = "freq",
  # highlight unique (degree = 1) intersections
  queries = list(
    list(query = intersects, params = list("MF"), color = "#F39C12", active = TRUE),
    list(query = intersects, params = list("TW"), color = "#F39C12", active = TRUE),
    list(query = intersects, params = list("XS"), color = "#F39C12", active = TRUE),
    list(query = intersects, params = list("C"),  color = "#F39C12", active = TRUE)
  )
)

```
## ANCOM-BC2
### Enrichment analysis between male and female

**Note:** Create a filtered phyloseq object excluding structural zeros for ANCOM-BC2 analysis. Replace zeros with a small pseudocount (0.00001) to enable log-transformation.

```{r zero_phylo, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#phyloseq object considering structural zeros
phylo_samples <- sample_metadata %>% 
                    column_to_rownames("sample") %>% 
                    sample_data() #convert to phyloseq sample_data object
phylo_genome <- genome_counts_filt %>% 
                    filter(!genome %in% structural_zeros$genome) %>% # remove structural zeros
                    column_to_rownames("genome") %>% 
                    mutate_all(~ replace(., . == 0, 0.00001)) %>% 
                    otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>% 
                    filter(genome %in% rownames(phylo_genome)) %>% # remove structural zeros
                    mutate(genome2=genome) %>% #create a pseudo genome name column
                    column_to_rownames("genome2") %>% 
                    dplyr::select(domain,phylum,class,order,family,genus,species,genome) %>% #add an additional taxonomic level to ensure genome-level analysis (as not all genomes have species-level taxonomic assignments. Otherwise, ANCOMBC2 aggregates analyses per species)
                    as.matrix() %>% 
                    tax_table() #convert to phyloseq tax_table object

physeq_genome_filtered <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)

```

**Note:** Define helper functions to reduce code duplication for ANCOM-BC2 analyses, taxonomy processing, color schemes, and visualization.

```{r helper_functions_ancombc, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Helper function to run ANCOM-BC2 analysis
run_ancombc2_analysis <- function(data, tax_level = NULL, rand_formula = NULL, 
                                   struc_zero = FALSE, group = NULL, 
                                   lme_control = NULL, seed = 1234) {
  set.seed(seed)
  ancombc2(data = data,
           assay_name = "counts",
           tax_level = tax_level,
           fix_formula = "sex",
           rand_formula = rand_formula,
           p_adj_method = "holm",
           pseudo_sens = TRUE,
           prv_cut = 0,
           lib_cut = 0,
           s0_perc = 0.05,
           group = group,
           struc_zero = struc_zero,
           neg_lb = FALSE,
           alpha = 0.05,
           n_cl = 2,
           verbose = TRUE,
           global = FALSE,
           pairwise = FALSE,
           dunnet = FALSE,
           trend = FALSE,
           iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
           em_control = list(tol = 1e-5, max_iter = 100),
           lme_control = lme_control,
           mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
           trend_control = NULL)
}

# Helper function to clean taxonomy table
clean_taxonomy_table <- function(physeq_obj, include_unclassified = TRUE) {
  taxonomy <- data.frame(physeq_obj@tax_table) %>%
    rownames_to_column(., "taxon") %>%
    mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))
  
  if (include_unclassified) {
    taxonomy <- taxonomy %>%
      mutate(across(c(family, genus, species), na_if, "")) %>%
      mutate(family = coalesce(family, paste('Unclassified', order)),
             genus = coalesce(genus, 
                            if_else(grepl('^Unclassified', family),
                                    family, paste('Unclassified', family))),
             species = coalesce(species, 
                            if_else(grepl('^Unclassified', genus),
                                    genus, paste('Unclassified', genus))))
  }
  return(taxonomy)
}

# Helper function to get phylum colors
get_phylum_colors <- function(taxonomy, add_alpha = TRUE) {
  colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", "")) %>%
    right_join(taxonomy, by = join_by(phylum == phylum)) %>%
    dplyr::select(phylum, colors) %>%
    mutate(colors = if (add_alpha) str_c(colors, "80") else colors) %>%
    unique() %>%
    dplyr::arrange(phylum)
  
  return(colors_alphabetic)
}

# Helper function to extract significant results
extract_significant_results <- function(ancom_output, taxonomy, p_threshold = 0.05) {
  results <- ancom_output$res %>%
    dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
    filter(p_sexMale < p_threshold) %>%
    dplyr::arrange(p_sexMale) %>%
    merge(., taxonomy, by = "taxon") %>%
    mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", "")) %>%
    dplyr::arrange(lfc_sexMale)
  return(results)
}

# Helper function to prepare color vector for plots
prepare_tax_color_vector <- function(results_table, taxonomy) {
  colors_alphabetic <- get_phylum_colors(taxonomy, add_alpha = TRUE)
  tax_table <- as.data.frame(unique(results_table$phylum))
  colnames(tax_table)[1] <- "phylum"
  tax_color <- merge(tax_table, colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()
  return(tax_color)
}

# Helper function to create volcano plot
plot_volcano_ancombc <- function(ancom_output, genome_metadata, genome_tree, 
                                   p_threshold = 0.05) {
  phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by = join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name = phylum)
  
  ancom_output$res %>%
    na.omit() %>%
    dplyr::select(genome = taxon, lfc_sexMale, p_sexMale) %>%
    inner_join(genome_metadata, by = join_by(genome == genome)) %>%
    mutate(phylum = ifelse(p_sexMale < p_threshold, phylum, NA)) %>%
    ggplot(., aes(x = lfc_sexMale, y = -log(p_sexMale), color = phylum)) +
    geom_point(size = 3, show.legend = FALSE) +
    scale_color_manual(values = phylum_colors) +
    labs(color = "Significant phylum", 
         x = "Log-fold difference between sample types", 
         y = "p-value") +
    theme_classic()
}

# Helper function to create bar plot
plot_ancombc_barplot <- function(results_table, tax_color, y_label = "Species") {
  results_table %>%
    mutate(genome = factor(genome, levels = results_table$genome)) %>%
    ggplot(., aes(x = lfc_sexMale, 
                   y = forcats::fct_reorder(genome, lfc_sexMale), 
                   fill = phylum)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    geom_hline(yintercept = 0) +
    theme(panel.background = element_blank(),
          axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_text(size = 8),
          axis.title = element_text(size = 14, face = "bold"),
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14, face = "bold"),
          legend.position = "right", legend.box = "vertical") +
    xlab("log2FoldChange") +
    ylab(y_label) +
    guides(fill = guide_legend(title = "Phylum"))
}
```

### With random effect
#### MAG level

**Note:** Run ANCOM-BC2 analysis at MAG level with random effects for broad_environment to test differential abundance between sexes, accounting for environmental variation.

```{r ancom_rand_pond, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = NULL,
  rand_formula = "(1|broad_environment)",
  struc_zero = FALSE,
  group = NULL,
  lme_control = lme4::lmerControl()
)
```

**Note:** Process ANCOM-BC2 results to extract significant MAGs (p < 0.05), merge with taxonomy information, and prepare color scheme for phylum-level visualization.

```{r ancom_rand, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
taxonomy <- clean_taxonomy_table(physeq_genome_filtered, include_unclassified = TRUE)
ancombc_rand_table_mag <- extract_significant_results(ancom_rand_output, taxonomy)
ancombc_rand_table_mag
tax_color <- prepare_tax_color_vector(ancombc_rand_table_mag, taxonomy)
```
**Note:** Create a bar plot showing log2 fold changes for significantly differentially abundant MAGs between sexes, colored by phylum.

```{r ancombc_rand_plot_phy, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, }
plot_ancombc_barplot(ancombc_rand_table_mag, tax_color, y_label = "Species")
```


**Note:** Create a volcano plot showing log-fold change vs. p-value for all MAGs, with significant phyla highlighted in color.

```{r ancom_rand_volcano, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
plot_volcano_ancombc(ancom_rand_output, genome_metadata, genome_tree)
```

#### Controlling structural components

**Note:** Run ANCOM-BC2 analysis at MAG level with random effects and structural zero detection enabled, grouping by sex to identify taxa that are completely absent in one group.

```{r ancom_rand_season, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_mag_Struc <- run_ancombc2_analysis(
  data = physeq_all,
  tax_level = NULL,
  rand_formula = "(1|broad_environment)",
  struc_zero = TRUE,
  group = "sex",
  lme_control = lme4::lmerControl()
)
```

**Note:** Process ANCOM-BC2 results with structural zero control, extract significant MAGs, merge with taxonomy, and prepare phylum color scheme for visualization.

```{r ancom_rand1, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
taxonomy <- clean_taxonomy_table(physeq_genome_filtered, include_unclassified = FALSE)
ancombc_rand_table_mag_Struc <- extract_significant_results(ancom_rand_output_mag_Struc, taxonomy)
ancombc_rand_table_mag_Struc
tax_color <- prepare_tax_color_vector(ancombc_rand_table_mag_Struc, taxonomy)
```
**Note:** Create a bar plot showing log2 fold changes for significantly differentially abundant MAGs (with structural zero control) between sexes, colored by phylum.

```{r ancombc_rand_plot_mag, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, }
plot_ancombc_barplot(ancombc_rand_table_mag_Struc, tax_color, y_label = "Species")
```

**Note:** Create a volcano plot for ANCOM-BC2 results with structural zero control, showing log-fold change vs. p-value with significant phyla highlighted.

```{r ancom_rand_volcano_mag, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
plot_volcano_ancombc(ancom_rand_output_mag_Struc, genome_metadata, genome_tree)
```

#### Phylum level

**Note:** Run ANCOM-BC2 analysis aggregated at phylum level with random effects for broad_environment to test differential abundance between sexes.

```{r ancom_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_phylum <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = "phylum",
  rand_formula = "(1|broad_environment)",
  struc_zero = FALSE,
  group = NULL,
  lme_control = lme4::lmerControl()
)
```

**Note:** Extract and display significantly differentially abundant phyla (p < 0.05) from the ANCOM-BC2 results, sorted by p-value.

```{r ancom_rand_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_phylum$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Family level

**Note:** Run ANCOM-BC2 analysis aggregated at family level with random effects for broad_environment to test differential abundance between sexes.

```{r ancom_family, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_family <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = "family",
  rand_formula = "(1|broad_environment)",
  struc_zero = FALSE,
  group = NULL,
  lme_control = lme4::lmerControl()
)
```

**Note:** Extract and display significantly differentially abundant families (p < 0.05) from the ANCOM-BC2 results, sorted by p-value.

```{r ancom_rand_family, comment="", echo=FALSE, message=FALSE, warning=FALSE}

ancom_rand_output_family$res %>% 
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Genera

**Note:** Run ANCOM-BC2 analysis aggregated at genus level with random effects for broad_environment to test differential abundance between sexes.

```{r ancom_genera, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_gen <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = "genus",
  rand_formula = "(1|broad_environment)",
  struc_zero = FALSE,
  group = NULL,
  lme_control = lme4::lmerControl()
)
```

**Note:** Extract and display significantly differentially abundant genera (p < 0.05) from the ANCOM-BC2 results, sorted by p-value.

```{r ancom_rand_genera, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_gen$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```



### Without random effect
#### MAG level

**Note:** Run ANCOM-BC2 analysis at MAG level without random effects to test differential abundance between sexes (chunk is set to eval=FALSE, not executed).

```{r ancom_rand_pond2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
ancom_rand_output_norand <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = NULL,
  rand_formula = NULL,
  struc_zero = FALSE,
  group = NULL,
  lme_control = NULL
)
```

**Note:** Process ANCOM-BC2 results without random effects, extract significant MAGs, merge with taxonomy, and prepare phylum color scheme (chunk is set to eval=FALSE, not executed).

```{r ancom_rand2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
taxonomy <- clean_taxonomy_table(physeq_genome_filtered, include_unclassified = TRUE)
ancombc_rand_table_mag_norand <- extract_significant_results(ancom_rand_output_norand, taxonomy)
tax_color <- prepare_tax_color_vector(ancombc_rand_table_mag_norand, taxonomy)
```

**Note:** Create a bar plot for ANCOM-BC2 results without random effects showing log2 fold changes for significant MAGs (chunk is set to eval=FALSE, not executed).

```{r ancombc_rand_plot_mag2, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
plot_ancombc_barplot(ancombc_rand_table_mag_norand, tax_color, y_label = "Species")
```


**Note:** Create a volcano plot for ANCOM-BC2 results without random effects (chunk is set to eval=FALSE, not executed).

```{r ancom_rand_volcano2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
plot_volcano_ancombc(ancom_rand_output_norand, genome_metadata, genome_tree)
```

#### Controlling structural components

**Note:** Run ANCOM-BC2 analysis at MAG level without random effects but with structural zero detection enabled, grouping by sex.

```{r ancom_rand_struct12, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_mag_Struc_norand <- run_ancombc2_analysis(
  data = physeq_all,
  tax_level = NULL,
  rand_formula = NULL,
  struc_zero = TRUE,
  group = "sex",
  lme_control = NULL
)
```

**Note:** Process ANCOM-BC2 results without random effects but with structural zero control, extract significant MAGs, merge with taxonomy, and prepare phylum color scheme.

```{r ancom_rand12, comment="", echo=FALSE, message=FALSE, warning=FALSE,  }
taxonomy <- clean_taxonomy_table(physeq_genome_filtered, include_unclassified = FALSE)
ancombc_rand_table_mag_Struc_norand <- extract_significant_results(ancom_rand_output_mag_Struc_norand, taxonomy)
tax_color <- prepare_tax_color_vector(ancombc_rand_table_mag_Struc_norand, taxonomy)
```

**Note:** Create a bar plot showing log2 fold changes for significantly differentially abundant MAGs (without random effects, with structural zero control) between sexes, colored by phylum.

```{r ancombc_rand_plot_mag12, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, }
plot_ancombc_barplot(ancombc_rand_table_mag_Struc_norand, tax_color, y_label = "Species")
```

**Note:** Create a volcano plot for ANCOM-BC2 results without random effects but with structural zero control, showing log-fold change vs. p-value with significant phyla highlighted.

```{r ancom_rand_volcano_mag12, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
plot_volcano_ancombc(ancom_rand_output_mag_Struc_norand, genome_metadata, genome_tree)
```

#### Phylum level

**Note:** Run ANCOM-BC2 analysis aggregated at phylum level without random effects to test differential abundance between sexes.

```{r ancom_phylum2, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_phylum_norand <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = "phylum",
  rand_formula = NULL,
  struc_zero = FALSE,
  group = NULL,
  lme_control = NULL
)
```

**Note:** Extract and display significantly differentially abundant phyla (p < 0.05) from ANCOM-BC2 results without random effects, sorted by p-value.

```{r ancom_rand_phylum2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
ancom_rand_output_phylum_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Family level

**Note:** Run ANCOM-BC2 analysis aggregated at family level without random effects to test differential abundance between sexes.

```{r ancom_family2, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_family_norand <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = "family",
  rand_formula = NULL,
  struc_zero = FALSE,
  group = NULL,
  lme_control = NULL
)
```

**Note:** Extract and display significantly differentially abundant families (p < 0.05) from ANCOM-BC2 results without random effects, sorted by p-value.

```{r ancom_rand_family2, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_family_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Genera

**Note:** Run ANCOM-BC2 analysis aggregated at genus level without random effects to test differential abundance between sexes.

```{r ancom_genera2, comment="", echo=FALSE, message=FALSE, warning=FALSE,  }
ancom_rand_output_gen_norand <- run_ancombc2_analysis(
  data = physeq_genome_filtered,
  tax_level = "genus",
  rand_formula = NULL,
  struc_zero = FALSE,
  group = NULL,
  lme_control = NULL
)
```

**Note:** Extract and display significantly differentially abundant genera (p < 0.05) from ANCOM-BC2 results without random effects, sorted by p-value.

```{r ancom_rand_genera2, comment="", echo=FALSE, message=FALSE, warning=FALSE,  }
ancom_rand_output_gen_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

**Note:** Save all ANCOM-BC2 analysis results to an Rdata file for later use.

```{r ancom_save, comment="", echo=FALSE, message=FALSE, warning=FALSE }
 save(ancom_rand_output,
     ancom_rand_output_mag_Struc,
     ancom_rand_output_phylum,
     ancom_rand_output_family,
     ancom_rand_output_gen,
     #ancom_rand_output_norand,
     ancom_rand_output_mag_Struc_norand,
     ancom_rand_output_phylum_norand,
     ancom_rand_output_family_norand,
     ancom_rand_output_gen_norand,
     file="data/ancombc_all.Rdata")
```