
# Differential abundance analysis

ANCOM-BC2-style models test which taxa change with sex, with and without random effects and while accounting for structural zeros. Results are summarised at MAG, phylum, family, and genus levels.

```{r load_data_mag_filtdamr_diffe, comment="", echo=FALSE, message=FALSE, warning=FALSE}

load("data/data.Rdata")
```

```{r exclude Sample7, message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
valid_samples <- sample_metadata$sample

genome_counts_filt <- genome_counts_filt %>%
  dplyr::select(genome, tidyselect::any_of(valid_samples))
```

```{r genom_count_filt, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))
```

```{r phyloseq, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
#Phyloseq object
count_phy <- genome_counts_filt %>%
  column_to_rownames(var="genome")%>%
  otu_table(., taxa_are_rows=T)

sample_info_tab_phy <- sample_metadata%>%
  column_to_rownames(var="sample")%>%
  sample_data()

TAX <- genome_metadata%>%
  column_to_rownames(var="genome")%>%
  select(1:7)%>%
  as.matrix()%>%
  tax_table()
tree <- phy_tree(genome_tree)

physeq_all = phyloseq(count_phy, TAX, sample_info_tab_phy, tree)
```

## Structural zeros

```{r struct_zero, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# groups
Female_samples <- sample_metadata %>% filter(sex == "Female") %>% pull(sample)
Male_samples   <- sample_metadata %>% filter(sex == "Male")   %>% pull(sample)

# keep only samples that exist in the counts
existing_samples <- colnames(genome_counts_filt)
Female_samples <- intersect(Female_samples, existing_samples)
Male_samples   <- intersect(Male_samples,   existing_samples)

# compute structural zeros row-wise
structural_zeros <- genome_counts_filt %>% 
  rowwise() %>%
  mutate(
    all_zeros_Female = all(c_across(all_of(Female_samples)) == 0),
    all_zeros_Male   = all(c_across(all_of(Male_samples))   == 0),
    average_Female   = mean(c_across(all_of(Female_samples)), na.rm = TRUE),
    average_Male     = mean(c_across(all_of(Male_samples)),   na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(all_zeros_Female | all_zeros_Male) %>%
  mutate(
    # present only in the *other* group (the one that is not all zeros)
    present = dplyr::case_when(
      all_zeros_Female & !all_zeros_Male ~ "Male",
      all_zeros_Male   & !all_zeros_Female ~ "Female",
      TRUE ~ NA_character_
    ),
    average = dplyr::case_when(
      present == "Female" ~ average_Female,
      present == "Male"   ~ average_Male,
      TRUE ~ NA_real_
    )
  )

# attach taxonomy
structural_zeros <- structural_zeros %>%
  dplyr::left_join(genome_metadata %>% dplyr::select(genome, phylum), by = "genome")

# normalize possible taxonomic prefixes on both sides
structural_zeros <- structural_zeros %>%
  dplyr::mutate(phylum = stringr::str_replace(phylum, "^[dpcofgs]__", ""))

```

```{r structu_Zero_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#top 30 strongest “present-only-in” genomes
structural_zeros_top <- structural_zeros %>%
  dplyr::mutate(abs_avg = abs(average)) %>%
  dplyr::arrange(dplyr::desc(abs_avg)) %>%
  dplyr::slice_head(n = 30)

xmax <- max(structural_zeros_top$abs_avg, na.rm = TRUE) * 1.1

ggplot(structural_zeros_top,
       aes(x = average,
           y = forcats::fct_reorder(genome, average),
           fill = present)) +
  geom_col(width = 0.8) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c(Female = "#C33C54", Male = "#2A9D8F"),
                    name = "Present only in") +
  coord_cartesian(xlim = c(-xmax, xmax), expand = FALSE) +
  labs(x = "Genome counts (negative = Male only, positive = Female only)",
       y = "MAG (top 30 by |difference|)") +
  theme_classic() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 6),
        plot.margin = margin(6, 6, 6, 6))

```

```{r chart2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#phylum overview of "present only in"
phylum_summary_sz <- structural_zeros %>%
  dplyr::filter(!is.na(phylum), !is.na(present)) %>%
  dplyr::count(phylum, present, name = "n") %>%
  tidyr::complete(phylum, present = c("Female","Male"), fill = list(n = 0)) %>%
  dplyr::group_by(phylum) %>%
  dplyr::mutate(total = sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(dplyr::desc(total))

ggplot(phylum_summary_sz,
       aes(x = reorder(phylum, total), y = n, fill = present)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c(Female = "#C33C54", Male = "#2A9D8F"),
                    name = "Present only in") +
  labs(x = "Phylum", y = "Number of MAGs present only in group") +
  theme_classic() +
  theme(legend.position = "top")

## MAGs in different locations and shared among locations
```

```{r chart1, comment="", echo=FALSE, message=FALSE, warning=FALSE}
locationcolors=c('#408892','#c04062')
locationcolors=c('#408892','#c04062')

genome_counts_rel <- genome_counts %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  column_to_rownames(., "genome")

genome_counts_rel_fil<- genome_counts_filt%>% 
    select(one_of(c("genome",sample_metadata$sample))) %>% 
  column_to_rownames(., "genome")
  
genome_counts_rel_pa=1*(genome_counts_rel_fil>0)
#MAGrel_pa[1:6,1:6]
table_upset_analysis_cont=t(aggregate(t(genome_counts_rel_pa),by=list(sample_metadata$sex),FUN=sum)[,-1])
colnames(table_upset_analysis_cont)=levels(as.factor(sample_metadata$sex))
table_upset_analysis=(table_upset_analysis_cont>0)*1
table_upset_analysis=data.frame(table_upset_analysis)
table_upset_analysis=apply(table_upset_analysis,2,as.integer)
rownames(table_upset_analysis) <- rownames(genome_counts_rel_pa)

# === UpSet with Unique vs Shared colors (ComplexUpset) ===
if (!requireNamespace("ComplexUpset", quietly = TRUE)) install.packages("ComplexUpset")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
library(ComplexUpset)
library(ggplot2)

# 1) Presence/absence: genomes (rows) x samples (cols)
pa <- genome_counts_filt %>%
  tibble::column_to_rownames("genome") %>%
  dplyr::mutate(across(everything(), ~ as.integer(. > 0)))

# 2) Align sample order in metadata to PA columns, then aggregate by environment
env_by_sample <- setNames(sample_metadata$broad_environment, sample_metadata$sample)
env_vec <- unname(env_by_sample[colnames(pa)])  # reorder
agg <- aggregate(t(pa), by = list(env_vec), FUN = function(x) as.integer(sum(x) > 0))
mat <- t(as.matrix(agg[, -1])); colnames(mat) <- agg$Group.1

# 3) Long → short labels
env_map <- c(
  "1000198 - Mixed forest"       = "MF",
  "1000221 - Temperate woodland" = "TW",
  "1000218 - Xeric shrubland"    = "XS",
  "1000245 - Cropland"           = "C"
)
colnames(mat) <- ifelse(colnames(mat) %in% names(env_map),
                        unname(env_map[colnames(mat)]),
                        colnames(mat))

# 4) Final boolean data frame for ComplexUpset
table_upset_analysis <- as.data.frame((mat > 0) * 1)

# 5) Sets & colors
wanted_sets <- c("MF","TW","XS","C")
set_colors  <- c(MF="#f56042", TW="#429ef5", XS="#42f58d", C="#b142f5")
uniqshare_colors <- c(Unique = "#F39C12", Shared = "#7F8C8D")  # tweak if you like

# 6) Plot:
# ensure UpSetR
if (!requireNamespace("UpSetR", quietly = TRUE)) install.packages("UpSetR")
library(UpSetR)

# `table_upset_analysis` is your 0/1 dataframe with columns MF,TW,XS,C
UpSetR::upset(
  table_upset_analysis,
  sets = c("MF","TW","XS","C"),
  keep.order = TRUE,
  sets.bar.color = c("#f56042", "#429ef5", "#42f58d", "#b142f5"),
  order.by = "freq",
  # highlight unique (degree = 1) intersections
  queries = list(
    list(query = intersects, params = list("MF"), color = "#F39C12", active = TRUE),
    list(query = intersects, params = list("TW"), color = "#F39C12", active = TRUE),
    list(query = intersects, params = list("XS"), color = "#F39C12", active = TRUE),
    list(query = intersects, params = list("C"),  color = "#F39C12", active = TRUE)
  )
)

```
## Ancomb2
### Enrichment analysis between male and female

```{r zero_phylo, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#phyloseq object considering structual zeros
phylo_samples <- sample_metadata %>% 
                    column_to_rownames("sample") %>% 
                    sample_data() #convert to phyloseq sample_data object
phylo_genome <- genome_counts_filt %>% 
                    filter(!genome %in% structural_zeros$genome) %>% # remove structural zeros
                    column_to_rownames("genome") %>% 
                    mutate_all(~ replace(., . == 0, 0.00001)) %>% 
                    otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>% 
                    filter(genome %in% rownames(phylo_genome)) %>% # remove structural zeros
                    mutate(genome2=genome) %>% #create a pseudo genome name column
                    column_to_rownames("genome2") %>% 
                    dplyr::select(domain,phylum,class,order,family,genus,species,genome) %>% #add an additional taxonomic level to ensure genome-level analysis (as no all genomes have species-level taxonomic assignments. Otherwise, ANCOMBC2 aggregates analyses per species)
                    as.matrix() %>% 
                    tax_table() #convert to phyloseq tax_table object

physeq_genome_filtered <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)

```

### With random effect
#### MAG level

```{r ancom_rand_pond, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
set.seed(1234) #set seed for reproducibility
ancom_rand_output = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = NULL, #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
                  rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  s0_perc = 0.05,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```

```{r ancom_rand, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
taxonomy <- data.frame(physeq_genome_filtered@tax_table) %>%
  rownames_to_column(., "taxon") %>%
  mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  mutate(across(c(family,genus, species), na_if, ""))%>%
    mutate(family = coalesce(family, paste('Unclassified', order)),
           genus = coalesce(genus, 
                              if_else(grepl('^Unclassified', family),
                                      family, paste('Unclassified', family))),
           species = coalesce(species, 
                              if_else(grepl('^Unclassified', genus),
                                      genus, paste('Unclassified', genus))))

ancombc_rand_table_mag <- ancom_rand_output$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) %>%
  merge(., taxonomy, by="taxon") %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_sexMale)

ancombc_rand_table_mag
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```
```{r ancombc_rand_plot_phy, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, }
ancombc_rand_table_mag%>%
      mutate(genome=factor(genome,levels=ancombc_rand_table_mag$genome)) %>%
ggplot(., aes(x=lfc_sexMale, y=forcats::fct_reorder(genome,lfc_sexMale), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
#  coord_flip()+
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```


```{r ancom_rand_volcano, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

#pdf("figures/different_species_StrucZero_new_violin.pdf",width=12, height=6)
ancom_rand_output$res %>%
  na.omit() %>%
  dplyr::select(genome=taxon, lfc_sexMale, p_sexMale) %>%
  inner_join(genome_metadata, by = join_by(genome == genome)) %>%
  mutate(phylum = ifelse(p_sexMale < 0.05, phylum, NA)) %>%
  ggplot(., aes(x = lfc_sexMale, y = -log(p_sexMale), color = phylum)) +
  geom_point(size=3, show.legend = FALSE) +
  #xlim(c(-10,4)) +
  scale_color_manual(values = phylum_colors) +
  labs(color = "Significant phylum", x = "Log-fold difference between sample types", y = "p-value") +
  theme_classic()
#dev.off()
```

#### Controling structural components
```{r ancom_rand_season, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
set.seed(1234) #set seed for reproducibility
 ancom_rand_output_mag_Struc = ancombc2(data = physeq_all, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "sex", 
                  rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "sex", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```

```{r ancom_rand1, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
 taxonomy <- data.frame(physeq_genome_filtered@tax_table) %>%
  rownames_to_column(., "taxon")%>%
  mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))

ancombc_rand_table_mag_Struc <- ancom_rand_output_mag_Struc$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) %>%
  merge(., taxonomy, by="taxon") %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_sexMale)

ancombc_rand_table_mag_Struc

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag_Struc$phylum))
  
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```
```{r ancombc_rand_plot_mag, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, }
 ancombc_rand_table_mag_Struc%>%
      mutate(genome=factor(genome,levels=ancombc_rand_table_mag_Struc$genome)) %>%
ggplot(., aes(x=lfc_sexMale, y=forcats::fct_reorder(genome,lfc_sexMale), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
#  coord_flip()+
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```

```{r ancom_rand_volcano_mag, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

#pdf("figures/different_species_StrucZero_new_violin.pdf",width=12, height=6)
ancom_rand_output_mag_Struc$res %>%
  na.omit() %>%
  dplyr::select(genome=taxon, lfc_sexMale, p_sexMale) %>%
  inner_join(genome_metadata, by = join_by(genome == genome)) %>%
  mutate(phylum = ifelse(p_sexMale < 0.05, phylum, NA)) %>%
  ggplot(., aes(x = lfc_sexMale, y = -log(p_sexMale), color = phylum)) +
  geom_point(size=3, show.legend = FALSE) +
  #xlim(c(-10,4)) +
  scale_color_manual(values = phylum_colors) +
  labs(color = "Significant phylum", x = "Log-fold difference between sample types", y = "p-value") +
  theme_classic()
#dev.off()
```

#### Phylum level

```{r ancom_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE, }

ancom_rand_output_phylum = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = "phylum", #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
                  rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut = 0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r ancom_rand_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_phylum$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Family level
```{r ancom_family, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_family = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = "family", #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
                  rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut = 0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r ancom_rand_family, comment="", echo=FALSE, message=FALSE, warning=FALSE}

ancom_rand_output_family$res %>% 
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Genera
```{r ancom_genera, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_gen = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = "genus", #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
                  rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut = 0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r ancom_rand_genera, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_gen$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```



### Without random effect
#### MAG level

```{r ancom_rand_pond2, comment="", echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE }
set.seed(1234) #set seed for reproducibility
ancom_rand_output_norand = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = NULL, #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
                  rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL,
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```

```{r ancom_rand2, comment="", echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE}
taxonomy <- data.frame(physeq_genome_filtered@tax_table) %>%
  rownames_to_column(., "taxon") %>%
  mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  mutate(across(c(family,genus, species), na_if, ""))%>%
    mutate(family = coalesce(family, paste('Unclassified', order)),
           genus = coalesce(genus, 
                              if_else(grepl('^Unclassified', family),
                                      family, paste('Unclassified', family))),
           species = coalesce(species, 
                              if_else(grepl('^Unclassified', genus),
                                      genus, paste('Unclassified', genus))))

ancombc_rand_table_mag_norand <- ancom_rand_output_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) %>%
  merge(., taxonomy, by="taxon") %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_sexMale)
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag_norand$phylum))
  
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```

```{r ancombc_rand_plot_mag2, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, eval = FALSE }
ancombc_rand_table_mag_norand%>%
      mutate(genome=factor(genome,levels=ancombc_rand_table_mag_norand$genome)) %>%
ggplot(., aes(x=lfc_sexMale, y=forcats::fct_reorder(genome,lfc_sexMale), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
#  coord_flip()+
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```


```{r ancom_rand_volcano2, comment="", echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

#pdf("figures/different_species_StrucZero_new_violin.pdf",width=12, height=6)
ancom_rand_output_norand$res %>%
  na.omit() %>%
  dplyr::select(genome=taxon, lfc_sexMale, p_sexMale) %>%
  inner_join(genome_metadata, by = join_by(genome == genome)) %>%
  mutate(phylum = ifelse(p_sexMale < 0.05, phylum, NA)) %>%
  ggplot(., aes(x = lfc_sexMale, y = -log(p_sexMale), color = phylum)) +
  geom_point(size=3, show.legend = FALSE) +
  #xlim(c(-10,4)) +
  scale_color_manual(values = phylum_colors) +
  labs(color = "Significant phylum", x = "Log-fold difference between sample types", y = "p-value") +
  theme_classic()
#dev.off()
```

#### Controling structural components
```{r ancom_rand_struct12, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_mag_Struc_norand = ancombc2(data = physeq_all, 
                  assay_name = "counts", 
                  tax_level = NULL, #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
 #                 rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "sex", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL,
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```

```{r ancom_rand12, comment="", echo=FALSE, message=FALSE, warning=FALSE,  }
taxonomy <- data.frame(physeq_genome_filtered@tax_table) %>%
  rownames_to_column(., "taxon")%>%
  mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  mutate(across(c(family,genus, species), na_if, ""))

ancombc_rand_table_mag_Struc_norand <- ancom_rand_output_mag_Struc_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) %>%
  merge(., taxonomy, by="taxon") %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_sexMale)
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag_Struc_norand$phylum))
  
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```

```{r ancombc_rand_plot_mag12, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE, }
ancombc_rand_table_mag_Struc_norand%>%
      mutate(genome=factor(genome,levels=ancombc_rand_table_mag_Struc_norand$genome)) %>%
ggplot(., aes(x=lfc_sexMale, y=forcats::fct_reorder(genome,lfc_sexMale), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
#  coord_flip()+
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```

```{r ancom_rand_volcano_mag12, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

#pdf("figures/different_species_StrucZero_new_violin.pdf",width=12, height=6)
ancom_rand_output_mag_Struc_norand$res %>%
  na.omit() %>%
  dplyr::select(genome=taxon, lfc_sexMale, p_sexMale) %>%
  inner_join(genome_metadata, by = join_by(genome == genome)) %>%
  mutate(phylum = ifelse(p_sexMale < 0.05, phylum, NA)) %>%
  ggplot(., aes(x = lfc_sexMale, y = -log(p_sexMale), color = phylum)) +
  geom_point(size=3, show.legend = FALSE) +
  #xlim(c(-10,4)) +
  scale_color_manual(values = phylum_colors) +
  labs(color = "Significant phylum", x = "Log-fold difference between sample types", y = "p-value") +
  theme_classic()
#dev.off()
```

#### Phylum level

```{r ancom_phylum2, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_phylum_norand = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = "phylum", #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
  #                rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut = 0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL,
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r ancom_rand_phylum2, comment="", echo=FALSE, message=FALSE, warning=FALSE}
ancom_rand_output_phylum_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Family level
```{r ancom_family2, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_family_norand = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = "family", #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
   #               rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut = 0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL,
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r ancom_rand_family2, comment="", echo=FALSE, message=FALSE, warning=FALSE, }
ancom_rand_output_family_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

#### Genera
```{r ancom_genera2, comment="", echo=FALSE, message=FALSE, warning=FALSE,  }
ancom_rand_output_gen_norand = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = "genus", #change to agglomerate analysis to a higher taxonomic range
                  fix_formula = "sex", #fixed variable(s)
 #                 rand_formula = "(1|broad_environment)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut = 0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL,
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r ancom_rand_genera2, comment="", echo=FALSE, message=FALSE, warning=FALSE,  }
ancom_rand_output_gen_norand$res %>%
  dplyr::select(taxon, lfc_sexMale, p_sexMale) %>%
  filter(p_sexMale < 0.05) %>%
  dplyr::arrange(p_sexMale) 
```

```{r ancom_save, comment="", echo=FALSE, message=FALSE, warning=FALSE }
 save(ancom_rand_output,
     ancom_rand_output_mag_Struc,
     ancom_rand_output_phylum,
     ancom_rand_output_family,
     ancom_rand_output_gen,
     #ancom_rand_output_norand,
     ancom_rand_output_mag_Struc_norand,
     ancom_rand_output_phylum_norand,
     ancom_rand_output_family_norand,
     ancom_rand_output_gen_norand,
     file="data/ancombc_all.Rdata")
```