#Enviromentypes

##Landcover percentage

# Load libraries

# required libraries
```{r load_libraries_enviroment, warning=FALSE, comments="", message=FALSE}      
library(sf)
library(terra)
library(exactextractr)
```

# 1. sample points
```{r sample_points, warning=FALSE, comments="", message=FALSE}
sample_points <- read.csv("sample_points.csv")
points_sf <- st_as_sf(sample_points, coords = c("longitude", "latitude"), crs = 4326)
points_proj <- st_transform(points_sf, 28355)  # GDA94 / MGA Zone 55 — meters
```

# 2. 150-meter buffer zones
```{r buffer_zones, warning=FALSE, comment="", message=FALSE, eval=FALSE}
   buffers <- st_buffer(points_proj, dist = 150)

landcover <- rast("Merged.tif")  # adjust if name differs
landcover <- project(landcover, "EPSG:28355")  # match CRS to buffer layer

extracted <- exact_extract(landcover, buffers)

summary_list <- lapply(seq_along(extracted), function(i) {
  df <- extracted[[i]]
  agg <- aggregate(df$coverage_fraction, by = list(class = round(df$value)), sum)
  colnames(agg) <- c("land_cover_class", "fraction")
  agg$percent <- round(100 * agg$fraction / sum(agg$fraction), 2)
  agg$id <- sample_points$id[i]
  agg
})
```


# 3. Combine buffer summaries with landcover classes
```{r buffer_table, warning=FALSE, comments="", message=FALSE}
result_df <- do.call(rbind, summary_list)

# Load legend CSV
legend <- read.csv("C:\\Users\\Lukas\\Documents\\GitHub\\btp_metagenomics\\Full_Legend_with_Simplified_Labels.csv")
groupinglabels <- read.csv("C:/Users/Lukas/Documents/GitHub/btp_metagenomics/Detailed_Land_Cover_Groupings.csv")



result_df <- result_df %>%
  left_join(legend[, c("land_cover_class", "land_cover_description")], by = "land_cover_class")

result_df <- result_df %>%
  left_join(groupinglabels[, c("land_cover_description", "detailed_category_group")], by = "land_cover_description")


```

#4 Production of contious variabels  (TODO : Check for the NAs in the landcover - where do they come from)
```{r continous variables, warning=FALSE, comments="", message=FALSE}

# Remove rows with missing land_cover_description
result_df_clean <- result_df %>%
  filter(!is.na(land_cover_description))

# Pivot to wide format using full land cover descriptions
landcover_wide <- result_df_clean %>%
  pivot_wider(
    id_cols = id,
    names_from = land_cover_description,
    values_from = percent,
    values_fill = list(percent = 0)
  )

# Make sure all percent columns are numeric
landcover_wide <- landcover_wide %>%
  mutate(across(-id, ~ as.numeric(unlist(.))))

# Create continuous summary variables per buffer
landcover_wide <- landcover_wide %>%
  mutate(
    Cultivated_Total          = rowSums(select(., starts_with("Cultivated"))),
    Native_Terrestrial_Total  = rowSums(select(., starts_with("Natural Terrestrial Vegetated"))),
    Native_Aquatic_Total      = rowSums(select(., starts_with("Natural Aquatic Vegetated"))),
    Water      = rowSums(select(., starts_with("Water"))),
    Bare_Urban                = rowSums(select(., starts_with("Natural Surface")), na.rm = TRUE)
  )

```

#5 Visualisation
```{r stackedbarplot, warning=FALSE, comments="", message=FALSE}

# Set order of categories
result_df$detailed_category_group <- factor(
  result_df$detailed_category_group,
  levels = c(
    "Cultivated Vegetation - Woody",
    "Cultivated Vegetation - Herbaceous",
    "Natural Vegetation - Woody",
    "Natural Vegetation - Herbaceous",
    "Natural Vegetation - General",
    "Natural Vegetation - Sparse",
    "Bare Surface",
    "Urban/Artificial",
    "Aquatic Vegetation",
    "Water",
    "Other"
  )
)

# Define colors (same order as above)
group_colors <- c(
  "Cultivated Vegetation - Woody"      = "#6a51a3",
  "Cultivated Vegetation - Herbaceous" = "#9e9ac8",

  "Natural Vegetation - Woody"         = "#006400",
  "Natural Vegetation - Herbaceous"    = "#1b9e77",
  "Natural Vegetation - Sparse"        = "#e6c229",  
  "Natural Vegetation - General"       = "#b2df8a",

  "Aquatic Vegetation"                 = "#41b6c4",
  "Bare Surface"                       = "#e6ab02",
  "Urban/Artificial"                   = "#757575",
  "Water"                              = "#1f78b4",
  "Other"                              = "#cccccc"
)





# Generate the plot
p <- ggplot(result_df, aes(x = id, y = percent, fill = detailed_category_group)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  scale_fill_manual(values = group_colors, na.value = "gray90") +
  labs(
    title = "Land Cover Composition per Sample",
    x = "Sample ID",
    y = "Land Cover (%)",
    fill = "Land Cover Category"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    panel.grid.major.x = element_blank()
  )

print(p)

# Save high-res output
ggsave("land_cover_composition_per_sample.png", p, width = 12, height = 6, dpi = 300)

```
#Merging Metadata with Landcover



# Join with landcover data by sample ID
```{r fusing, warning=FALSE, comments="", message=FALSE}
NaturalvsCultivated <- left_join(sample_metadata, landcover_wide, by = c("sample" = "id"))
```


```{r Cultivated_diversity, warning=FALSE, comments="", message=FALSE}
ggplot(NaturalvsCultivated, aes(x = Cultivated_Total, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Microbiome Diversity vs. Cultivated Land")
```

```{r Native_diversity, warning=FALSE, comments="", message=FALSE}
ggplot(NaturalvsCultivated, aes(x = Native_Terrestrial_Total, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Microbiome Diversity vs. Native_Land")
```

```{r Cultivated_diversity2, warning=FALSE, comments="", message=FALSE}
ggplot(NaturalvsCultivated, aes(x = Bare_Urban, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Microbiome Diversity vs. Urban_Land")
```

```{r model_diversitys, warning=FALSE, comments="", message=FALSE}
lm_mod <- lm(diversity ~ Cultivated_Total + Native_Terrestrial_Total + Bare_Urban, data = NaturalvsCultivated)
summary(lm_mod)



ggplot(NaturalvsCultivated, aes(x = broad_environment, y = diversity)) +
  geom_boxplot() +
  theme_minimal()
```

```{r model_diversitys_2, warning=FALSE, comments="", message=FALSE}
NaturalvsCultivated$native_group <- cut(
  NaturalvsCultivated$Native_Terrestrial_Total,
  breaks = c(-1, 25, 75, 100),
  labels = c("Low", "Medium", "High")
)

ggplot(NaturalvsCultivated, aes(x = native_group, y = diversity)) +
  geom_boxplot(fill = "#1b9e77") +
  theme_minimal() +
  labs(title = "Diversity by Native Vegetation Group", x = "Native Cover Group", y = "Diversity")

aov_model <- aov(diversity ~ native_group, data = NaturalvsCultivated)
summary(aov_model)
```

```{r model_diversitys_3, warning=FALSE, comments="", message=FALSE}
NaturalvsCultivated$native_group <- cut(
  NaturalvsCultivated$Cultivated_Total,
  breaks = c(-1, 25, 75, 100),
  labels = c("Low", "Medium", "High")
)

ggplot(NaturalvsCultivated, aes(x = native_group, y = diversity)) +
  geom_boxplot(fill = "#1b9e77") +
  theme_minimal() +
  labs(title = "Diversity by Native Vegetation Group", x = "Native Cover Group", y = "Diversity")

aov_model <- aov(diversity ~ native_group, data = NaturalvsCultivated)
summary(aov_model)
```

```{r continous_enviroment, warning=FALSE, comments="", message=FALSE}
# Select only percent cover columns
landcover_numeric <- landcover_wide %>%
  select(Cultivated_Total, Native_Terrestrial_Total, Native_Aquatic_Total, Bare_Urban, Water) %>%
  na.omit()  # Ensure no NAs

# Standardize (mean = 0, sd = 1) before PCA
landcover_scaled <- scale(landcover_numeric)

# Run PCA
pca_result <- prcomp(landcover_scaled)

# Inspect variance explained
summary(pca_result)

# Add first PC (environmental gradient) to landcover data
landcover_wide$EnvGradient_PC1 <- pca_result$x[, 1]
```

```{r continous_enviroment_2, warning=FALSE, comments="", message=FALSE}

# Merge PCA-derived gradient to microbiome data
NaturalvsCultivated <- left_join(sample_metadata, landcover_wide, by = c("sample" = "id"))


ggplot(NaturalvsCultivated, aes(x = EnvGradient_PC1, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Microbiome Diversity vs. Environmental Gradient (PC1)",
       x = "Environmental Gradient (PC1)",
       y = "Diversity") +
  theme_minimal()
```
```{r continous_woody_herbacous, warning=FALSE, comments="", message=FALSE}

# === 1. Define column groups ===

# Columns that represent woody vegetation cover
woody_cols <- c(
  "Natural Terrestrial Vegetated: Woody Closed (> 65 %)",
  "Natural Terrestrial Vegetated: Woody Open (40 to 65 %)",
  "Natural Terrestrial Vegetated: Woody Sparse (4 to 15 %)",
  "Natural Terrestrial Vegetated: Woody Scattered (1 to 4 %)"
)

# Columns that represent herbaceous vegetation cover
herbaceous_cols <- c(
  "Natural Terrestrial Vegetated: Herbaceous Closed (> 65 %)",
  "Natural Terrestrial Vegetated: Herbaceous Open (40 to 65 %)",
  "Natural Terrestrial Vegetated: Herbaceous Open (15 to 40 %)",
  "Natural Terrestrial Vegetated: Herbaceous Sparse (4 to 15 %)",
  "Natural Terrestrial Vegetated: Herbaceous Scattered (1 to 4 %)"
)

# === 2. Create gradients with PCA ===

# Scale and run PCA for woody
woody_pca <- prcomp(landcover_wide[, woody_cols], center = TRUE, scale. = TRUE)
landcover_wide$Woody_Gradient_PC1 <- woody_pca$x[, 1]

# Scale and run PCA for herbaceous
herb_pca <- prcomp(landcover_wide[, herbaceous_cols], center = TRUE, scale. = TRUE)
landcover_wide$Herbaceous_Gradient_PC1 <- herb_pca$x[, 1]

# === 3. View summary or write out ===
summary(woody_pca)
summary(herb_pca)

# Optional: write to CSV for merging later
write.csv(landcover_wide[, c("id", "Woody_Gradient_PC1", "Herbaceous_Gradient_PC1")],
          "environmental_gradients.csv", row.names = FALSE)
```


```{r weigthed_cover, warning=FALSE, comments="", message=FALSE}

# Add estimated % cover per class
landcover_wide_weighted <- landcover_wide %>%
  mutate(
    Woody_Cover = 
      `Natural Terrestrial Vegetated: Woody Closed (> 65 %)`       * 0.85 +
      `Natural Terrestrial Vegetated: Woody Open (40 to 65 %)`     * 0.525 +
      `Natural Terrestrial Vegetated: Woody Sparse (4 to 15 %)`    * 0.095 +
      `Natural Terrestrial Vegetated: Woody Scattered (1 to 4 %)`  * 0.025,

    Herbaceous_Cover = 
      `Natural Terrestrial Vegetated: Herbaceous Closed (> 65 %)`       * 0.85 +
      `Natural Terrestrial Vegetated: Herbaceous Open (40 to 65 %)`     * 0.525 +
      `Natural Terrestrial Vegetated: Herbaceous Open (15 to 40 %)`     * 0.275 +
      `Natural Terrestrial Vegetated: Herbaceous Sparse (4 to 15 %)`    * 0.095 +
      `Natural Terrestrial Vegetated: Herbaceous Scattered (1 to 4 %)`  * 0.025
  )


```

```{r true_cover_fuse, warning=FALSE, comments="", message=FALSE}
weigthed_continues <-left_join(sample_metadata, landcover_wide_weighted, by = c("sample" = "id"))
```
                                 
```{r true_cover_woddy, warning=FALSE, comments="", message=FALSE}                                 
ggplot(weigthed_continues, aes(x = Woody_Cover, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Estimated Woody Cover (%)", y = "Microbiome Diversity")
```
```{r true%_cover_c_woody, warning=FALSE, comments="", message=FALSE, eval = FALSE}                                 
ggplot(weigthed_continues, aes(x = Woody_Cover, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Estimated Woody Cover (%)", y = "Microbiome Diversity")
```

```{r true%_cover_herbaceous, warning=FALSE, comments="", message=FALSE}                                 
ggplot(weigthed_continues, aes(x = Herbaceous_Cover, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Estimated Herbaceous Cover (%)", y = "Microbiome Diversity")
```

```{r true%_cover_c_herbaceous, warning=FALSE, comments="", message=FALSE}                                 
ggplot(weigthed_continues, aes(x = Herbaceous_Cover, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Estimated Herbaceous Cover (%)", y = "Microbiome Diversity")
```


```{r true%_cover_div, warning=FALSE, comments="", message=FALSE}  

weigthed_continues$Total_Veg_Cover <- weigthed_continues$Woody_Cover + weigthed_continues$Herbaceous_Cover

ggplot(weigthed_continues, aes(x = Total_Veg_Cover, y = diversity)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Total Vegetative Cover (%)", y = "Microbiome Diversity") +
  theme_minimal()
```

```{r true%_cover_c, warning=FALSE, comments="", message=FALSE}  

ggplot(weigthed_continues, aes(x = Total_Veg_Cover, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Total Vegetative Cover (%)", y = "Microbiome Diversity") +
  theme_minimal()
```

```{r true%_cover_regression_div, warning=FALSE, comments="", message=FALSE}  
model <- lm(diversity ~ Woody_Cover + Herbaceous_Cover, data = weigthed_continues)
summary(model)
```

```{r true%_cover_regression_c, warning=FALSE, comments="", message=FALSE}  
model <- lm(C ~ Woody_Cover + Herbaceous_Cover, data = weigthed_continues)
summary(model)
```
```{r genom_count_vegcover, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Prep genome counts
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts),] 
rownames(genome_counts_filt) <- NULL
```

```{r gift2, comment="", echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE}
# Aggregate GIFTs
GIFTs_elements <- to.elements(genome_gifts, GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome, ] %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db)
GIFTs_domains   <- to.domains(GIFTs_functions, GIFT_db)

# Community-weighted GIFTs
genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome), ~ ./sum(.)) %>% 
  column_to_rownames("genome")

GIFTs_elements_community   <- to.community(GIFTs_elements_filtered, genome_counts_row, GIFT_db)
GIFTs_functions_community  <- to.community(GIFTs_functions, genome_counts_row, GIFT_db)
GIFTs_domains_community    <- to.community(GIFTs_domains, genome_counts_row, GIFT_db)
```

## Functional variation by Vegetational Gradient

```{r linear_models_gradient, echo=TRUE, results=TRUE, warning=FALSE, message=FALSE, eval = FALSE}
# Linear models per function using Total_Veg_Cover as predictor
lm_results <- apply(GIFTs_functions_community, 1, function(func) {
  summary(lm(func ~ sample_metadata$Total_Veg_Cover))$coefficients[2, ]
})

# Structure results
lm_df <- as.data.frame(t(lm_results))
colnames(lm_df) <- c("Estimate", "StdError", "t_value", "p_value")

# Sort by significance
lm_df <- lm_df %>% arrange(p_value)

# View top hits
head(lm_df)
```

```{r visualize_significant_functions, echo=TRUE, warning=FALSE, message=FALSE, eval = FALSE}
# Plot top functions
library(ggplot2)
lm_df$Function <- rownames(lm_df)
top_hits <- lm_df %>% filter(p_value < 0.05) %>% head(10)

ggplot(top_hits, aes(x = reorder(Function, Estimate), y = Estimate)) +
  geom_col(fill = "#6a51a3") +
  coord_flip() +
  labs(title = "Top Functional Shifts by Vegetation Gradient",
       x = "Functional Category", y = "Effect Size (Estimate)") +
  theme_minimal()
```

```{r heatmap_significant_functions, echo=TRUE, warning=FALSE, message=FALSE, eval = FALSE}
# Heatmap of significant functions
library(pheatmap)

sig_functions <- rownames(lm_df)[lm_df$p_value < 0.05]
heat_data <- GIFTs_functions_community[sig_functions, ]

# Scale across samples for visualization
heat_data_scaled <- t(scale(t(heat_data)))

pheatmap(heat_data_scaled,
         main = "Heatmap of Significant Functional Shifts",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize_row = 8,
         fontsize_col = 6)
```


```{r ancom_save2, comment="", echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
save(weigthed_continues, Total_Veg_Cover, landcover_wide_weighted
     file="data/vegcover.Rdata")
```