
# functional analysis

Community functional capacity (GIFT framework): overall MCI differences, Wilcoxon/BH tests for element/function/domain contrasts, and compact visuals for the significant functions by sex.

**Note:** Load all required data objects including genome counts, metadata, GIFT database, and color schemes.

```{r load_data_func,comment="", echo=FALSE, message=FALSE, warning=FALSE}
load("data/data.Rdata")
```

**Note:** Filter out sample EHI01340 and filter genome counts and GIFT community matrices to only include valid samples. This ensures consistency across all downstream analyses.

```{r exclude_Sample8, message=FALSE, warning=FALSE}

sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
valid_samples <- sample_metadata$sample

genome_counts_filt <- genome_counts_filt %>%
  dplyr::select(c(genome, tidyselect::any_of(valid_samples)))

# Note: GIFTs_elements_community, GIFTs_functions_community, and GIFTs_domains_community
# are created in the 'gift' chunk from the filtered genome_counts_filt, so they will
# automatically have the correct samples. No need to filter them here.


```

**Note:** Set factor levels for sex variable and define color schemes for plots. Filter genome counts to only include genomes that have GIFT annotations, and reset row names for downstream processing.

```{r colourscheming,comment="", echo=FALSE, message=FALSE, warning=FALSE}
sample_metadata$sex <- factor(sample_metadata$sex, levels=c("Male", "Female"))
treatment_colors <- c("#f56042","#429ef5")
# Define gradient colors for GIFT plots (reused across multiple plots)
gift_gradient_colors <- rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da"))
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts),]

```

**Note:** Define helper function `transpose_and_convert` that performs a common transformation pattern: transposes a dataframe, converts row names to a column, and converts character columns to numeric. This function is reused throughout the analysis for consistent data reshaping.

```{r helper_functions_functional, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Helper function for repeated transformation pattern: transpose, convert row names to column names, convert character to numeric
transpose_and_convert <- function(df, row_name_col = "name") {
  df %>%
    t() %>%
    row_to_names(row_number = 1) %>%
    as.data.frame() %>%
    mutate(across(where(is.character), as.numeric)) %>%
    rownames_to_column(., row_name_col)
}
```

**Note:** Aggregate GIFT functional annotations at multiple hierarchical levels: bundle-level to elements (compounds), elements to functions, and functions to domains (Biosynthesis, Degradation, Structural). Filter elements to remove those with zero abundance across all genomes. Calculate community-weighted average GIFTs per sample by normalizing genome counts to relative abundances and applying weighted aggregation.

```{r gift, comment="", echo=FALSE, message=FALSE, warning=FALSE}
#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select(where(~ !is.numeric(.) | (is.numeric(.) && sum(.) != 0)))

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

#Get community-weighed average GIFTs per sample
genome_counts_row <- genome_counts_filt %>%
  mutate(across(-genome, ~ .x / sum(.x))) %>%
  column_to_rownames("genome")
GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)

save(
  GIFTs_elements_community,
  GIFTs_functions_community,
  GIFTs_domains_community,
  file = "data/gift_community.Rdata"
)
```

## MCI

**Note:** Calculate Mean Community Index (MCI) as the average of all function-level GIFTs per sample. Compute summary statistics (mean ± SD) by sex, test for normality using Shapiro-Wilk test, and perform Wilcoxon rank-sum test to compare MCI between sexes.

```{r gitfs_functional_wild, echo=TRUE,results=TRUE, message=FALSE, warning=FALSE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample)) %>%
  group_by(sex) %>%
  summarise(MCI = mean(value), sd = sd(value), .groups = "drop")

MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample))

shapiro.test(MCI$value)
if (length(unique(MCI$sex)) >= 2L && length(unique(MCI$value)) >= 2L) {
  wilcox.test(value ~ sex, data = MCI)
}

```

**Note:** Create a heatmap visualization of element-level GIFTs across all samples. Map element codes to readable element names and function codes to function names. Facet by function and sex to show patterns in functional capacity at the element level.

```{r comunity_elem_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=15, fig.width=10, fig.fullwidth=TRUE}
GIFTs_elements_community %>%
    as.data.frame() %>%
    rownames_to_column(var="sample")%>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample)) %>%
    mutate(
      functionid = substr(trait, 1, 3),
      trait = case_when(
        trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
        TRUE ~ trait
      ),
      functionid = case_when(
        functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
        TRUE ~ functionid
      ),
      trait = factor(trait, levels=unique(GIFT_db$Element)),
      functionid = factor(functionid, levels=unique(GIFT_db$Function))
    ) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=gift_gradient_colors)+
        facet_grid(functionid ~ sex, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=8),
              strip.text.y = element_text(angle = 0)
              ) +
        labs(y="Traits",x="Samples",fill="GIFT")
```

**Note:** Create a heatmap visualization of function-level GIFTs across all samples. Map function codes to readable function names and facet by sex to show differences in functional capacity at the function level.

```{r comunity_funct_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
unique_funct_db<- GIFT_db[c(3,4,5)] %>% 
  distinct(Code_function, .keep_all = TRUE)

GIFTs_functions_community %>%
   t() %>%
  as.data.frame() %>%
  mutate(across(where(is.character), as.numeric))  %>%
  rownames_to_column(., "Code_function")  %>%
  left_join(.,unique_funct_db[c(1,3)],by = join_by(Code_function))  %>%
  select(-Code_function) %>%
  column_to_rownames(., "Function")%>%
   t()  %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample)) %>%
    ggplot(aes(x=trait,y=sample,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=gift_gradient_colors)+
        facet_grid(sex ~ ., scales="free",space="free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10),
        axis.text.y = element_text(size=8),
        strip.background = element_blank(),
        strip.text = element_text(size = 12, color="black",face="bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)),
        panel.background= element_blank()
              ) +
        labs(x="Function", y="Sample",fill="GIFT")
```

**Note:** Create a heatmap visualization of domain-level GIFTs (Biosynthesis, Degradation, Structural) across all samples. Facet by sex to show differences in functional capacity at the highest hierarchical level.

```{r comunity_dom_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
GIFTs_domains_community %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample)) %>%
    ggplot(aes(x=trait,y=sample,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=gift_gradient_colors)+
        facet_grid(sex ~ ., scales="free",space="free")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size=8),
        strip.background = element_blank(),
        strip.text = element_text(size = 12, color="black",face="bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)),
        panel.background= element_blank()
        ) +
        labs(x="Function", y="Sample",fill="GIFT")
```

## Wilcoxon

### Community elements differences:

**Note:** Prepare element-level GIFT data by converting to dataframe format and joining with sample metadata (sex and broad_environment). This creates the base dataset for element-level differential analysis.

```{r comunity_elem, comment="", echo=FALSE, message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata %>% select(sample,sex, broad_environment), by=join_by(sample))
```

**Note:** Perform Wilcoxon rank-sum tests for each element-level GIFT between sexes. Apply Benjamini-Hochberg (BH) correction for multiple testing and filter to significant elements (p_adjust < 0.05). Create a lookup table mapping elements to functions. Filter element data to only significant elements, calculate mean abundances by sex, and compute difference table showing which sex has higher functional capacity for each element.

```{r commun_wilcox_elem, comment="", echo=FALSE, message=FALSE, warning=FALSE,}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements <- element_gift %>%
    pivot_longer(-c(sample,sex, broad_environment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(
      p_value = {
        vals <- value
        grp <- sex
        if (length(unique(stats::na.omit(grp))) < 2L || length(unique(stats::na.omit(vals))) < 2L) {
          NA_real_
        } else {
          tryCatch(stats::wilcox.test(vals ~ grp)$p.value, error = function(e) NA_real_)
        }
      },
      .groups = "drop"
    ) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05) %>%
    rename(Elements = trait) %>%
    left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(Elements == Code_element))

element_gift_t <- transpose_and_convert(element_gift, "trait")

element_gift_filt <- element_gift_t %>% 
  filter(trait %in% significant_elements$Elements) %>% 
  transpose_and_convert("sample") %>%
  left_join(sample_metadata %>% select(sample, sex, broad_environment), by = join_by(sample)) %>%
  filter(!is.na(sex)) 

element_gift_filt %>%
  select(-c(sample,broad_environment))%>%
  group_by(sex) %>%
  summarise(across(everything(), mean), .groups = "drop")%>%
  transpose_and_convert("Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(Elements == Code_element))

difference_table <- element_gift_filt %>%
  select(-sample) %>%
  group_by(sex) %>%
  summarise(across(everything(), mean), .groups = "drop") %>%
  transpose_and_convert("Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element))

# Arrange by Function if it exists, otherwise by first column
if("Function" %in% names(difference_table)) {
  difference_table <- difference_table %>% arrange(Function)
}

difference_table <- difference_table %>%
  mutate(Difference=Female-Male, group_color = ifelse(Difference <0, "Male","Female")) 

```

**Note:** Calculate log2 fold change for each significant element by comparing mean abundances between Male and Female samples. This quantifies the magnitude and direction of differences in functional capacity.

```{r log_fold_calc, comment="", echo=FALSE, message=FALSE, warning=FALSE}
means_gift <- element_gift_filt %>% 
  select(-broad_environment) %>% 
  pivot_longer(!c(sample, sex), names_to = "elements", values_to = "abundance") %>% 
  group_by(sex, elements) %>%
  summarise(mean=mean(abundance), .groups = "drop")

log_fold <- means_gift %>%
  group_by(elements) %>%
  summarise(
    logfc_high_low = log2((mean[sex == "Male"] + 1e-10) / (mean[sex == "Female"] + 1e-10)),
    .groups = "drop"
    )
```

**Note:** Create boxplots with jittered points showing the distribution of each significant element-level GIFT by sex. Map element codes to readable element names and function names for better visualization. Use facet_wrap to display all significant elements in a grid layout.

```{r elements_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE}
element_gift_names <- element_gift_filt%>%
  select(-c(sex,broad_environment)) %>%
  transpose_and_convert("Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(Elements == Code_element))%>%
  select(Function, everything(), -Elements)%>%
  transpose_and_convert("sample")%>% 
  left_join(sample_metadata %>% select(sample,sex,broad_environment), by = join_by(sample))

# Optimize loop using pivot_longer and facet_wrap
element_gift_names %>%
  pivot_longer(-c(sample, sex, broad_environment), names_to = "element", values_to = "value") %>%
  ggplot(aes(x=sex, y=value, color = sex, fill=sex)) +
    geom_boxplot(alpha = 0.2, outlier.shape = NA, width = 0.3, show.legend = FALSE) +
    geom_jitter(width = 0.1, show.legend = TRUE) +
    scale_color_manual(values=treatment_colors)+
    scale_fill_manual(values=treatment_colors) +
    theme_minimal() +
    theme(
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()) +
    facet_wrap(~element, scales = "free_y")
```

**Note:** Create a bar plot showing the mean difference in element-level GIFTs between sexes, ordered by function. Positive values indicate higher abundance in Females, negative values indicate higher abundance in Males. Color bars by which sex has higher functional capacity.

```{r commun_wilcox_elem_plot_gut, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=12, fig.fullwidth=TRUE, }
difference_table %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
  scale_fill_manual(values=treatment_colors) +
  geom_hline(yintercept=0) +
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Microbial Functional Capacity") + 
  ylab("Mean difference")+
  labs(fill="Sex")
```


**Note:** Create a volcano plot combining statistical significance (-log10 adjusted p-value) and effect size (log2 fold change) for all significant elements. Color points by functional trait category and size by absolute mean difference. Add text labels for element names to identify the most significant and largest effect size differences.

```{r plot3, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=14, fig.fullwidth=TRUE}

uniqueGIFT <- unique(GIFT_db[c(2,3,4,5,6)])

code_function <- difference_table %>%
  left_join(uniqueGIFT[c(1:3)], by=join_by(Elements==Code_element))

unique_codes<-unique(code_function$Code_function)

gift_colors <- read_tsv("data/gift_colors.tsv") %>% 
  filter(Code_function %in% unique_codes)%>% 
  mutate(legend=str_c(Code_function," - ",Function))

code_function %>%
  left_join(significant_elements, by=join_by(Elements)) %>%
  left_join(log_fold, by=join_by(Elements==elements)) %>% 
  left_join(gift_colors, by=join_by(Code_function)) %>% 
  ggplot(., aes(x = logfc_high_low, y = -log(p_adjust), color=legend, size=abs(Difference))) +
  geom_jitter(width = 0.2, height = 0.2)+
  geom_vline(xintercept=0) +
  scale_color_manual(values = gift_colors$Color)+
  theme_classic()+
  labs(size="Mean difference (abs)", color="Functional trait")+
  labs(x = "Log-fold change", y="-Log adjusted p-value") +
  geom_text_repel(aes(label = Element), min.segment.length = 0.4, size=2.5, max.overlaps = Inf)
```

### Community functions differences

**Note:** Prepare function-level GIFT data by converting to dataframe format and joining with sample metadata (sex and broad_environment). This creates the base dataset for function-level differential analysis.

```{r comunity_func_gut, comment="", message=FALSE, warning=FALSE}
function_gift <- GIFTs_functions_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata %>% select(sample,sex,broad_environment), by=join_by(sample))
```

**Note:** Perform Wilcoxon rank-sum tests for each function-level GIFT between sexes. Apply Benjamini-Hochberg (BH) correction for multiple testing and filter to significant functions (p_adjust < 0.05). Join with function names for readability.

```{r commun_wilcox_func_gut, comment="", message=FALSE, warning=FALSE}
unique_funct_db<- GIFT_db[c(3,4,5)] %>% 
  distinct(Code_function, .keep_all = TRUE)

significant_functional <- function_gift %>%
    pivot_longer(-c(sample,sex,broad_environment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(
      p_value = {
        vals <- value
        grp <- sex
        if (length(unique(stats::na.omit(grp))) < 2L || length(unique(stats::na.omit(vals))) < 2L) {
          NA_real_
        } else {
          tryCatch(stats::wilcox.test(vals ~ grp)$p.value, error = function(e) NA_real_)
        }
      },
      .groups = "drop"
    ) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  left_join(.,unique_funct_db[c(1,3)],by = join_by(trait == Code_function))
```

**Note:** Filter function data to only significant functions, transpose for easier manipulation, and calculate mean abundances by sex. Join with function names to create a readable summary table of significant functional differences.

```{r  function_sig_gut, comment="", message=FALSE, warning=FALSE}
function_gift_t <- function_gift  %>% 
  select(-sex)  %>% 
  transpose_and_convert("trait")

function_gift_filt <- function_gift_t %>% 
  filter(trait %in% significant_functional$trait) %>% 
  transpose_and_convert("sample")%>% 
  left_join(., sample_metadata %>% select(sample,sex,broad_environment), by = join_by(sample)) %>%
  filter(!is.na(sex))

function_gift_filt %>%
  select(-c(sample,broad_environment)) %>%
  group_by(sex)  %>%
  summarise(across(everything(), mean), .groups = "drop")%>%
  transpose_and_convert("Code_function") %>%
  left_join(unique_funct_db[c(1,3)],by = join_by(Code_function))
```

**Note:** Create boxplots with jittered points showing the distribution of each significant function-level GIFT by sex. Map function codes to readable function names for better visualization. Use facet_wrap to display all significant functions in a grid layout.

```{r function_plot_gut, comment="", message=FALSE, warning=FALSE}
function_gift_names <- function_gift_filt%>%
  select(-c(sex,broad_environment)) %>%
  transpose_and_convert("Code_function") %>%
  left_join(.,unique_funct_db[c(1,3)],by = join_by(Code_function))%>%
  select(Function, everything(), -Code_function)%>%
  transpose_and_convert("sample")%>% 
  left_join(sample_metadata %>% select(sample,sex,broad_environment), by = join_by(sample))

# Optimize loop using pivot_longer and facet_wrap
function_gift_names %>%
  pivot_longer(-c(sample, sex, broad_environment), names_to = "function_name", values_to = "value") %>%
  ggplot(aes(x=sex, y=value, color = sex, fill=sex)) +
    geom_boxplot(alpha = 0.2, outlier.shape = NA, width = 0.3, show.legend = FALSE) +
    geom_jitter(width = 0.1, show.legend = TRUE) +
    scale_color_manual(values=treatment_colors)+
    scale_fill_manual(values=treatment_colors)+
    theme_minimal() +
    theme(
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()) +
    facet_wrap(~function_name, scales = "free_y")
```

### Community domains differences

**Note:** Prepare domain-level GIFT data by converting to dataframe format and joining with sample metadata (sex). This creates the base dataset for domain-level differential analysis.

```{r comunity_dom_gut, comment="", message=FALSE, warning=FALSE}
domain_gift <- GIFTs_domains_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata %>% select(sample,sex), by=join_by(sample))
```

**Note:** Perform Wilcoxon rank-sum tests for each domain-level GIFT (Biosynthesis, Degradation, Structural) between sexes. Apply Benjamini-Hochberg (BH) correction for multiple testing and filter to significant domains (p_adjust < 0.05).

```{r commun_wilcox_dom_gut, comment="", message=FALSE, warning=FALSE}
unique_domain_db<- GIFT_db[c(4)] %>% 
  distinct(Domain, .keep_all = TRUE)

significant_domain <- domain_gift %>%
    pivot_longer(-c(sample,sex), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(
      p_value = {
        vals <- value
        grp <- sex
        if (length(unique(stats::na.omit(grp))) < 2L || length(unique(stats::na.omit(vals))) < 2L) {
          NA_real_
        } else {
          tryCatch(stats::wilcox.test(vals ~ grp)$p.value, error = function(e) NA_real_)
        }
      },
      .groups = "drop"
    ) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)
```

**Note:** Filter domain data to only significant domains, transpose for easier manipulation, and calculate mean abundances by sex. This creates a summary table showing which domains differ significantly between sexes.

```{r  domain_sig, comment="", echo=FALSE, message=FALSE, warning=FALSE}
domain_gift_t <- domain_gift  %>% 
  select(-sex)  %>% 
  transpose_and_convert("trait")

domain_gift_filt <- domain_gift_t %>% 
  filter(trait %in% significant_domain$trait) %>% 
  transpose_and_convert("sample") %>%
  left_join(sample_metadata[c("sample", "sex")], by = join_by(sample)) %>%
  filter(!is.na(sex))

domain_gift_filt %>%
  select(-sample)%>%
  group_by(sex)  %>%
  summarise(across(everything(), mean), .groups = "drop")%>%
  transpose_and_convert("Code_domain")

```

**Note:** Create boxplots with jittered points showing the distribution of each significant domain-level GIFT by sex. Use facet_wrap to display all significant domains in a grid layout, allowing visual comparison of functional capacity differences at the highest hierarchical level.

```{r domain_plot, comment="", echo=FALSE, message=FALSE, warning=FALSE}
domain_gift_names <- domain_gift_filt%>%
  select(-sex)%>%
  left_join(., sample_metadata[c("sample", "sex")], by = join_by(sample))

# Optimize loop using pivot_longer and facet_wrap
domain_gift_names %>%
  pivot_longer(-c(sample, sex), names_to = "domain", values_to = "value") %>%
  ggplot(aes(x=sex, y=value, color = sex, fill=sex)) +
    geom_boxplot(alpha = 0.2, outlier.shape = NA, width = 0.3, show.legend = FALSE) +
    geom_jitter(width = 0.1, show.legend = TRUE) +
    scale_color_manual(values=treatment_colors)+
    scale_fill_manual(values=treatment_colors)+
    theme_minimal() +
    theme(
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()) +
    facet_wrap(~domain, scales = "free_y")
```