
# environmental permanova

We test whether environmental α-diversity (Hill h0/h1/h2 of land-cover) relates to microbiome β-diversity (Bray–Curtis), and visualise relationships (e.g., NMDS axes vs. Hill metrics) and distance-matrix structure.

```{r load_data_comparison, comment="", echo=FALSE, message=FALSE, warning=FALSE}
load("data/data.Rdata")
load("data/beta.Rdata")
load("data/hill.Rdata")
```

```{r exclude Sample10, message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
hill_long       <- hill_long       %>% dplyr::filter(id     != "EHI01340")
```

```{r Pivot the Hill-long to wide, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Creates columns: env_hill_h0, env_hill_h1, env_hill_h2
env_hill_df <- hill_long %>%
  dplyr::rename(sample = id) %>%
  tidyr::pivot_wider(
    id_cols      = sample,
    names_from   = q,
    values_from  = value,
    names_glue   = "env_hill_{q}"
  )

meta <- sample_metadata %>%
  dplyr::left_join(env_hill_df, by = "sample")

```

```{r PERMANOVAs for q0, q1n, q1f, comment="", echo=FALSE, message=FALSE, warning=FALSE}
run_permanova <- function(beta_obj, meta) {
  dist_obj <- beta_obj$S
  samples  <- attr(dist_obj, "Labels")

  meta_aligned <- meta %>%
    dplyr::filter(sample %in% samples) %>%
    dplyr::arrange(match(sample, samples))
  stopifnot(identical(meta_aligned$sample, samples))

  hill_vars <- c("env_hill_h0", "env_hill_h1", "env_hill_h2")

  for (v in hill_vars) {
    meta_tmp <- meta_aligned[, v, drop = FALSE]
    colnames(meta_tmp) <- "alpha"
    res <- vegan::adonis2(dist_obj ~ alpha, data = meta_tmp, permutations = 999)
    cat("\n--- PERMANOVA for", v, "---\n")
    print(res)
  }
}

run_permanova(beta_q0n, meta)
run_permanova(beta_q1n, meta)
run_permanova(beta_q1f, meta)

```

```{r Prepare distance matrices, comment="", echo=FALSE, message=FALSE, warning=FALSE}
bray_dist0n <- beta_q0n$S
bray_dist1n <- beta_q1n$S
bray_dist1f <- beta_q1f$S
```


```{r Helper to make each plot, comment="", echo=FALSE, message=FALSE, warning=FALSE}
make_nmds_plot <- function(dist_mat, meta_aligned, qvar, title) {
  nmds <- vegan::metaMDS(dist_mat, k = 2, trymax = 500, trace = 0)
  pts  <- as.data.frame(nmds$points)
  pts[[qvar]] <- meta_aligned[[qvar]]

  ggplot2::ggplot(pts, ggplot2::aes(MDS1, MDS2, color = .data[[qvar]])) +
    ggplot2::geom_point(size = 3, alpha = 0.8, shape = 21, stroke = 0.5) +
    ggplot2::theme_minimal(base_size = 14) +
    ggplot2::scale_color_viridis_c(option = "plasma", direction = -1) +
    ggplot2::labs(color = title, x = "NMDS1", y = "NMDS2") +
    ggplot2::theme(
      panel.grid = element_blank(),
      axis.text  = element_text(size = 12),
      axis.title = element_text(face = "bold"),
      legend.title = element_text(face = "bold")
    )
}


```

```{r Align metadata once for q0, comment="", echo=FALSE, message=FALSE, warning=FALSE}
aligned0 <- meta %>%
  dplyr::filter(sample %in% attr(bray_dist0n, "Labels")) %>%
  dplyr::arrange(match(sample, attr(bray_dist0n, "Labels")))

aligned1n <- meta %>%
  dplyr::filter(sample %in% attr(bray_dist1n, "Labels")) %>%
  dplyr::arrange(match(sample, attr(bray_dist1n, "Labels")))

aligned1f <- meta %>%
  dplyr::filter(sample %in% attr(bray_dist1f, "Labels")) %>%
  dplyr::arrange(match(sample, attr(bray_dist1f, "Labels")))

p1 <- make_nmds_plot(bray_dist0n, aligned0, "env_hill_h0", "Environmental Hill q0")
p2 <- make_nmds_plot(bray_dist1n, aligned1n, "env_hill_h1", "Environmental Hill q1")
p3 <- make_nmds_plot(bray_dist1f, aligned1f, "env_hill_h2", "Environmental Hill q2")

cowplot::plot_grid(p1, p2, p3, ncol = 3)

```

```{r Envfit, comment="", echo=FALSE, message=FALSE, warning=FALSE}
fit <- vegan::envfit(nmds0n, aligned0[, c("env_hill_h0", "env_hill_h1", "env_hill_h2")])

# Scores extrahieren
fit_scores <- as.data.frame(scores(fit, display = "vectors"))
fit_scores$variable <- rownames(fit_scores)

ggplot(pts0n, aes(MDS1, MDS2)) +
  geom_point(color = "grey50", size = 2) +
  geom_segment(data = fit_scores,
               aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text(data = fit_scores,
            aes(x = NMDS1, y = NMDS2, label = variable),
            vjust = -0.8, color = "red", fontface = "bold") +
  theme_minimal(base_size = 14)


```

```{r Betadisper by tertiles, comment="", echo=FALSE, message=FALSE, warning=FALSE}
bd <- vegan::betadisper(
  bray_dist0n,
  ggplot2::cut_number(aligned0$env_hill_h0, 3)
)
plot(bd)
anova(bd)

```

```{r  Splitting into quantiles or medians, comment="", echo=FALSE, message=FALSE, warning=FALSE}
meta <- meta %>%
  dplyr::mutate(
    env_hill_q0_group     = ggplot2::cut_number(env_hill_h0, 3, labels = c("Low","Mid","High")),
    env_hill_q1_group     = ggplot2::cut_number(env_hill_h1, 3, labels = c("Low","Mid","High")),
    env_hill_q2_group     = ggplot2::cut_number(env_hill_h2, 3, labels = c("Low","Mid","High")),
    env_hill_q0_group_med = ifelse(env_hill_h0 <= median(env_hill_h0, na.rm = TRUE), "Low", "High"),
    env_hill_q1_group_med = ifelse(env_hill_h1 <= median(env_hill_h1, na.rm = TRUE), "Low", "High"),
    env_hill_q2_group_med = ifelse(env_hill_h2 <= median(env_hill_h2, na.rm = TRUE), "Low", "High")
  )

table(meta$env_hill_q0_group)
table(meta$env_hill_q0_group_med)

```

```{r checking validty, comment="", echo=FALSE, message=FALSE, warning=FALSE}
table(meta$env_hill_q0_group)
table(meta$env_hill_q0_group_med)
```
**visualisation**

Associations between environmental α-diversity (Hill h1) and microbiome ordination (NMDS1/2), with fitted vectors and a subset heatmap of Bray–Curtis distances.

```{r tidy NMDS‐points data frame, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nmds0n <- vegan::metaMDS(bray_dist0n, k = 2, trymax = 500, trace = 0)
pts0n <- as.data.frame(nmds0n$points) %>%
  dplyr::mutate(sample = rownames(.)) %>%
  dplyr::left_join(meta %>% dplyr::select(sample, env_hill_q0_group), by = "sample")

p_box1 <- ggplot2::ggplot(pts0n, ggplot2::aes(x = env_hill_q0_group, y = MDS1, fill = env_hill_q0_group)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_fill_viridis_d(name = "q0 group") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "Hill q0 tertile", y = "NMDS1", title = "NMDS1 by q0 group")

p_box2 <- ggplot2::ggplot(pts0n, ggplot2::aes(x = env_hill_q0_group, y = MDS2, fill = env_hill_q0_group)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_fill_viridis_d(name = "q0 group") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "Hill q0 tertile", y = "NMDS2", title = "NMDS2 by q0 group")

cowplot::plot_grid(p_box1, p_box2, ncol = 2)

```

```{r Scatterplots: Hill α‐diversity vs. NMDS axes}
# Scatterplots: Hill α‐diversity (q1) vs. NMDS axes
nmds1n <- vegan::metaMDS(bray_dist1n, k = 2, trymax = 500, trace = 0)

pts1n <- as.data.frame(nmds1n$points) %>%
  dplyr::mutate(sample = rownames(.)) %>%
  dplyr::left_join(aligned1n %>% dplyr::select(sample, env_hill_h1), by = "sample")

ggplot2::ggplot(pts1n, ggplot2::aes(env_hill_h1, MDS1)) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(method = "lm", se = FALSE) +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "Hill q1 (environmental α)", y = "NMDS1", title = "NMDS1 vs Hill q1")

ggplot2::ggplot(pts1n, ggplot2::aes(env_hill_h1, MDS2)) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(method = "lm", se = FALSE) +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "Hill q1 (environmental α)", y = "NMDS2", title = "NMDS2 vs Hill q1")

```

```{r Envfit arrows over ggplot NMDS s}
# Fit environmental Hill vectors (h0/h1/h2) onto the q0 NMDS
nmds0n <- vegan::metaMDS(bray_dist0n, k = 2, trymax = 500, trace = 0)

fit <- vegan::envfit(
  nmds0n,
  aligned0[, c("env_hill_h0", "env_hill_h1", "env_hill_h2")],
  permutations = 999
)

# Base plotting (robust and simple)
plot(nmds0n, type = "n")
points(nmds0n, pch = 19, col = "grey30")
plot(fit, p.max = 0.05, col = "black")

```

```{r Heatmap of pairwise dissimilarities}
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
library(pheatmap)

set.seed(42)
labels0 <- attr(bray_dist0n, "Labels")
sub_samps <- sample(labels0, min(30, length(labels0)))

mat_sub <- as.matrix(bray_dist0n)[sub_samps, sub_samps]
dm_sub  <- as.dist(mat_sub)

pheatmap::pheatmap(
  mat_sub,
  clustering_distance_rows = dm_sub,
  clustering_distance_cols = dm_sub,
  main = "Bray–Curtis dissimilarity (q0) — 30 samples"
)

```
