
# environmental permanova

We test whether environmental α-diversity (Hill h0/h1/h2 of land-cover) relates to microbiome β-diversity (Bray–Curtis), and visualise relationships (e.g., NMDS axes vs. Hill metrics) and distance-matrix structure.

## Load data

**Note:** Load all required data objects including sample metadata, beta diversity matrices, and Hill diversity metrics from the RData files created in previous chapters.

```{r load_data_comparison, comment="", echo=FALSE, message=FALSE, warning=FALSE}
load("data/data.Rdata")
load("data/beta.Rdata")
load("data/hill.Rdata")
```

## Data preparation

**Note:** Filter out sample EHI01340 from the metadata and Hill diversity data. This sample was removed because it could not be mapped to MAGs, ensuring consistency across downstream analyses.

```{r exclude_Sample10_permanova, message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% dplyr::filter(sample != "EHI01340")
hill_long       <- hill_long       %>% dplyr::filter(id     != "EHI01340")
```

**Note:** Transform Hill diversity data from long to wide format, creating separate columns for each Hill diversity metric (env_hill_h0, env_hill_h1, env_hill_h2). Join with sample metadata to create a combined metadata object for downstream analyses.

```{r Pivot the Hill-long to wide, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Creates columns: env_hill_h0, env_hill_h1, env_hill_h2
env_hill_df <- hill_long %>%
  dplyr::rename(sample = id) %>%
  tidyr::pivot_wider(
    id_cols      = sample,
    names_from   = q,
    values_from  = value,
    names_glue   = "env_hill_{q}"
  )

meta <- sample_metadata %>%
  dplyr::left_join(env_hill_df, by = "sample")
```

## Helper functions

**Note:** Helper function to align metadata with distance matrices. This eliminates code duplication when aligning metadata for multiple distance matrices. Defined early as it's needed by multiple analysis sections.

```{r Helper function to align metadata, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# OPTIMIZED: Helper function to align metadata with distance matrix
align_metadata <- function(meta, dist_mat) {
  meta %>%
    dplyr::filter(sample %in% attr(dist_mat, "Labels")) %>%
    dplyr::arrange(match(sample, attr(dist_mat, "Labels")))
}
```

**Note:** Optimized helper function that accepts pre-calculated NMDS object instead of recalculating. This eliminates redundant NMDS calculations when creating multiple plots.

```{r Helper to make each plot OPTIMIZED, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# OPTIMIZED: Accept pre-calculated NMDS instead of recalculating
make_nmds_plot <- function(nmds_obj, meta_aligned, qvar, title) {
  pts  <- as.data.frame(nmds_obj$points)
  pts[[qvar]] <- meta_aligned[[qvar]]

  ggplot2::ggplot(pts, ggplot2::aes(MDS1, MDS2, color = .data[[qvar]])) +
    ggplot2::geom_point(size = 3, alpha = 0.8, shape = 21, stroke = 0.5) +
    ggplot2::theme_minimal(base_size = 14) +
    ggplot2::scale_color_viridis_c(option = "plasma", direction = -1) +
    ggplot2::labs(color = title, x = "NMDS1", y = "NMDS2") +
    ggplot2::theme(
      panel.grid = element_blank(),
      axis.text  = element_text(size = 12),
      axis.title = element_text(face = "bold"),
      legend.title = element_text(face = "bold")
    )
}
```

## Ordination analysis

**Note:** Extract distance matrices from beta diversity objects for use in ordination analyses. These matrices represent Bray-Curtis dissimilarities between samples at different diversity orders.

```{r Prepare distance matrices, comment="", echo=FALSE, message=FALSE, warning=FALSE}
bray_dist0n <- beta_q0n$S
bray_dist1n <- beta_q1n$S
bray_dist1f <- beta_q1f$S
```

**Note:** Calculate all NMDS ordinations once upfront. These are expensive operations (trymax = 500 iterations each) and were previously calculated redundantly 3+ times. This optimization saves significant computation time.

```{r Calculate all NMDS ordinations once, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Calculate all NMDS ordinations upfront - expensive operation, do once
nmds0n <- vegan::metaMDS(bray_dist0n, k = 2, trymax = 500, trace = 0)
nmds1n <- vegan::metaMDS(bray_dist1n, k = 2, trymax = 500, trace = 0)
nmds1f <- vegan::metaMDS(bray_dist1f, k = 2, trymax = 500, trace = 0)
```

**Note:** Align metadata with each distance matrix using the helper function. This ensures proper sample order matching for ordination and visualization.

```{r Align metadata for all matrices, comment="", echo=FALSE, message=FALSE, warning=FALSE}
aligned0 <- align_metadata(meta, bray_dist0n)
aligned1n <- align_metadata(meta, bray_dist1n)
aligned1f <- align_metadata(meta, bray_dist1f)

# OPTIMIZED: Use pre-calculated NMDS objects instead of recalculating
p1 <- make_nmds_plot(nmds0n, aligned0, "env_hill_h0", "Environmental Hill q0")
p2 <- make_nmds_plot(nmds1n, aligned1n, "env_hill_h1", "Environmental Hill q1")
p3 <- make_nmds_plot(nmds1f, aligned1f, "env_hill_h2", "Environmental Hill q2")

cowplot::plot_grid(p1, p2, p3, ncol = 3)
```

## Grouping variables

**Note:** Create grouping variables for environmental Hill diversity metrics. Calculate medians once to avoid redundant calculations. Split into tertiles (Low, Mid, High) and median-based groups (Low, High) for visualization and statistical testing.

```{r Splitting into quantiles or medians OPTIMIZED, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# OPTIMIZED: Calculate medians once instead of three times
med_h0 <- median(meta$env_hill_h0, na.rm = TRUE)
med_h1 <- median(meta$env_hill_h1, na.rm = TRUE)
med_h2 <- median(meta$env_hill_h2, na.rm = TRUE)

meta <- meta %>%
  dplyr::mutate(
    env_hill_q0_group     = ggplot2::cut_number(env_hill_h0, 3, labels = c("Low","Mid","High")),
    env_hill_q1_group     = ggplot2::cut_number(env_hill_h1, 3, labels = c("Low","Mid","High")),
    env_hill_q2_group     = ggplot2::cut_number(env_hill_h2, 3, labels = c("Low","Mid","High")),
    env_hill_q0_group_med = ifelse(env_hill_h0 <= med_h0, "Low", "High"),
    env_hill_q1_group_med = ifelse(env_hill_h1 <= med_h1, "Low", "High"),
    env_hill_q2_group_med = ifelse(env_hill_h2 <= med_h2, "Low", "High")
  )
```

**Note:** Verify the grouping variables were created correctly by displaying frequency tables. This ensures balanced group sizes for downstream analyses.

```{r checking validity, comment="", echo=FALSE, message=FALSE, warning=FALSE}
table(meta$env_hill_q0_group)
table(meta$env_hill_q0_group_med)
```

## PERMANOVA analysis

**Note:** Define function to run PERMANOVA tests for each environmental Hill diversity metric (h0, h1, h2) against beta diversity distance matrices. The function aligns metadata with distance matrix samples, ensures sample order matches, and performs PERMANOVA with 999 permutations for each Hill metric. This eliminates code duplication when testing multiple beta diversity matrices.

```{r PERMANOVAs for q0, q1n, q1f, comment="", echo=FALSE, message=FALSE, warning=FALSE}
run_permanova <- function(beta_obj, meta) {
  dist_obj <- beta_obj$S
  samples  <- attr(dist_obj, "Labels")

  meta_aligned <- meta %>%
    dplyr::filter(sample %in% samples) %>%
    dplyr::arrange(match(sample, samples))
  stopifnot(identical(meta_aligned$sample, samples))

  hill_vars <- c("env_hill_h0", "env_hill_h1", "env_hill_h2")

  for (v in hill_vars) {
    meta_tmp <- meta_aligned[, v, drop = FALSE]
    colnames(meta_tmp) <- "alpha"
    res <- vegan::adonis2(dist_obj ~ alpha, data = meta_tmp, permutations = 999)
    cat("\n--- PERMANOVA for", v, "---\n")
    print(res)
  }
}

run_permanova(beta_q0n, meta)
run_permanova(beta_q1n, meta)
run_permanova(beta_q1f, meta)
```

## Mantel tests

**Note:** Test correlations between environmental Hill diversity and beta diversity using Mantel tests. This is more appropriate than linear regression on NMDS axes since it tests relationships directly on distance matrices, which is the appropriate scale for beta diversity analysis.

```{r Mantel tests for Hill diversity, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Create distance matrices for environmental variables (Euclidean distance)
# Need to align with distance matrix samples first
meta_for_mantel <- align_metadata(meta, bray_dist0n)

env_dist_h0 <- dist(meta_for_mantel$env_hill_h0)
env_dist_h1 <- dist(meta_for_mantel$env_hill_h1)
env_dist_h2 <- dist(meta_for_mantel$env_hill_h2)

# Mantel tests for q0n
mantel_h0_q0n <- vegan::mantel(bray_dist0n, env_dist_h0, permutations = 999)
mantel_h1_q0n <- vegan::mantel(bray_dist0n, env_dist_h1, permutations = 999)
mantel_h2_q0n <- vegan::mantel(bray_dist0n, env_dist_h2, permutations = 999)

cat("\n=== Mantel tests for q0n beta diversity ===\n")
cat("\nHill h0:\n")
print(mantel_h0_q0n)
cat("\nHill h1:\n")
print(mantel_h1_q0n)
cat("\nHill h2:\n")
print(mantel_h2_q0n)

# Mantel tests for q1n
meta_for_mantel_q1n <- align_metadata(meta, bray_dist1n)
env_dist_h0_q1n <- dist(meta_for_mantel_q1n$env_hill_h0)
env_dist_h1_q1n <- dist(meta_for_mantel_q1n$env_hill_h1)
env_dist_h2_q1n <- dist(meta_for_mantel_q1n$env_hill_h2)

mantel_h0_q1n <- vegan::mantel(bray_dist1n, env_dist_h0_q1n, permutations = 999)
mantel_h1_q1n <- vegan::mantel(bray_dist1n, env_dist_h1_q1n, permutations = 999)
mantel_h2_q1n <- vegan::mantel(bray_dist1n, env_dist_h2_q1n, permutations = 999)

cat("\n=== Mantel tests for q1n beta diversity ===\n")
cat("\nHill h0:\n")
print(mantel_h0_q1n)
cat("\nHill h1:\n")
print(mantel_h1_q1n)
cat("\nHill h2:\n")
print(mantel_h2_q1n)

# Mantel tests for q1f
meta_for_mantel_q1f <- align_metadata(meta, bray_dist1f)
env_dist_h0_q1f <- dist(meta_for_mantel_q1f$env_hill_h0)
env_dist_h1_q1f <- dist(meta_for_mantel_q1f$env_hill_h1)
env_dist_h2_q1f <- dist(meta_for_mantel_q1f$env_hill_h2)

mantel_h0_q1f <- vegan::mantel(bray_dist1f, env_dist_h0_q1f, permutations = 999)
mantel_h1_q1f <- vegan::mantel(bray_dist1f, env_dist_h1_q1f, permutations = 999)
mantel_h2_q1f <- vegan::mantel(bray_dist1f, env_dist_h2_q1f, permutations = 999)

cat("\n=== Mantel tests for q1f beta diversity ===\n")
cat("\nHill h0:\n")
print(mantel_h0_q1f)
cat("\nHill h1:\n")
print(mantel_h1_q1f)
cat("\nHill h2:\n")
print(mantel_h2_q1f)
```

## Betadisper tests

**Note:** Test for homogeneity of multivariate dispersions (variance) across Hill diversity tertile groups using betadisper. This tests whether groups differ in their dispersion around the centroid, which is an assumption of PERMANOVA. Tests all three beta diversity matrices for consistency.

```{r Betadisper by tertiles, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Betadisper for q0n
bd_q0n <- vegan::betadisper(
  bray_dist0n,
  ggplot2::cut_number(aligned0$env_hill_h0, 3)
)
cat("\n=== Betadisper for q0n beta diversity ===\n")
plot(bd_q0n)
anova(bd_q0n)

# Betadisper for q1n
bd_q1n <- vegan::betadisper(
  bray_dist1n,
  ggplot2::cut_number(aligned1n$env_hill_h1, 3)
)
cat("\n=== Betadisper for q1n beta diversity ===\n")
plot(bd_q1n)
anova(bd_q1n)

# Betadisper for q1f
bd_q1f <- vegan::betadisper(
  bray_dist1f,
  ggplot2::cut_number(aligned1f$env_hill_h2, 3)
)
cat("\n=== Betadisper for q1f beta diversity ===\n")
plot(bd_q1f)
anova(bd_q1f)
```

## Correlation tests

**Note:** Test correlations between environmental Hill diversity variables and NMDS ordination axes using Spearman correlation. This is more appropriate than linear regression for non-metric ordination and provides correlation coefficients and significance tests.

```{r Correlation tests: Hill diversity vs NMDS axes, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Correlation tests for all Hill metrics vs NMDS axes
cor_results <- list()

# For q0 NMDS
for (hill_var in c("env_hill_h0", "env_hill_h1", "env_hill_h2")) {
  cor_mds1 <- cor.test(aligned0[[hill_var]], nmds0n$points[,1], method = "spearman")
  cor_mds2 <- cor.test(aligned0[[hill_var]], nmds0n$points[,2], method = "spearman")
  cor_results[[paste0(hill_var, "_MDS1")]] <- cor_mds1
  cor_results[[paste0(hill_var, "_MDS2")]] <- cor_mds2
}

# For q1n NMDS
for (hill_var in c("env_hill_h0", "env_hill_h1", "env_hill_h2")) {
  cor_mds1 <- cor.test(aligned1n[[hill_var]], nmds1n$points[,1], method = "spearman")
  cor_mds2 <- cor.test(aligned1n[[hill_var]], nmds1n$points[,2], method = "spearman")
  cor_results[[paste0(hill_var, "_q1n_MDS1")]] <- cor_mds1
  cor_results[[paste0(hill_var, "_q1n_MDS2")]] <- cor_mds2
}

# For q1f NMDS
for (hill_var in c("env_hill_h0", "env_hill_h1", "env_hill_h2")) {
  cor_mds1 <- cor.test(aligned1f[[hill_var]], nmds1f$points[,1], method = "spearman")
  cor_mds2 <- cor.test(aligned1f[[hill_var]], nmds1f$points[,2], method = "spearman")
  cor_results[[paste0(hill_var, "_q1f_MDS1")]] <- cor_mds1
  cor_results[[paste0(hill_var, "_q1f_MDS2")]] <- cor_mds2
}

# Display results
cat("\n=== Spearman correlations: Hill diversity vs NMDS axes ===\n")
cat("\n--- q0n NMDS ---\n")
for (name in grep("_MDS", names(cor_results), value = TRUE)) {
  if (!grepl("_q1", name)) {
    cat("\n", name, ":\n")
    print(cor_results[[name]])
  }
}
cat("\n--- q1n NMDS ---\n")
for (name in grep("_q1n_MDS", names(cor_results), value = TRUE)) {
  cat("\n", name, ":\n")
  print(cor_results[[name]])
}
cat("\n--- q1f NMDS ---\n")
for (name in grep("_q1f_MDS", names(cor_results), value = TRUE)) {
  cat("\n", name, ":\n")
  print(cor_results[[name]])
}
```

## Visualization

Associations between environmental α-diversity (Hill h0/h1/h2) and microbiome ordination (NMDS1/2), with fitted vectors and a subset heatmap of Bray–Curtis distances.

**Note:** Create combined scatterplots showing relationships between all environmental Hill diversity metrics and NMDS ordination axes. This provides a comprehensive view of how environmental diversity relates to microbiome β-diversity ordination. Uses loess smoothing instead of linear regression, which is more appropriate for potentially non-linear relationships.

```{r Combined scatterplot: All Hill metrics vs NMDS, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Create combined data frame for all Hill metrics
pts_all <- as.data.frame(nmds0n$points) %>%
  dplyr::mutate(sample = rownames(.)) %>%
  dplyr::left_join(aligned0 %>% dplyr::select(sample, env_hill_h0, env_hill_h1, env_hill_h2), by = "sample") %>%
  tidyr::pivot_longer(cols = c(env_hill_h0, env_hill_h1, env_hill_h2), 
                      names_to = "hill_metric", 
                      values_to = "hill_value",
                      names_prefix = "env_hill_") %>%
  tidyr::pivot_longer(cols = c(MDS1, MDS2), 
                      names_to = "nmds_axis", 
                      values_to = "nmds_value")

# Create faceted scatterplot
p_combined <- ggplot2::ggplot(pts_all, ggplot2::aes(x = hill_value, y = nmds_value)) +
  ggplot2::geom_point(alpha = 0.6, size = 1.5) +
  ggplot2::geom_smooth(method = "loess", se = TRUE, color = "red", linewidth = 1) +
  ggplot2::facet_grid(nmds_axis ~ hill_metric, scales = "free_x") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(x = "Environmental Hill diversity", y = "NMDS axis value") +
  ggplot2::theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p_combined
```

**Note:** Fit environmental Hill diversity vectors (h0, h1, h2) onto the q0 NMDS ordination using envfit. This tests the correlation between environmental variables and ordination axes, displaying significant vectors as arrows on the ordination plot. Improved visualization using ggplot2 for consistency with other plots.

```{r Envfit ggplot visualization, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# OPTIMIZED: Use pre-calculated nmds0n from above
# Fit environmental Hill vectors (h0/h1/h2) onto the q0 NMDS
fit <- vegan::envfit(
  nmds0n,
  aligned0[, c("env_hill_h0", "env_hill_h1", "env_hill_h2")],
  permutations = 999
)

# Extract scores for significant vectors only
fit_scores <- as.data.frame(vegan::scores(fit, display = "vectors")) %>%
  dplyr::mutate(variable = rownames(.)) %>%
  dplyr::mutate(p_value = fit$vectors$pvals) %>%
  dplyr::filter(p_value < 0.05)  # Only significant vectors

# Create base plot data
pts0n_plot <- as.data.frame(nmds0n$points) %>%
  dplyr::mutate(sample = rownames(.))

# Create ggplot visualization
p_envfit <- ggplot2::ggplot(pts0n_plot, ggplot2::aes(x = MDS1, y = MDS2)) +
  ggplot2::geom_point(color = "grey50", size = 2.5, alpha = 0.7) +
  ggplot2::geom_segment(data = fit_scores,
                       ggplot2::aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
                       arrow = ggplot2::arrow(length = ggplot2::unit(0.3, "cm"), type = "closed"),
                       color = "red", linewidth = 1.2) +
  ggplot2::geom_text(data = fit_scores,
                    ggplot2::aes(x = NMDS1 * 1.15, y = NMDS2 * 1.15, label = variable),
                    hjust = 0, vjust = 0.5, color = "red", fontface = "bold", size = 4.5) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::coord_fixed() +
  ggplot2::labs(
    title = "NMDS ordination with environmental vectors (p < 0.05)",
    x = "NMDS1",
    y = "NMDS2"
  ) +
  ggplot2::theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_text(size = 12),
    axis.title = element_text(face = "bold")
  )

p_envfit
```

**Note:** Create an improved heatmap visualization of Bray-Curtis dissimilarity matrix with sample grouping information. This shows clustering patterns in microbiome β-diversity and how they relate to environmental Hill diversity groups. Uses all samples (or a representative subset if too large) with grouping annotations.

```{r Improved heatmap with grouping, comment="", echo=FALSE, message=FALSE, warning=FALSE}

# Get sample grouping information from meta (which has the grouping variables)
# Join with the distance matrix labels to ensure proper alignment
sample_groups <- meta %>%
  dplyr::filter(sample %in% attr(bray_dist0n, "Labels")) %>%
  dplyr::select(sample, env_hill_q0_group) %>%
  dplyr::arrange(match(sample, attr(bray_dist0n, "Labels")))

# Create annotation for heatmap
annotation_row <- data.frame(
  Hill_q0_group = sample_groups$env_hill_q0_group
)
rownames(annotation_row) <- sample_groups$sample

# Color palette for groups (viridis-like colors)
ann_colors <- list(
  Hill_q0_group = c("Low" = "#440154FF", "Mid" = "#21908CFF", "High" = "#FDE725FF")
)

# If too many samples, use a subset for visualization
n_samples <- length(attr(bray_dist0n, "Labels"))
if (n_samples > 50) {
  set.seed(42)
  sub_samps <- sample(attr(bray_dist0n, "Labels"), 50)
  mat_plot <- as.matrix(bray_dist0n)[sub_samps, sub_samps]
  annotation_plot <- annotation_row[sub_samps, , drop = FALSE]
  title_text <- "Bray–Curtis dissimilarity (q0) — 50 samples (subset)"
} else {
  mat_plot <- as.matrix(bray_dist0n)
  annotation_plot <- annotation_row
  title_text <- paste("Bray–Curtis dissimilarity (q0) —", n_samples, "samples")
}

# Create heatmap with grouping
pheatmap::pheatmap(
  mat_plot,
  clustering_distance_rows = as.dist(mat_plot),
  clustering_distance_cols = as.dist(mat_plot),
  annotation_row = annotation_plot,
  annotation_colors = ann_colors,
  main = title_text,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("#440154", "#21908C", "#FDE725"))(100)
)

```
