# HMSC final model setup

This chapter defines the definitive HMSC model for downstream analyses. The model is restricted to Tasmania samples, uses 200 m environmental buffers, and includes a devil × temperature interaction term.

```{r load_libraries_final, comment="", message=FALSE, warning=FALSE}
library(tidyverse)
library(Hmsc)
library(ape)
```

## Load data

```{r load_data_final, comment="", message=FALSE, warning=FALSE}
source("R/chapter_deps.R")
load("data/data.Rdata")
ensure_hmsc_env_csvs()

animal_variables <- read_csv("data/animal_var.csv")
location_variables <- read_csv("data/environment_var_r200.csv")

predictors <- inner_join(animal_variables, location_variables, by = "id") %>%
  rename(sample = id) %>%
  filter(sample %in% sample_metadata$sample[sample_metadata$island == "Tasmania"])
```

## Filter genomes

The genome set is pinned in `data/model_final_genomes.tsv` (576 genomes) so it matches the fitted model on ERDA. Do not re-derive this list from prevalence filters unless you intend to refit the model on HPC.

```{r filter_genomes_final, comment="", message=FALSE, warning=FALSE}
genomes_to_model <- read_tsv("data/model_final_genomes.tsv", show_col_types = FALSE) %>%
  pull(genome)
```

## Build model inputs

```{r hmsc_input_final, warning=FALSE, comments="", message=FALSE}
StudyDesign <- predictors %>%
  select(sample, site) %>%
  mutate(animal = sample) %>%
  mutate(animal = factor(animal), site = factor(site)) %>%
  column_to_rownames("sample")

normalisation_factors <- genome_metadata %>%
  mutate(norm_factor = median(length) / length) %>%
  select(genome, norm_factor)

YData <- read_counts %>%
  left_join(normalisation_factors, by = "genome") %>%
  mutate(across(-c(genome, norm_factor), ~ round(. * norm_factor, 0))) %>%
  select(-norm_factor) %>%
  mutate(across(where(is.numeric), ~ . + 1)) %>%
  mutate(across(where(is.numeric), ~ log(.))) %>%
  filter(genome %in% genomes_to_model) %>%
  column_to_rownames("genome") %>%
  select(all_of(row.names(StudyDesign))) %>%
  as.data.frame() %>%
  t()

XData <- predictors %>%
  select(sample, temperature, precipitation, diversity, density, devil) %>%
  mutate(logseqdepth = read_counts %>%
    select(all_of(row.names(StudyDesign))) %>%
    colSums() %>%
    log()) %>%
  column_to_rownames("sample")

PData <- genome_tree %>%
  keep.tip(tip = genomes_to_model)
```

## Model formula

```{r hmsc_formulas_final, warning=FALSE, comments="", message=FALSE}
XFormula = ~ devil + temperature + diversity + logseqdepth + devil:temperature

rL.animal = HmscRandomLevel(units = levels(StudyDesign$animal))
rL.site = HmscRandomLevel(units = levels(StudyDesign$site))
```

## Define and save unfitted model

The unfitted model object is saved as `hmsc/model_final`. The version in the repository (55 Tasmania samples, 576 genomes) is the canonical model for fitting and analysis. Re-running the chunks above may produce a different genome set because filtering was tuned for the final model; only rebuild if you intend to replace `hmsc/model_final` deliberately.

```{r hmsc_models_final, warning=FALSE, comments="", message=FALSE}
m = Hmsc(
  Y = YData,
  XData = XData,
  XFormula = XFormula,
  studyDesign = StudyDesign,
  phyloTree = PData,
  ranLevels = list("animal" = rL.animal, "site" = rL.site),
  distr = "normal",
  YScale = TRUE
)

save(m, file = "hmsc/model_final")
```

## In the HPC

The commands below are for the HPC cluster only. They are shown for reference and are **not executed** when knitting this book locally. The launch script is also saved as `launch_hmsc_model_final.sh` in the project root.

### Prepare the initiation object

```bash
conda activate hmsc-hpc

model="hmsc/model_final"
init="init/model_final.rds"

export samples=250
export thin=1000
export transient=$(( samples * thin ))
export nChains=4
export verbose=TRUE
export model
export init

# Get initiation object (hmsc.r must be available on the HPC)
Rscript hmsc.r $model $init
```

### Launch script

```bash
cat <<'EOF' > launch_hmsc_model_final.sh
#!/bin/bash
#SBATCH --job-name=hmsc_model_final
#SBATCH --nodes=1
#SBATCH --partition=cpuqueue
#SBATCH --qos=normal
#SBATCH --gres=gpu:1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=1
#SBATCH --mem=32gb
#SBATCH --time=150:00:00

# Activate conda environment
source activate hmsc-hpc

# Set output files
output="output/Hmsc_model_final.rds"
mkdir output

# Run model fit
srun python3 -m hmsc.run_gibbs_sampler --input $init --output $output --samples $samples --transient $transient --thin $thin --verbose 100
EOF
```

### Submit job

```bash
sbatch launch_hmsc_model_final.sh
```
