# HMSC final model setup

This chapter defines the definitive HMSC model for downstream analyses. The model is restricted to Tasmania samples, uses 200 m environmental buffers, and includes a devil × temperature interaction term.

```{r load_libraries_final, comment="", message=FALSE, warning=FALSE, eval=FALSE}
library(tidyverse)
library(Hmsc)
library(ape)
```

## Load data

```{r load_data_final, comment="", message=FALSE, warning=FALSE, eval=FALSE}
load("data/data.Rdata")

animal_variables <- read_csv("data/animal_var.csv")
location_variables <- read_csv("data/environment_var_r200.csv")

predictors <- inner_join(animal_variables, location_variables, by = "id") %>%
  rename(sample = id) %>%
  filter(sample %in% sample_metadata$sample[sample_metadata$island == "Tasmania"])
```

## Filter genomes

```{r filter_genomes_final, comment="", message=FALSE, warning=FALSE, eval=FALSE}
prevalent_genomes <- genome_counts_filt %>%
  select(genome, all_of(predictors$sample)) %>%
  rowwise() %>%
  mutate(nonzero_prop = mean(c_across(-genome) > 0)) %>%
  ungroup() %>%
  filter(nonzero_prop > 0.10) %>%
  select(-nonzero_prop) %>%
  pull(genome)

widespread_genomes <- genome_counts_filt %>%
  pivot_longer(!genome, names_to = "sample", values_to = "count") %>%
  left_join(location_variables, by = join_by("sample" == "id")) %>%
  filter(sample %in% predictors$sample) %>%
  group_by(genome, site) %>%
  summarise(nonzero_samples = sum(count > 0), .groups = "drop") %>%
  group_by(genome) %>%
  summarise(nonzero_sites = sum(nonzero_samples > 0), .groups = "drop") %>%
  filter(nonzero_sites > 5) %>%
  pull(genome)

genomes_to_model <- intersect(prevalent_genomes, widespread_genomes)
```

## Build model inputs

```{r hmsc_input_final, warning=FALSE, comments="", message=FALSE, eval=FALSE}
StudyDesign <- predictors %>%
  select(sample, site) %>%
  mutate(animal = sample) %>%
  mutate(animal = factor(animal), site = factor(site)) %>%
  column_to_rownames("sample")

normalisation_factor <- genome_metadata %>%
  mutate(factor = median(length) / length) %>%
  pull(factor)

YData <- read_counts %>%
  mutate(across(where(is.numeric), ~ round(. * normalisation_factor, 0))) %>%
  mutate(across(where(is.numeric), ~ . + 1)) %>%
  mutate(across(where(is.numeric), ~ log(.))) %>%
  filter(genome %in% genomes_to_model) %>%
  column_to_rownames("genome") %>%
  select(all_of(row.names(StudyDesign))) %>%
  as.data.frame() %>%
  t()

XData <- predictors %>%
  select(sample, temperature, precipitation, diversity, density, devil) %>%
  mutate(logseqdepth = read_counts %>%
    select(all_of(row.names(StudyDesign))) %>%
    colSums() %>%
    log()) %>%
  column_to_rownames("sample")

PData <- genome_tree %>%
  keep.tip(tip = genomes_to_model)
```

## Model formula

```{r hmsc_formulas_final, warning=FALSE, comments="", message=FALSE, eval=FALSE}
XFormula = ~ devil + temperature + diversity + logseqdepth + devil:temperature

rL.animal = HmscRandomLevel(units = levels(StudyDesign$animal))
rL.site = HmscRandomLevel(units = levels(StudyDesign$site))
```

## Define and save unfitted model

The unfitted model object is saved as `hmsc/model_final`. The version in the repository (55 Tasmania samples, 576 genomes) is the canonical model for fitting and analysis. Re-running the chunks above may produce a different genome set because filtering was tuned for the final model; only rebuild if you intend to replace `hmsc/model_final` deliberately.

```{r hmsc_models_final, warning=FALSE, comments="", message=FALSE, eval=FALSE}
m = Hmsc(
  Y = YData,
  XData = XData,
  XFormula = XFormula,
  studyDesign = StudyDesign,
  phyloTree = PData,
  ranLevels = list("animal" = rL.animal, "site" = rL.site),
  distr = "normal",
  YScale = TRUE
)

save(m, file = "hmsc/model_final")
```

## In the HPC

```{sh, eval=FALSE}
conda activate hmsc-hpc

model="hmsc/model_final"
init="init/model_final.rds"

export samples=250
export thin=1000
export transient=$(( samples * thin ))
export nChains=4
export verbose=TRUE
export model
export init

# Get initiation object (hmsc.r must be available on the HPC, as for other models in chapter 13)
Rscript hmsc.r $model $init
```

```{sh, eval=FALSE}
cat <<'EOF' > launch_hmsc_model_final.sh
#!/bin/bash
#SBATCH --job-name=hmsc_model_final
#SBATCH --nodes=1
#SBATCH --partition=cpuqueue
#SBATCH --qos=normal
#SBATCH --gres=gpu:1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=1
#SBATCH --mem=32gb
#SBATCH --time=150:00:00

# Activate conda environment
source activate hmsc-hpc

# Set output files
output="output/Hmsc_model_final.rds"
mkdir -p output

# Run model fit
srun python3 -m hmsc.run_gibbs_sampler --input $init --output $output --samples $samples --transient $transient --thin $thin --verbose 100
EOF
```

```{sh, eval=FALSE}
sbatch launch_hmsc_model_final.sh
```

After the HPC run completes, upload `output/Hmsc_model_final.rds` to ERDA and update the download URL in `18_hmsc_analysis_final.Rmd`.
