# Hmsc analysis

**Note:** Load all required data objects including genome counts, metadata, GIFT database, and color schemes. Read GIFT color mapping file to create legend labels for functional groups.

```{r load_data_hmsc_18}
load("data/data.Rdata")

gift_colors <- read_tsv("data/gift_colors.tsv") %>% 
  mutate(legend=str_c(Code_function," - ",Function))
```

**Note:** Set HMSC model parameters for MCMC sampling: number of samples, thinning interval, transient (burn-in) period, and posterior support thresholds for determining significant associations.

```{r parameters_hmsc_18}
#Parameters
nSamples=250
thin=1000
transient=nSamples * thin
support_threshold=0.9
negsupport_threshold=1-support_threshold
```

**Note:** Load the unfitted HMSC model object and import the fitted model posterior samples from HPC. The fitted model is downloaded from a remote server, then posterior samples are imported and processed according to the specified sampling parameters.

```{r load_model_18, warning=FALSE, comments="", message=FALSE, cache=TRUE}
#Load unfitted
load(file = "hmsc/model_int")

#Load fitted (as a temporary file)
tmp <- tempfile(fileext = ".rds")
options(timeout = max(600, getOption("timeout"))) 
download.file("https://sid.erda.dk/share_redirect/Gh0WbM42c4/Hmsc_model_int.rds", tmp, mode = "wb")
importFromHPC = readRDS(tmp)[[1]]
unlink(tmp)

postList = importFromHPC[1:4]
fit_model_18 = importPosteriorFromHPC(m, postList, nSamples, thin, transient)
```

**Note:** Compute variance partitioning to quantify the proportion of variance explained by each environmental variable and random effect (animal, site) for each genome. Calculate predicted values and evaluate model fit using R², displaying the distribution of R² values across genomes.

```{r varpart_all_18, warning=FALSE, comments="", message=FALSE, cache=TRUE}
# Compute variance partitioning
varpart=computeVariancePartitioning(fit_model_18)

varpart$vals %>%
   as.data.frame() %>%
   rownames_to_column(var="variable") %>%
   pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
   mutate(variable=factor(variable)) %>%
   group_by(variable) %>%
   summarise(mean=mean(value)*100,sd=sd(value)*100) %>%
   tt()

preds = computePredictedValues(fit_model_18) 
MF = evaluateModelFit(hM=fit_model_18, predY=preds) 
hist(MF$R2, xlim = c(0,1), main=paste0("Mean = ", round(mean(MF$R2),2)))
```

## Correlation analysis of modeling variables

**Note:** Calculate Pearson correlations between all numeric variables used in the HMSC model. This analysis helps identify multicollinearity and understand relationships between environmental predictors, with particular focus on correlations with devil density.

```{r correlation_analysis_18, warning=FALSE, comments="", message=FALSE, fig.width=10, fig.height=8}
# Extract predictor variables from the model
predictor_data <- m$XData %>%
  as.data.frame() %>%
  # Select only numeric variables (exclude factors like development)
  select_if(is.numeric)

# Calculate Pearson correlation matrix
cor_matrix <- cor(predictor_data, use = "complete.obs", method = "pearson")

# Display full correlation matrix as table
cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "variable") %>%
  tt()

# Extract correlations with devil
devil_correlations <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "variable") %>%
  select(variable, devil) %>%
  filter(variable != "devil") %>%
  arrange(desc(abs(devil))) %>%
  rename(correlation_with_devil = devil)

# Display correlations with devil
cat("\n### Correlations with devil density:\n")
devil_correlations %>% tt()

# Create correlation plot for all variables
corrplot::corrplot(
  cor_matrix,
  method = "circle",
  type = "upper",
  order = "hclust",
  tl.cex = 0.9,
  tl.col = "black",
  addCoef.col = "black",
  number.cex = 0.8,
  col = RColorBrewer::brewer.pal(n = 8, name = "RdYlBu"),
  title = "Pearson correlation matrix of modeling variables",
  mar = c(0, 0, 2, 0)
)

# Bar plot of correlations with devil
devil_correlations %>%
  ggplot(aes(x = reorder(variable, abs(correlation_with_devil)), 
             y = correlation_with_devil,
             fill = correlation_with_devil > 0)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_fill_manual(
    values = c("FALSE" = "#d73027", "TRUE" = "#4575b4"),
    labels = c("FALSE" = "Negative", "TRUE" = "Positive"),
    name = "Correlation"
  ) +
  labs(
    x = "Variable",
    y = "Pearson correlation with devil density",
    title = "Correlations between devil density and other modeling variables"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

# Summary statistics
cat("\n### Summary of correlations with devil:\n")
cat("Strong positive (r > 0.5):", sum(devil_correlations$correlation_with_devil > 0.5, na.rm = TRUE), "\n")
cat("Moderate positive (0.3 < r ≤ 0.5):", sum(devil_correlations$correlation_with_devil > 0.3 & devil_correlations$correlation_with_devil <= 0.5, na.rm = TRUE), "\n")
cat("Weak (|r| ≤ 0.3):", sum(abs(devil_correlations$correlation_with_devil) <= 0.3, na.rm = TRUE), "\n")
cat("Moderate negative (-0.5 ≤ r < -0.3):", sum(devil_correlations$correlation_with_devil >= -0.5 & devil_correlations$correlation_with_devil < -0.3, na.rm = TRUE), "\n")
cat("Strong negative (r < -0.5):", sum(devil_correlations$correlation_with_devil < -0.5, na.rm = TRUE), "\n")
```

## MCMC convergence diagnostics

**Note:** Assess MCMC chain convergence using multiple diagnostics: R-hat (Gelman-Rubin statistic), effective sample size (ESS), trace plots, and autocorrelation. Good convergence is indicated by R-hat < 1.1, ESS > 400, and well-mixed chains without trends.

```{r mcmc_convergence_18, warning=FALSE, comments="", message=FALSE, fig.width=12, fig.height=10, eval=FALSE}
# Convert posterior to coda format for diagnostics
mpost <- convertToCodaObject(fit_model_18)

# Load coda package for diagnostics (if not already loaded)
if (!requireNamespace("coda", quietly = TRUE)) {
  stop("Please install coda package: install.packages('coda')")
}
library(coda)

# 1. R-hat (Gelman-Rubin diagnostic) - should be < 1.1 for good convergence
cat("=== R-hat (Gelman-Rubin) Diagnostic ===\n")
cat("Values should be < 1.1 for good convergence\n\n")

# Calculate R-hat for Beta coefficients
if (!is.null(mpost$Beta)) {
  beta_rhat <- gelman.diag(mpost$Beta, multivariate = FALSE)
  cat("Beta coefficients R-hat:\n")
  print(beta_rhat$psrf)
  
  # Summary of R-hat values
  beta_rhat_values <- beta_rhat$psrf[, "Point est."]
  cat("\nBeta R-hat summary:\n")
  cat("  Mean:", round(mean(beta_rhat_values, na.rm = TRUE), 4), "\n")
  cat("  Max:", round(max(beta_rhat_values, na.rm = TRUE), 4), "\n")
  cat("  Values > 1.1:", sum(beta_rhat_values > 1.1, na.rm = TRUE), "out of", length(beta_rhat_values), "\n")
}

# Calculate R-hat for Rho (phylogenetic signal)
if (!is.null(mpost$Rho)) {
  rho_rhat <- gelman.diag(mpost$Rho, multivariate = FALSE)
  cat("\nRho (phylogenetic signal) R-hat:\n")
  print(rho_rhat$psrf)
}

# Calculate R-hat for V (variance components)
if (!is.null(mpost$V)) {
  v_rhat <- gelman.diag(mpost$V, multivariate = FALSE)
  cat("\nV (variance components) R-hat:\n")
  print(v_rhat$psrf)
}

# 2. Effective Sample Size (ESS) - should be > 400 for reliable estimates
cat("\n\n=== Effective Sample Size (ESS) ===\n")
cat("Values should be > 400 for reliable estimates\n\n")

if (!is.null(mpost$Beta)) {
  beta_ess <- effectiveSize(mpost$Beta)
  cat("Beta coefficients ESS:\n")
  cat("  Mean:", round(mean(beta_ess, na.rm = TRUE), 2), "\n")
  cat("  Min:", round(min(beta_ess, na.rm = TRUE), 2), "\n")
  cat("  Values < 400:", sum(beta_ess < 400, na.rm = TRUE), "out of", length(beta_ess), "\n")
  
  # Show ESS for key parameters (e.g., first few Beta coefficients)
  cat("\nESS for first 10 Beta coefficients:\n")
  print(head(beta_ess, 10))
}

if (!is.null(mpost$Rho)) {
  rho_ess <- effectiveSize(mpost$Rho)
  cat("\nRho ESS:\n")
  print(rho_ess)
}

# 3. Trace plots for key parameters
cat("\n\n=== Trace Plots ===\n")
cat("Chains should be well-mixed (no trends, good overlap between chains)\n\n")

# Trace plot for Rho (phylogenetic signal)
if (!is.null(mpost$Rho)) {
  par(mfrow = c(2, 2))
  traceplot(mpost$Rho, main = "Rho (phylogenetic signal)")
}

# Trace plots for a few Beta coefficients (intercept and key predictors)
if (!is.null(mpost$Beta)) {
  # Get Beta for intercept and devil, temperature, temperature:devil
  beta_names <- dimnames(mpost$Beta[[1]])[[2]]
  
  # Find indices for key parameters
  intercept_idx <- grep("Intercept", beta_names, ignore.case = TRUE)[1]
  devil_idx <- grep("devil", beta_names, ignore.case = TRUE)[1]
  temp_idx <- grep("^temperature$", beta_names, ignore.case = TRUE)[1]
  interaction_idx <- grep("temperature.*devil|devil.*temperature", beta_names, ignore.case = TRUE)[1]
  
  key_indices <- c(intercept_idx, devil_idx, temp_idx, interaction_idx)
  key_indices <- key_indices[!is.na(key_indices)]
  
  if (length(key_indices) > 0) {
    par(mfrow = c(2, 2))
    for (idx in key_indices[1:min(4, length(key_indices))]) {
      if (idx <= length(beta_names)) {
        # Extract this parameter across all chains
        param_chains <- lapply(mpost$Beta, function(x) x[, idx])
        param_mcmc <- mcmc.list(param_chains)
        traceplot(param_mcmc, main = paste("Beta:", beta_names[idx]))
      }
    }
  }
}

# 4. Autocorrelation plots
cat("\n\n=== Autocorrelation ===\n")
cat("Autocorrelation should drop quickly to near zero\n\n")

if (!is.null(mpost$Rho)) {
  par(mfrow = c(1, 1))
  autocorr.plot(mpost$Rho, main = "Autocorrelation: Rho", lag.max = 50)
}

# 5. Summary table of convergence diagnostics
cat("\n\n=== Convergence Summary Table ===\n")

convergence_summary <- data.frame(
  Parameter = character(),
  Rhat_mean = numeric(),
  Rhat_max = numeric(),
  Rhat_over_threshold = numeric(),
  ESS_mean = numeric(),
  ESS_min = numeric(),
  ESS_under_threshold = numeric(),
  stringsAsFactors = FALSE
)

if (!is.null(mpost$Beta)) {
  beta_rhat_vals <- beta_rhat$psrf[, "Point est."]
  convergence_summary <- rbind(convergence_summary, data.frame(
    Parameter = "Beta",
    Rhat_mean = round(mean(beta_rhat_vals, na.rm = TRUE), 4),
    Rhat_max = round(max(beta_rhat_vals, na.rm = TRUE), 4),
    Rhat_over_threshold = sum(beta_rhat_vals > 1.1, na.rm = TRUE),
    ESS_mean = round(mean(beta_ess, na.rm = TRUE), 2),
    ESS_min = round(min(beta_ess, na.rm = TRUE), 2),
    ESS_under_threshold = sum(beta_ess < 400, na.rm = TRUE)
  ))
}

if (!is.null(mpost$Rho)) {
  rho_rhat_vals <- rho_rhat$psrf[, "Point est."]
  convergence_summary <- rbind(convergence_summary, data.frame(
    Parameter = "Rho",
    Rhat_mean = round(mean(rho_rhat_vals, na.rm = TRUE), 4),
    Rhat_max = round(max(rho_rhat_vals, na.rm = TRUE), 4),
    Rhat_over_threshold = sum(rho_rhat_vals > 1.1, na.rm = TRUE),
    ESS_mean = round(mean(rho_ess, na.rm = TRUE), 2),
    ESS_min = round(min(rho_ess, na.rm = TRUE), 2),
    ESS_under_threshold = sum(rho_ess < 400, na.rm = TRUE)
  ))
}

convergence_summary %>% tt()

# 6. Overall convergence assessment
cat("\n=== Overall Convergence Assessment ===\n")
all_rhat <- c()
all_ess <- c()

if (!is.null(mpost$Beta)) {
  all_rhat <- c(all_rhat, beta_rhat_values)
  all_ess <- c(all_ess, beta_ess)
}
if (!is.null(mpost$Rho)) {
  all_rhat <- c(all_rhat, rho_rhat_vals)
  all_ess <- c(all_ess, rho_ess)
}

if (length(all_rhat) > 0) {
  max_rhat <- max(all_rhat, na.rm = TRUE)
  min_ess <- min(all_ess, na.rm = TRUE)
  
  cat("Maximum R-hat:", round(max_rhat, 4), "\n")
  cat("Minimum ESS:", round(min_ess, 2), "\n\n")
  
  if (max_rhat < 1.1 && min_ess > 400) {
    cat("✓ CONVERGENCE: GOOD\n")
    cat("  All R-hat values < 1.1 and ESS > 400\n")
  } else if (max_rhat < 1.2 && min_ess > 200) {
    cat("⚠ CONVERGENCE: ACCEPTABLE\n")
    cat("  Some parameters may need more samples\n")
  } else {
    cat("✗ CONVERGENCE: POOR\n")
    cat("  Consider running more MCMC iterations\n")
  }
}

# Reset plotting parameters
par(mfrow = c(1, 1))
```

**Note:** Create a phylogenetic tree visualization with variance partitioning results displayed as stacked bars next to each tip. The tree shows phylum annotations and variance contributions from each predictor variable, allowing visualization of how environmental and random effects vary across the phylogenetic tree.

```{r varpart_tree_plot_18, warning=FALSE, message=FALSE, fig.height=8, fig.width=7, eval = FALSE}
# 1) Build the phylogeny in the same species set/order as varpart
vp_tree_phylo <- genome_tree %>%
  keep.tip(tip = colnames(varpart$vals)) %>%
  force.ultrametric(method = "extend")

vp_tree <- ggtree(vp_tree_phylo, size = 0.3)

# 2) Phylum annotation in tree tip order
phylums_vp <- genome_metadata %>%
  filter(genome %in% vp_tree$data$label) %>%
  arrange(match(genome, vp_tree$data$label)) %>%
  mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
  column_to_rownames("genome") %>%
  select(phylum)

vp_tree <- gheatmap(
  vp_tree,
  phylums_vp,
  offset    = -0.2,
  width     = 0.1,
  colnames  = FALSE
) +
  scale_fill_manual(values = phylum_colors) +
  labs(fill = "Phylum")

# start a new fill scale for the variance bars
vp_tree <- vp_tree + new_scale_fill()

# 3) Long variance-partitioning table
varpart_long <- varpart$vals %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  pivot_longer(
    cols      = -variable,
    names_to  = "label",   # must be 'label' to match tree tip labels
    values_to = "value"
  ) %>%
  mutate(
    # reorder predictors in a sensible way for the legend
    variable = factor(
      variable,
      levels = rev(c(
        "devil", "temperature", "density",
        "development", "weight",
        "precipitation", "diversity",
        "logseqdepth",
        "Random: animal", "Random: site"
      ))
    )
  )

# 4) Add stacked bars with geom_fruit
vp_tree +
  geom_fruit(
    data   = varpart_long,
    geom   = geom_bar,
    mapping = aes(
      x    = value,
      y    = label,
      fill = variable
    ),
    stat        = "identity",
    orientation = "y",
    pwidth      = 2,
    offset      = 0.05
  ) +
  labs(fill = "Variable")

```
## Phylogenetic signal

**Note:** Assess phylogenetic signal in the model by extracting the Rho parameter from the posterior distribution. Rho quantifies the strength of phylogenetic correlation in species responses, with values near 1 indicating strong phylogenetic signal and values near 0 indicating weak signal. Calculate credible intervals and visualize the posterior distribution.

```{r phylogenetic_signal_18, warning=FALSE, message=FALSE, eval = FALSE}
# Convert to coda object to inspect Rho (phylogenetic signal)
mpost <- convertToCodaObject(fit_model_18)

# Overall Rho posterior (phylogenetic signal on the species random effect)
rho_vec <- unlist(mpost$Rho)

# 90% credible interval and median
rho_summary <- quantile(rho_vec, probs = c(0.05, 0.5, 0.95))
rho_summary

# Optional: density plot for the appendix
plot(density(rho_vec),
     main = "Posterior of phylogenetic signal (Rho)",
     xlab = "Rho")
abline(v = rho_summary[2], col = "red", lty = 2)
```

## Create posterior estimates table

**Note:** Extract posterior support values and create a table for downstream analysis. This is separated from the visualization chunk to allow execution even when ggtree plots are disabled.

```{r post_table_18, warning=FALSE, comments="", message=FALSE, cache=TRUE}
hmsc_tree <- genome_tree %>%
        keep.tip(., tip=m$spNames) 

# Posterior estimates 
post_estimates <- getPostEstimate(hM=fit_model_18, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support_threshold ~ "Positive",
          value <= negsupport_threshold ~ "Negative",
          TRUE ~ "Neutral")) 

post_table <- post_estimates %>%
    mutate(genome=factor(genome, levels=rev(hmsc_tree$tip.label))) %>%
    select(-trend) %>%
    mutate(value = case_when(
          value >= support_threshold ~ "Positive",
          value <= negsupport_threshold ~ "Negative",
          TRUE ~ "Neutral")) %>%
    mutate(value=factor(value, levels=c("Positive","Neutral","Negative"))) %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    rename(intercept=2) %>%
    rename(temperature_devil = `temperature:devil`) %>%
   column_to_rownames(var="genome") %>%
  select(-intercept)
```

## Posterior estimates visualization

**Note:** Visualize posterior support values on a phylogenetic tree. This chunk is set to `eval = FALSE` due to ggtree compatibility issues.

```{r posteriors_18, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE, eval = FALSE} 

# Basal ggtree
postestimates_tree <- hmsc_tree %>%
        force.ultrametric(.,method="extend") %>%
        ggtree(., size = 0.3)
        

# Phylums
phylums <- genome_metadata %>%
    filter(genome %in% hmsc_tree$tip.label) %>%
    arrange(match(genome, hmsc_tree$tip.label)) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome") %>%
    select(phylum)


#Add phylum colors next to the tree tips
postestimates_tree <- gheatmap(postestimates_tree, phylums, offset=-0.2, width=0.1, colnames=FALSE) +
      scale_fill_manual(values=phylum_colors)+
      labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
postestimates_tree <- postestimates_tree + new_scale_fill()

# Add posterior significant heatmap

postestimates_tree <- gheatmap(postestimates_tree, post_table, offset=0, width=0.5, colnames=TRUE, colnames_position="top",colnames_angle=90, colnames_offset_y=1, hjust=0) +
        scale_fill_manual(values=c("#be3e2b","#f4f4f4","#b2b530"))+
        labs(fill="Trend") 

postestimates_tree +
        vexpand(.30, 1)  # expand top 

```

**Note:** Examine the congruence between devil and temperature associations at the taxonomic level. Create contingency tables showing how many genomes have each combination of devil and temperature responses (positive, negative, neutral). Visualize taxonomic patterns by phylum to identify whether certain taxonomic groups show consistent response patterns.

```{r congruence_devil_temperature_taxa_18, warning=FALSE, message=FALSE}
# Long table of trends for devil and temperature per genome
devil_temp_trends <- post_table %>%
  rownames_to_column("genome") %>%
  select(genome, devil, temperature) %>%
  left_join(genome_metadata, by = "genome")

# Overall contingency of trends
devil_temp_tab <- devil_temp_trends %>%
  count(devil, temperature) %>%
  arrange(desc(n))

devil_temp_tab %>% tt()

# Phylum-level congruence: how many genomes in each combo
devil_temp_phylum <- devil_temp_trends %>%
  group_by(phylum, devil, temperature) %>%
  summarise(n = n(), .groups = "drop")

# Keep only non-zero combos (for cleaner plots)
devil_temp_phylum %>%
  filter(n > 0) %>%
  ggplot(aes(x = devil, y = temperature, fill = n)) +
  geom_tile() +
  facet_wrap(~ phylum) +
  scale_fill_gradient(low = "lightgrey", high = "blue") +
  labs(
    x = "Devil association",
    y = "Temperature association",
    fill = "Number of genomes",
    title = "Taxonomic congruence of devil vs temperature responses"
  ) +
  theme_minimal(base_size = 8)
```

**Note:** Map GIFT functional annotations (bundle-level) to element-level for genomes included in the HMSC model. Join functional trait data with posterior support trends to enable analysis of functional differences between genomes with different environmental associations.

```{r functional_response_18, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
elements_response <- genome_gifts[rownames(post_table),colnames(genome_gifts) %in% GIFT_db$Code_bundle] %>%
  to.elements(.,GIFT_db=GIFT_db) %>% 
  as.data.frame() %>% 
  rownames_to_column(var="genome") %>% 
  pivot_longer(!genome,names_to="GIFT", values_to="value") %>% 
  left_join(post_table %>% rownames_to_column(var = "genome"), by="genome")
```

**Note:** Examine functional congruence between devil and temperature associations. Filter to genomes with non-neutral responses for both variables, then summarize functional element signals by devil-temperature response combinations. Aggregate to functional groups and visualize the top 12 most variable functional groups to identify functional traits that show different patterns across response combinations.

```{r congruence_devil_temperature_functional_18, warning=FALSE, message=FALSE}
# elements_response has: genome, GIFT, value, and trend columns (devil, temperature, ...)
# Focus on non-neutral responses for both variables
elem_dev_temp <- elements_response %>%
  filter(devil != "Neutral",
         temperature != "Neutral") %>%
  mutate(
    devil      = droplevels(devil),
    temperature = droplevels(temperature)
  )

# Summarise at the element level: mean value per devil/temperature combo
func_congruence_elem <- elem_dev_temp %>%
  group_by(GIFT, devil, temperature) %>%
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    n        = n(),
    .groups  = "drop"
  )

# Collapse to functional group (Code_function)
func_congruence_group <- func_congruence_elem %>%
  left_join(GIFT_db, by = c("GIFT" = "Code_bundle")) %>%
  group_by(Code_function, devil, temperature) %>%
  summarise(
    mean_val = mean(mean_val, na.rm = TRUE),
    .groups  = "drop"
  )

# For readability, keep only functional groups that show some variation
interesting_funcs <- func_congruence_group %>%
  group_by(Code_function) %>%
  summarise(range_val = max(mean_val) - min(mean_val), .groups = "drop") %>%
  slice_max(order_by = range_val, n = 12) %>%
  pull(Code_function)

func_congruence_group %>%
  filter(Code_function %in% interesting_funcs) %>%
  ggplot(aes(x = devil, y = temperature, fill = mean_val)) +
  geom_tile() +
  facet_wrap(~ Code_function) +
  scale_fill_gradient2() +
  labs(
    x = "Devil association",
    y = "Temperature association",
    fill = "Mean element signal",
    title = "Functional congruence of devil vs temperature responses"
  ) +
  theme_minimal(base_size = 8)
``` 

## Association with devil

**Note:** Summarize devil associations across taxonomic levels. Count and calculate percentages of genomes with positive and negative devil associations at phylum, order, family, and genus levels. This helps identify which taxonomic groups show the strongest associations with devil presence.

```{r association_devil_18, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
post_table$devil %>% table()
  
devil_phylum_perc <- post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  group_by(phylum) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100)

devil_phylum_perc %>% 
  arrange(-percent_positive)

devil_phylum_perc %>% 
  arrange(-percent_negative)

# Sub-phylum level

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(phylum=="p__Bacillota_A") %>% 
  group_by(order) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(order=="o__TANB77") %>% 
  group_by(family) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(family=="f__CAG-508") %>% 
  group_by(genus) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100) %>% 
  arrange(-percent_positive)
```
**Note:** Extract posterior mean and support values for the devil coefficient (β) for each genome. Create a scatter plot showing the relationship between the mean β coefficient and posterior support, colored by phylum. This visualization helps identify genomes with strong, well-supported associations with devil presence.

```{r beta_devil_18, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
# Mean beta for 'devil' per genome
devil_estimate <- getPostEstimate(hM = fit_model_18, parName = "Beta")$mean %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "devil") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
  dplyr::select(genome, mean)

# Posterior support for 'devil' per genome
devil_support <- getPostEstimate(hM = fit_model_18, parName = "Beta")$support %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "devil") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
  dplyr::select(genome, support)

devil_beta <- inner_join(devil_estimate, devil_support, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

devil_beta %>%
  mutate(
    significant   = support >= support_threshold | support <= negsupport_threshold,
    support_plot  = ifelse(mean < 0, 1 - support, support), # support for the sign of beta
    phylum_sig    = ifelse(significant & support_plot > 0.9, phylum, NA)
  ) %>%
  ggplot(aes(x = mean, y = support_plot, colour = phylum_sig)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = phylum_colors, na.translate = FALSE) +
  geom_vline(xintercept = 0) +
  labs(
    x = "Devil β coefficient",
    y = "Posterior probability (|β| > 0)",
    colour = "Phylum"
  ) +
  theme_minimal()

```
**Note:** Calculate 80% credible intervals (10th and 90th percentiles) for the devil β coefficient for each genome. Filter to genomes with credible non-zero effects (where the credible interval does not include zero) and visualize these with error bars, ordered by mean β value.

```{r devil_credible_intervals_18, warning=FALSE, comments="", message=FALSE}
# Credible intervals for 'devil'
devil_q <- getPostEstimate(hM = fit_model_18,
                           parName = "Beta",
                           q = c(0.1, 0.9))$q

idx_devil <- which(m$covNames == "devil")
n_genomes <- length(m$spNames)

devil_ci <- map_dfr(1:n_genomes, ~ {
  tibble(
    genome = m$spNames[.x],
    q10 = devil_q[1, idx_devil, .x],
    q90 = devil_q[2, idx_devil, .x]
  )
}) %>%
  right_join(devil_estimate, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

# Genomes with credible non-zero devil effect
devil_ci %>%
  filter(q10 > 0 | q90 < 0) %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal() +
  labs(x = "Devil β coefficient",
       y = "Genomes") +
  theme(legend.position = "none")
```
**Note:** Identify and visualize the top 30 genomes with the strongest positive and top 30 with the strongest negative credible devil associations. Create a forest plot showing these top genomes with their credible intervals, ordered by mean β coefficient and colored by phylum.

```{r devil_topgenomes_18, warning=FALSE, comments="", message=FALSE}
## Association with devil – credible intervals (top genomes)

# Keep only genomes with credible non-zero effect
devil_ci_sig <- devil_ci %>%
  filter(q10 > 0 | q90 < 0)

# 4. Take top N positive and top N negative genomes
N_top <- 30  

devil_ci_top <- devil_ci_sig %>%
  slice_max(order_by = mean, n = N_top, with_ties = FALSE) %>%
  bind_rows(
    devil_ci_sig %>%
      slice_min(order_by = mean, n = N_top, with_ties = FALSE)
  ) %>%
  distinct(genome, .keep_all = TRUE)

# 5. Plot
devil_ci_top %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal(base_size = 9) +
  labs(
    x = "Devil β coefficient",
    y = "Top genomes with credible devil effect"
  ) +
  theme(legend.position = "right")

```
**Note:** Summarize devil β coefficients at the phylum level by calculating mean β and 80% credible intervals (10th and 90th percentiles) for each phylum. Visualize phylum-level associations with error bars to show the overall pattern of devil associations across major taxonomic groups.

```{r devil_phylum_ci_18, warning=FALSE, comments="", message=FALSE}
devil_phylum_ci <- devil_ci %>%
  group_by(phylum) %>%
  summarise(
    mean_beta = mean(mean),
    q10 = quantile(mean, 0.1),
    q90 = quantile(mean, 0.9),
    .groups = "drop"
  )

devil_phylum_ci %>%
  ggplot(aes(x = mean_beta,
             y = forcats::fct_reorder(phylum, mean_beta),
             xmin = q10, xmax = q90)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  theme_minimal() +
  labs(x = "Mean devil β (per phylum)",
       y = "Phylum")
```
**Note:** Create summary tables of genomes with credible positive and negative devil associations. Filter genomes where the 80% credible interval excludes zero, then extract taxonomic information and β statistics. Display these as formatted tables showing the genomes with the strongest associations.

```{r devil_overview_18, warning=FALSE, comments="", message=FALSE}
devil_positive <- devil_ci %>%
  filter(q10 > 0) %>%
  arrange(-mean) %>%
  dplyr::select(genome, phylum, class, order, family, genus, species, mean, q10, q90)

devil_negative <- devil_ci %>%
  filter(q90 < 0) %>%
  arrange(mean) %>%
  dplyr::select(genome, phylum, class, order, family, genus, species, mean, q10, q90)

devil_positive %>% tt()  # tinytable
devil_negative %>% tt()
```

**Note:** Compare functional element (GIFT) abundances between genomes with positive versus negative devil associations. For each functional element, calculate mean abundances in each group, compute differences, and perform t-tests. Adjust p-values using Benjamini-Hochberg FDR correction to identify functional elements that differ significantly between devil-associated groups.

```{r devil_functions_18, warning=FALSE, comments="", message=FALSE}
functional_differences_devil <- elements_response %>% 
  filter(devil !="Neutral") %>% 
  mutate(devil = droplevels(devil)) %>%
  group_by(GIFT) %>%
  summarise(
    n_pos   = sum(devil == "Positive"),
    n_neg   = sum(devil == "Negative"),
    mean_pos = mean(value[devil == "Positive"], na.rm = TRUE),
    mean_neg = mean(value[devil == "Negative"], na.rm = TRUE),
    diff     = mean_pos - mean_neg,
    p        = if (n_pos > 1 && n_neg > 1 &&
                   stats::sd(value[devil == "Positive"], na.rm = TRUE) > 0 &&
                   stats::sd(value[devil == "Negative"], na.rm = TRUE) > 0) {
                stats::t.test(value ~ devil, var.equal = FALSE)$p.value
              } else NA_real_
  , .groups = "drop") %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)
```


**Note:** Create a volcano plot visualizing functional differences between positive and negative devil-associated genomes. Plot mean difference (fold change) against -log10(FDR), with functional elements colored by their functional group. Highlight significant elements that meet both fold change and FDR thresholds, with labels for the most significant hits.

```{r devil_functions_plot_18, warning=FALSE, comments="", message=FALSE}
fc_threshold <- 0.2      # minimum absolute mean difference to call interesting
fdr_threshold <- 0.05    # BH-adjusted p-value cutoff

functional_differences_devil_logfold <- functional_differences_devil %>%
  mutate(func = substr(GIFT, 1, 3)) %>% 
  mutate(
    # guard against p_adj = 0 or NA
    p_adj_plot = ifelse(is.na(p_adj) | p_adj == 0, .Machine$double.xmin, p_adj),
    neglog10_fdr = -log10(p_adj_plot),
    hit = !is.na(p_adj) & p_adj < fdr_threshold & abs(diff) >= fc_threshold
  ) %>% 
  mutate(func = ifelse(hit, func, "NS")) %>% 
  filter(neglog10_fdr<20)

palette_df <- functional_differences_devil_logfold %>%
  dplyr::distinct(func) %>%
  dplyr::left_join(
    gift_colors %>% dplyr::select(func = Code_function, Color),
    by = "func"
  ) %>%
  dplyr::mutate(
    Color = dplyr::case_when(
      func == "NS" & is.na(Color) ~ "grey70",
      is.na(Color)                ~ "#999999",
      TRUE                        ~ Color
    )
  )

color_values <- setNames(palette_df$Color, palette_df$func)

functional_differences_devil_logfold %>% 
  ggplot(aes(x = diff, y = neglog10_fdr, color=func)) +
  geom_point(aes(color = func), alpha = 0.8, size = 2) +
  geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed") +
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed") +
  ylim(0,15)+
  xlim(-0.6,0.6)+
  ggrepel::geom_text_repel(
    data = subset(functional_differences_devil_logfold, hit),
    aes(label = GIFT),
    size = 3, max.overlaps = 20, box.padding = 0.3, min.segment.length = 0
  ) +
  scale_color_manual(values = color_values, breaks = names(color_values)) +
  labs(
    x = "Mean difference (Positive − Negative)",
    y = expression(-log[10]("FDR (BH)")),
    color = "Group",
    subtitle = paste0("Thresholds: |diff| ≥ ", fc_threshold, ", FDR < ", fdr_threshold)
  ) +
  theme_minimal(base_size = 12)
```

**Note:** Aggregate devil β coefficients to the functional group level by calculating mean β across all genomes carrying each functional trait. Identify the top 15 functional groups with strongest positive and negative associations, then create a forest plot showing functional group-level associations with credible intervals, colored by functional group.

```{r devil_function_top_forest_18, warning=FALSE, message=FALSE, fig.height=7}
# Map genomes to functional groups (GIFT 3-letter "func" code)
genome_func <- elements_response %>%
  dplyr::filter(value > 0) %>%
  dplyr::mutate(func = substr(GIFT, 1, 3)) %>%
  dplyr::distinct(genome, func)

# Summarise devil β across genomes for each functional group
devil_func_summary <- genome_func %>%
  dplyr::left_join(devil_ci %>% dplyr::select(genome, mean, q10, q90), by = "genome") %>%
  dplyr::group_by(func) %>%
  dplyr::summarise(
    n_genomes = sum(!is.na(mean)),
    mean_beta = mean(mean, na.rm = TRUE),
    q10_fun   = stats::quantile(mean, 0.10, na.rm = TRUE),
    q90_fun   = stats::quantile(mean, 0.90, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::filter(n_genomes >= 3)   # require at least 3 genomes per function

# 4) Take top 15 positive & top 15 negative functions
N_top <- 15

devil_func_top <- dplyr::bind_rows(
  devil_func_summary %>%
    dplyr::slice_max(order_by = mean_beta, n = N_top, with_ties = FALSE),
  devil_func_summary %>%
    dplyr::slice_min(order_by = mean_beta, n = N_top, with_ties = FALSE)
) %>%
  dplyr::distinct(func, .keep_all = TRUE)

# 5) Add nice labels & colours from gift_colors
devil_func_top <- devil_func_top %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::transmute(
        func        = Code_function,   # 3-letter code
        func_label  = legend,          # "B01 – Nucleic acid biosynthesis", etc.
        color       = Color
      ),
    by = "func"
  ) %>%
  dplyr::mutate(
    func_label = ifelse(is.na(func_label), func, func_label)
  )

# Colour lookup
func_cols <- setNames(devil_func_top$color, devil_func_top$func_label)

# 6) Forest plot: functions instead of genomes
ggplot(
  devil_func_top,
  aes(x = mean_beta,
      y = forcats::fct_reorder(func_label, mean_beta),
      xmin = q10_fun, xmax = q90_fun,
      colour = func_label)
) +
  geom_point(size = 2) +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = func_cols, guide = "none") +
  theme_minimal(base_size = 10) +
  labs(
    x = "Devil β (mean across genomes carrying the function)",
    y = "Functional trait (GIFT group)"
  )

```


## Association with temperature

**Note:** Summarize temperature associations across taxonomic levels. Count and calculate percentages of genomes with positive and negative temperature associations at phylum, order, family, and genus levels. This helps identify which taxonomic groups show the strongest associations with temperature.

```{r association_temperature_18, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
post_table$temperature %>% table()
  
temp_phylum_perc <- post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  group_by(phylum) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100)

temp_phylum_perc %>% 
  arrange(-percent_positive)

temp_phylum_perc %>% 
  arrange(-percent_negative)

# Sub-phylum level

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(phylum=="p__Bacillota_A") %>% 
  group_by(order) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(order=="o__TANB77") %>% 
  group_by(family) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(family=="f__CAG-508") %>% 
  group_by(genus) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100) %>% 
  arrange(-percent_positive)
```
**Note:** Extract posterior mean and support values for the temperature coefficient (β) for each genome. Create a scatter plot showing the relationship between the mean β coefficient and posterior support, colored by phylum. This visualization helps identify genomes with strong, well-supported associations with temperature.

```{r beta_temperature_18, warning=FALSE, comments="", message=FALSE}
temp_estimate <- getPostEstimate(hM = fit_model_18, parName = "Beta")$mean %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "temperature") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
  dplyr::select(genome, mean)

temp_support <- getPostEstimate(hM = fit_model_18, parName = "Beta")$support %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "temperature") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
  dplyr::select(genome, support)

temp_beta <- inner_join(temp_estimate, temp_support, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

temp_beta %>%
  mutate(significant = support >= support_threshold | support <= negsupport_threshold,
         support_plot = ifelse(mean < 0, 1 - support, support),
         phylum_sig = ifelse(significant & support_plot > 0.9, phylum, NA)) %>%
  ggplot(aes(x = mean, y = support_plot, colour = phylum_sig)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = phylum_colors, na.translate = FALSE) +
  geom_vline(xintercept = 0) +
  labs(x = "Temperature β coefficient",
       y = "Posterior probability (|β| > 0)",
       colour = "Phylum") +
  theme_minimal()
```
**Note:** Calculate 80% credible intervals (10th and 90th percentiles) for the temperature β coefficient for each genome. Filter to genomes with credible non-zero effects (where the credible interval does not include zero) and visualize these with error bars, ordered by mean β value.

```{r temperature_credible_intervals_18, warning=FALSE, comments="", message=FALSE}
temp_q <- getPostEstimate(hM = fit_model_18,
                          parName = "Beta",
                          q = c(0.1, 0.9))$q
idx_temp <- which(m$covNames == "temperature")
n_genomes <- length(m$spNames)

temp_ci <- map_dfr(1:n_genomes, ~ {
  tibble(
    genome = m$spNames[.x],
    q10 = temp_q[1, idx_temp, .x],
    q90 = temp_q[2, idx_temp, .x]
  )
}) %>%
  right_join(temp_estimate, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

temp_ci %>%
  filter(q10 > 0 | q90 < 0) %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal() +
  labs(x = "Temperature β coefficient",
       y = "Genomes") +
  theme(legend.position = "none")
```

**Note:** Identify and visualize the top 30 genomes with the strongest positive and top 30 with the strongest negative credible temperature associations. Create a forest plot showing these top genomes with their credible intervals, faceted by association direction (positive/negative), ordered by mean β coefficient and colored by phylum.

```{r temperature_ci_top30_18, fig.height=10, fig.width=6}
# --- Temperature credible intervals: top 30 genomes ---

# Keep only genomes with credible non-zero temperature effect
temp_ci_sig <- temp_ci %>%
  dplyr::filter(q10 > 0 | q90 < 0)

# 4. Take top 30 positive and top 30 negative genomes
N_top <- 30

temp_ci_top <- temp_ci_sig %>%
  dplyr::slice_max(order_by = mean, n = N_top, with_ties = FALSE) %>%
  dplyr::bind_rows(
    temp_ci_sig %>%
      dplyr::slice_min(order_by = mean, n = N_top, with_ties = FALSE)
  ) %>%
  dplyr::distinct(genome, .keep_all = TRUE) %>%
  dplyr::mutate(
    direction = ifelse(mean > 0, "Positive association", "Negative association"),
    genome    = forcats::fct_reorder(genome, mean)
  )

# 5. Plot credible intervals for these top genomes
temp_ci_top %>%
  ggplot2::ggplot(ggplot2::aes(x = mean,
                               y = genome,
                               xmin = q10, xmax = q90,
                               colour = phylum)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(width = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::scale_color_manual(values = phylum_colors) +
  ggplot2::facet_grid(direction ~ ., scales = "free_y", space = "free_y") +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::labs(
    x = "Temperature β coefficient",
    y = "Top genomes with credible temperature effect"
  ) +
  ggplot2::theme(legend.position = "right")
```

---

### 2) Temperature – phylum-level summary of β

**Note:** Summarize temperature β coefficients at the phylum level by calculating mean β and 80% credible intervals (10th and 90th percentiles) for each phylum. Visualize phylum-level associations with error bars to show the overall pattern of temperature associations across major taxonomic groups.

```{r temperature_phylum_ci_18, warning=FALSE, message=FALSE}
temp_phylum_ci <- temp_ci %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarise(
    mean_beta = mean(mean, na.rm = TRUE),
    q10       = stats::quantile(mean, 0.1, na.rm = TRUE),
    q90       = stats::quantile(mean, 0.9, na.rm = TRUE),
    .groups   = "drop"
  )

temp_phylum_ci %>%
  ggplot2::ggplot(ggplot2::aes(x = mean_beta,
                               y = forcats::fct_reorder(phylum, mean_beta),
                               xmin = q10, xmax = q90)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(width = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    x = "Mean temperature β (per phylum)",
    y = "Phylum"
  )

```


**Note:** Compare functional element (GIFT) abundances between genomes with positive versus negative temperature associations. For each functional element, calculate mean abundances in each group, compute differences, and perform t-tests. Adjust p-values using Benjamini-Hochberg FDR correction to identify functional elements that differ significantly between temperature-associated groups.

```{r functional_differences_temperature_18, warning=FALSE, comments="", message=FALSE}
functional_differences_temperature <- elements_response %>% 
  filter(temperature !="Neutral") %>% 
  mutate(temperature = droplevels(temperature)) %>%
  group_by(GIFT) %>%
  summarise(
    n_pos   = sum(temperature == "Positive"),
    n_neg   = sum(temperature == "Negative"),
    mean_pos = mean(value[temperature == "Positive"], na.rm = TRUE),
    mean_neg = mean(value[temperature == "Negative"], na.rm = TRUE),
    diff     = mean_pos - mean_neg,
    p        = if (n_pos > 1 && n_neg > 1 &&
                   stats::sd(value[temperature == "Positive"], na.rm = TRUE) > 0 &&
                   stats::sd(value[temperature == "Negative"], na.rm = TRUE) > 0) {
                stats::t.test(value ~ temperature, var.equal = FALSE)$p.value
              } else NA_real_
  , .groups = "drop") %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)
```


**Note:** Create a volcano plot visualizing functional differences between positive and negative temperature-associated genomes. Plot mean difference (fold change) against -log10(FDR), with functional elements colored by their functional group. Highlight significant elements that meet both fold change and FDR thresholds, with labels for the most significant hits.

```{r temperature_functions_plot_18, warning=FALSE, comments="", message=FALSE}
fc_threshold <- 0.2      # minimum absolute mean difference to call interesting
fdr_threshold <- 0.05    # BH-adjusted p-value cutoff

functional_differences_temperature_logfold <- functional_differences_temperature %>%
  mutate(func = substr(GIFT, 1, 3)) %>% 
  mutate(
    # guard against p_adj = 0 or NA
    p_adj_plot   = ifelse(is.na(p_adj) | p_adj == 0, .Machine$double.xmin, p_adj),
    neglog10_fdr = -log10(p_adj_plot),
    hit          = !is.na(p_adj) & p_adj < fdr_threshold & abs(diff) >= fc_threshold
  ) %>% 
  mutate(func = ifelse(hit, func, "NS")) %>%   # NS = non-significant / background
  filter(neglog10_fdr < 20)

# ---- Build a safe colour vector for all func levels ----
func_levels <- sort(unique(functional_differences_temperature_logfold$func))

palette_df <- tibble::tibble(func = func_levels) %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::select(func = Code_function, Color),
    by = "func"
  ) %>%
  dplyr::mutate(
    Color = dplyr::case_when(
      func == "NS" & is.na(Color) ~ "grey70",   # neutral colour for NS
      is.na(Color)                ~ "black",    # fallback for any unknown func
      TRUE                        ~ Color
    )
  )

color_values <- setNames(palette_df$Color, palette_df$func)

functional_differences_temperature_logfold %>% 
  ggplot(aes(x = diff, y = neglog10_fdr, color = func)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed") +
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed") +
  ylim(0, 15) +
  xlim(-0.6, 0.6) +
  ggrepel::geom_text_repel(
    data = subset(functional_differences_temperature_logfold, hit),
    aes(label = GIFT),
    size = 3, max.overlaps = 20, box.padding = 0.3, min.segment.length = 0
  ) +
  scale_color_manual(values = color_values) +
  labs(
    x = "Mean difference (Positive − Negative)",
    y = expression(-log[10]("FDR (BH)")),
    color = "Group",
    subtitle = paste0(
      "Thresholds: |diff| ≥ ", fc_threshold,
      ", FDR < ", fdr_threshold
    )
  ) +
  theme_minimal(base_size = 12)

```

**Note:** Aggregate temperature β coefficients to the functional group level by calculating mean β across all genomes carrying each functional trait. Identify the top 15 functional groups with strongest positive and negative associations, then create a forest plot showing functional group-level associations with credible intervals, colored by functional group.

```{r temperature_function_top_forest_18, warning=FALSE, message=FALSE, fig.height=7}
# Map genomes to functional groups (GIFT 3-letter codes)
genome_func_temp <- elements_response %>%
  dplyr::filter(value > 0) %>%
  dplyr::mutate(func = substr(GIFT, 1, 3)) %>%
  dplyr::distinct(genome, func)

# Summarise temperature β across genomes for each functional group
temp_func_summary <- genome_func_temp %>%
  dplyr::left_join(temp_ci %>% dplyr::select(genome, mean, q10, q90), by = "genome") %>%
  dplyr::group_by(func) %>%
  dplyr::summarise(
    n_genomes = sum(!is.na(mean)),
    mean_beta = mean(mean, na.rm = TRUE),
    q10_fun   = stats::quantile(mean, 0.10, na.rm = TRUE),
    q90_fun   = stats::quantile(mean, 0.90, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::filter(n_genomes >= 3)   # require at least 3 genomes per function


# 3) Take top 15 positive & top 15 negative functions ---------------------

N_top <- 15

temp_func_top <- dplyr::bind_rows(
  temp_func_summary %>%
    dplyr::slice_max(order_by = mean_beta, n = N_top, with_ties = FALSE),
  temp_func_summary %>%
    dplyr::slice_min(order_by = mean_beta, n = N_top, with_ties = FALSE)
) %>%
  dplyr::distinct(func, .keep_all = TRUE)

# 4) Add nice labels & colours from gift_colors --------------------------

temp_func_top <- temp_func_top %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::transmute(
        func        = Code_function,   # 3-letter code
        func_label  = legend,          # "B01 – Nucleic acid biosynthesis", etc.
        color       = Color
      ),
    by = "func"
  ) %>%
  dplyr::mutate(
    func_label = ifelse(is.na(func_label), func, func_label)
  )

func_cols_temp <- setNames(temp_func_top$color, temp_func_top$func_label)

# 5) Forest plot: functional groups for temperature ----------------------

ggplot(
  temp_func_top,
  aes(x = mean_beta,
      y = forcats::fct_reorder(func_label, mean_beta),
      xmin = q10_fun, xmax = q90_fun,
      colour = func_label)
) +
  geom_point(size = 2) +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = func_cols_temp, guide = "none") +
  theme_minimal(base_size = 10) +
  labs(
    x = "Temperature β (mean across genomes carrying the function)",
    y = "Functional trait (GIFT group)"
  )
```

## Association with devil + temperature interaction

**Note:** Summarize temperature:devil associations across taxonomic levels. Count and calculate percentages of genomes with positive and negative temperature:devil associations at phylum, order, family, and genus levels. This helps identify which taxonomic groups show the strongest associations with temperature:devil.

```{r association_temperature_devil_18, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
post_table$temperature_devil %>% table()
  
temp_phylum_perc <- post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  group_by(phylum) %>%
  summarise(positive = sum(temperature_devil == "Positive"),
            percent_positive = mean(temperature_devil == "Positive") * 100,
            negative = sum(temperature_devil == "Negative"),
            percent_negative = mean(temperature_devil == "Negative") * 100)

temp_phylum_perc %>% 
  arrange(-percent_positive)

temp_phylum_perc %>% 
  arrange(-percent_negative)

# Sub-phylum level

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(phylum=="p__Bacillota_A") %>% 
  group_by(order) %>%
  summarise(positive = sum(temperature_devil == "Positive"),
            percent_positive = mean(temperature_devil == "Positive") * 100,
            negative = sum(temperature_devil == "Negative"),
            percent_negative = mean(temperature_devil == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(order=="o__TANB77") %>% 
  group_by(family) %>%
  summarise(positive = sum(temperature_devil == "Positive"),
            percent_positive = mean(temperature_devil == "Positive") * 100,
            negative = sum(temperature_devil == "Negative"),
            percent_negative = mean(temperature_devil == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(family=="f__CAG-508") %>% 
  group_by(genus) %>%
  summarise(positive = sum(temperature_devil == "Positive"),
            percent_positive = mean(temperature_devil == "Positive") * 100,
            negative = sum(temperature_devil == "Negative"),
            percent_negative = mean(temperature_devil == "Negative") * 100) %>% 
  arrange(-percent_positive)
```
**Note:** Extract posterior mean and support values for the temperature:devil coefficient (β) for each genome. Create a scatter plot showing the relationship between the mean β coefficient and posterior support, colored by phylum. This visualization helps identify genomes with strong, well-supported associations with temperature:devil.

```{r beta_temperature_devil_18, warning=FALSE, comments="", message=FALSE}
temp_estimate <- getPostEstimate(hM = fit_model_18, parName = "Beta")$mean %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "temperature:devil") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
  dplyr::select(genome, mean)

temp_support <- getPostEstimate(hM = fit_model_18, parName = "Beta")$support %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "temperature:devil") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
  dplyr::select(genome, support)

temp_beta <- inner_join(temp_estimate, temp_support, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

temp_beta %>%
  mutate(significant = support >= support_threshold | support <= negsupport_threshold,
         support_plot = ifelse(mean < 0, 1 - support, support),
         phylum_sig = ifelse(significant & support_plot > 0.9, phylum, NA)) %>%
  ggplot(aes(x = mean, y = support_plot, colour = phylum_sig)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = phylum_colors, na.translate = FALSE) +
  geom_vline(xintercept = 0) +
  labs(x = "temperature:devil β coefficient",
       y = "Posterior probability (|β| > 0)",
       colour = "Phylum") +
  theme_minimal()
```
**Note:** Calculate 80% credible intervals (10th and 90th percentiles) for the temperature:devil β coefficient for each genome. Filter to genomes with credible non-zero effects (where the credible interval does not include zero) and visualize these with error bars, ordered by mean β value.

```{r temperature_devil_credible_intervals_18, warning=FALSE, comments="", message=FALSE}
temp_q <- getPostEstimate(hM = fit_model_18,
                          parName = "Beta",
                          q = c(0.1, 0.9))$q
idx_temp <- which(m$covNames == "temperature:devil")
n_genomes <- length(m$spNames)

temp_ci <- map_dfr(1:n_genomes, ~ {
  tibble(
    genome = m$spNames[.x],
    q10 = temp_q[1, idx_temp, .x],
    q90 = temp_q[2, idx_temp, .x]
  )
}) %>%
  right_join(temp_estimate, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

temp_ci %>%
  filter(q10 > 0 | q90 < 0) %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal() +
  labs(x = "temperature:devil β coefficient",
       y = "Genomes") +
  theme(legend.position = "none")
```

**Note:** Identify and visualize the top 30 genomes with the strongest positive and top 30 with the strongest negative credible temperature:devil associations. Create a forest plot showing these top genomes with their credible intervals, faceted by association direction (positive/negative), ordered by mean β coefficient and colored by phylum.

```{r temperature_devil_ci_top30_18, fig.height=10, fig.width=6}
# --- temperature:devil credible intervals: top 30 genomes ---

# Keep only genomes with credible non-zero temperature:devil effect
temp_ci_sig <- temp_ci %>%
  dplyr::filter(q10 > 0 | q90 < 0)

# 4. Take top 30 positive and top 30 negative genomes
N_top <- 30

temp_ci_top <- temp_ci_sig %>%
  dplyr::slice_max(order_by = mean, n = N_top, with_ties = FALSE) %>%
  dplyr::bind_rows(
    temp_ci_sig %>%
      dplyr::slice_min(order_by = mean, n = N_top, with_ties = FALSE)
  ) %>%
  dplyr::distinct(genome, .keep_all = TRUE) %>%
  dplyr::mutate(
    direction = ifelse(mean > 0, "Positive association", "Negative association"),
    genome    = forcats::fct_reorder(genome, mean)
  )

# 5. Plot credible intervals for these top genomes
temp_ci_top %>%
  ggplot2::ggplot(ggplot2::aes(x = mean,
                               y = genome,
                               xmin = q10, xmax = q90,
                               colour = phylum)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(width = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::scale_color_manual(values = phylum_colors) +
  ggplot2::facet_grid(direction ~ ., scales = "free_y", space = "free_y") +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::labs(
    x = "temperature:devil β coefficient",
    y = "Top genomes with credible temperature:devil effect"
  ) +
  ggplot2::theme(legend.position = "right")
```

---

### 2) temperature:devil – phylum-level summary of β

**Note:** Summarize temperature:devil β coefficients at the phylum level by calculating mean β and 80% credible intervals (10th and 90th percentiles) for each phylum. Visualize phylum-level associations with error bars to show the overall pattern of temperature:devil associations across major taxonomic groups.

```{r temperature_devil_phylum_ci_18, warning=FALSE, message=FALSE}
temp_phylum_ci <- temp_ci %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarise(
    mean_beta = mean(mean, na.rm = TRUE),
    q10       = stats::quantile(mean, 0.1, na.rm = TRUE),
    q90       = stats::quantile(mean, 0.9, na.rm = TRUE),
    .groups   = "drop"
  )

temp_phylum_ci %>%
  ggplot2::ggplot(ggplot2::aes(x = mean_beta,
                               y = forcats::fct_reorder(phylum, mean_beta),
                               xmin = q10, xmax = q90)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(width = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    x = "Mean temperature:devil β (per phylum)",
    y = "Phylum"
  )

```


**Note:** Compare functional element (GIFT) abundances between genomes with positive versus negative temperature:devil associations. For each functional element, calculate mean abundances in each group, compute differences, and perform t-tests. Adjust p-values using Benjamini-Hochberg FDR correction to identify functional elements that differ significantly between temperature:devil-associated groups.

```{r functional_differences_temperature_devil_18, warning=FALSE, comments="", message=FALSE}
functional_differences_temp_devil <- elements_response %>% 
  filter(temperature_devil != "Neutral") %>% 
  mutate(temp_devil_interaction = droplevels(temperature_devil)) %>%
  group_by(GIFT) %>%
  summarise(
    n_pos   = sum(temp_devil_interaction == "Positive"),
    n_neg   = sum(temp_devil_interaction == "Negative"),
    mean_pos = mean(value[temp_devil_interaction == "Positive"], na.rm = TRUE),
    mean_neg = mean(value[temp_devil_interaction == "Negative"], na.rm = TRUE),
    diff     = mean_pos - mean_neg,
    p        = if (n_pos > 1 && n_neg > 1 &&
                   stats::sd(value[temp_devil_interaction == "Positive"], na.rm = TRUE) > 0 &&
                   stats::sd(value[temp_devil_interaction == "Negative"], na.rm = TRUE) > 0) {
                stats::t.test(value ~ temp_devil_interaction, var.equal = FALSE)$p.value
              } else NA_real_
  , .groups = "drop") %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)
```


**Note:** Create a volcano plot visualizing functional differences between positive and negative temperature:devil-associated genomes. Plot mean difference (fold change) against -log10(FDR), with functional elements colored by their functional group. Highlight significant elements that meet both fold change and FDR thresholds, with labels for the most significant hits.

```{r temperature_devil_functions_plot_18, warning=FALSE, comments="", message=FALSE}
fc_threshold <- 0.2      # minimum absolute mean difference to call interesting
fdr_threshold <- 0.05    # BH-adjusted p-value cutoff

functional_differences_temp_devil_logfold <- functional_differences_temp_devil %>%
  mutate(func = substr(GIFT, 1, 3)) %>% 
  mutate(
    # guard against p_adj = 0 or NA
    p_adj_plot   = ifelse(is.na(p_adj) | p_adj == 0, .Machine$double.xmin, p_adj),
    neglog10_fdr = -log10(p_adj_plot),
    hit          = !is.na(p_adj) & p_adj < fdr_threshold & abs(diff) >= fc_threshold
  ) %>% 
  mutate(func = ifelse(hit, func, "NS")) %>%   # NS = non-significant / background
  filter(neglog10_fdr < 20)

# ---- Build a safe colour vector for all func levels ----
func_levels <- sort(unique(functional_differences_temp_devil_logfold$func))

palette_df <- tibble::tibble(func = func_levels) %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::select(func = Code_function, Color),
    by = "func"
  ) %>%
  dplyr::mutate(
    Color = dplyr::case_when(
      func == "NS" & is.na(Color) ~ "grey70",   # neutral colour for NS
      is.na(Color)                ~ "black",    # fallback for any unknown func
      TRUE                        ~ Color
    )
  )

color_values <- setNames(palette_df$Color, palette_df$func)

functional_differences_temp_devil_logfold %>% 
  ggplot(aes(x = diff, y = neglog10_fdr, color = func)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed") +
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed") +
  ylim(0, 15) +
  xlim(-0.6, 0.6) +
  ggrepel::geom_text_repel(
    data = subset(functional_differences_temp_devil_logfold, hit),
    aes(label = GIFT),
    size = 3, max.overlaps = 20, box.padding = 0.3, min.segment.length = 0
  ) +
  scale_color_manual(values = color_values) +
  labs(
    x = "Mean difference (Positive − Negative)",
    y = expression(-log[10]("FDR (BH)")),
    color = "Group",
    subtitle = paste0(
      "Thresholds: |diff| ≥ ", fc_threshold,
      ", FDR < ", fdr_threshold
    )
  ) +
  theme_minimal(base_size = 12)

```

**Note:** Aggregate temperature:devil β coefficients to the functional group level by calculating mean β across all genomes carrying each functional trait. Identify the top 15 functional groups with strongest positive and negative associations, then create a forest plot showing functional group-level associations with credible intervals, colored by functional group.

```{r temperature_devil_function_top_forest_18, warning=FALSE, message=FALSE, fig.height=7}
# Map genomes to functional groups (GIFT 3-letter codes)
genome_func_temp <- elements_response %>%
  dplyr::filter(value > 0) %>%
  dplyr::mutate(func = substr(GIFT, 1, 3)) %>%
  dplyr::distinct(genome, func)

# Summarise temperature:devil β across genomes for each functional group
temp_func_summary <- genome_func_temp %>%
  dplyr::left_join(temp_ci %>% dplyr::select(genome, mean, q10, q90), by = "genome") %>%
  dplyr::group_by(func) %>%
  dplyr::summarise(
    n_genomes = sum(!is.na(mean)),
    mean_beta = mean(mean, na.rm = TRUE),
    q10_fun   = stats::quantile(mean, 0.10, na.rm = TRUE),
    q90_fun   = stats::quantile(mean, 0.90, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::filter(n_genomes >= 3)   # require at least 3 genomes per function


# 3) Take top 15 positive & top 15 negative functions ---------------------

N_top <- 15

temp_func_top <- dplyr::bind_rows(
  temp_func_summary %>%
    dplyr::slice_max(order_by = mean_beta, n = N_top, with_ties = FALSE),
  temp_func_summary %>%
    dplyr::slice_min(order_by = mean_beta, n = N_top, with_ties = FALSE)
) %>%
  dplyr::distinct(func, .keep_all = TRUE)

# 4) Add nice labels & colours from gift_colors --------------------------

temp_func_top <- temp_func_top %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::transmute(
        func        = Code_function,   # 3-letter code
        func_label  = legend,          # "B01 – Nucleic acid biosynthesis", etc.
        color       = Color
      ),
    by = "func"
  ) %>%
  dplyr::mutate(
    func_label = ifelse(is.na(func_label), func, func_label)
  )

func_cols_temp <- setNames(temp_func_top$color, temp_func_top$func_label)

# 5) Forest plot: functional groups for temperature:devil ----------------------

ggplot(
  temp_func_top,
  aes(x = mean_beta,
      y = forcats::fct_reorder(func_label, mean_beta),
      xmin = q10_fun, xmax = q90_fun,
      colour = func_label)
) +
  geom_point(size = 2) +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = func_cols_temp, guide = "none") +
  theme_minimal(base_size = 10) +
  labs(
    x = "temperature:devil β (mean across genomes carrying the function)",
    y = "Functional trait (GIFT group)"
  )
```