> **Archived (former chapter 15).** Not in `_bookdown.yml`. See `archive/README.md`.

# Hmsc analysis
load("data/data.Rdata")

gift_colors <- read_tsv("data/gift_colors.tsv") %>% 
  mutate(legend=str_c(Code_function," - ",Function))
```

```{r parameters_hmsc_16}
#Parameters
nSamples=250
thin=1000
transient=nSamples * thin
support_threshold=0.9
negsupport_threshold=1-support_threshold
```

```{r load_model_16, warning=FALSE, comments="", message=FALSE, cache=TRUE}
#Load unfitted
load(file = "hmsc/model_tasmania")

#Load fitted (as a temporary file)
tmp <- tempfile(fileext = ".rds")
options(timeout = max(600, getOption("timeout"))) 
download.file("https://sid.erda.dk/share_redirect/Gh0WbM42c4/Hmsc_model_tasmania.rds", tmp, mode = "wb")
hpc_fit <- readRDS(tmp)
unlink(tmp)

importFromHPC <- hpc_fit$list
n_sp_fit <- ncol(importFromHPC[[1]][[1]]$Beta)
n_tasmania_samples <- sum(sample_metadata$island == "Tasmania")
if (length(m$spNames) != n_sp_fit) {
  stop(
    "model_tasmania: unfitted model (", length(m$spNames), " genomes) does not match fitted posterior (",
    n_sp_fit, " genomes). Re-run ch13 Tasmania setup and HPC fit.", call. = FALSE
  )
}
if (nrow(m$XData) != n_tasmania_samples) {
  stop(
    "model_tasmania: unfitted model has ", nrow(m$XData),
    " samples but Tasmania analysis expects ", n_tasmania_samples,
    ". Re-run ch13 Tasmania setup.", call. = FALSE
  )
}

postList = importFromHPC[1:4]
fit_model_16 = importPosteriorFromHPC(m, postList, nSamples, thin, transient)
```

```{r varpart_all_16, warning=FALSE, comments="", message=FALSE, cache=TRUE}
# Compute variance partitioning
varpart=computeVariancePartitioning(fit_model_16)

varpart$vals %>%
   as.data.frame() %>%
   rownames_to_column(var="variable") %>%
   pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
   mutate(variable=factor(variable)) %>%
   group_by(variable) %>%
   summarise(mean=mean(value)*100,sd=sd(value)*100) %>%
   tt()

preds = computePredictedValues(fit_model_16) 
MF = evaluateModelFit(hM=fit_model_16, predY=preds) 
hist(MF$R2, xlim = c(0,1), main=paste0("Mean = ", round(mean(MF$R2),2)))
```

```{r varpart_tree_plot_16, warning=FALSE, message=FALSE, fig.height=8, fig.width=7, eval = FALSE}
# 1) Build the phylogeny in the same species set/order as varpart
vp_tree_phylo <- genome_tree %>%
  keep.tip(tip = colnames(varpart$vals)) %>%
  force.ultrametric(method = "extend")

vp_tree <- ggtree(vp_tree_phylo, size = 0.3)

# 2) Phylum annotation in tree tip order
phylums_vp <- genome_metadata %>%
  filter(genome %in% vp_tree$data$label) %>%
  arrange(match(genome, vp_tree$data$label)) %>%
  mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
  column_to_rownames("genome") %>%
  select(phylum)

vp_tree <- gheatmap(
  vp_tree,
  phylums_vp,
  offset    = -0.2,
  width     = 0.1,
  colnames  = FALSE
) +
  scale_fill_manual(values = phylum_colors) +
  labs(fill = "Phylum")

# start a new fill scale for the variance bars
vp_tree <- vp_tree + new_scale_fill()

# 3) Long variance-partitioning table
varpart_long <- varpart$vals %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  pivot_longer(
    cols      = -variable,
    names_to  = "label",   # must be 'label' to match tree tip labels
    values_to = "value"
  ) %>%
  mutate(
    # reorder predictors in a sensible way for the legend
    variable = factor(
      variable,
      levels = rev(c(
        "devil", "temperature", "density",
        "development", "weight",
        "precipitation", "diversity",
        "logseqdepth",
        "Random: animal", "Random: site"
      ))
    )
  )

# 4) Add stacked bars with geom_fruit
vp_tree +
  geom_fruit(
    data   = varpart_long,
    geom   = geom_bar,
    mapping = aes(
      x    = value,
      y    = label,
      fill = variable
    ),
    stat        = "identity",
    orientation = "y",
    pwidth      = 2,
    offset      = 0.05
  ) +
  labs(fill = "Variable")

```
## Phylogenetic signal

```{r phylogenetic_signal_16, warning=FALSE, message=FALSE, eval = FALSE}
# Convert to coda object to inspect Rho (phylogenetic signal)
mpost <- convertToCodaObject(fit_model_16)

# Overall Rho posterior (phylogenetic signal on the species random effect)
rho_vec <- unlist(mpost$Rho)

# 90% credible interval and median
rho_summary <- quantile(rho_vec, probs = c(0.05, 0.5, 0.95))
rho_summary

# Optional: density plot for the appendix
plot(density(rho_vec),
     main = "Posterior of phylogenetic signal (Rho)",
     xlab = "Rho")
abline(v = rho_summary[2], col = "red", lty = 2)
```

```{r post_table_16, warning=FALSE, comments="", message=FALSE, cache=TRUE}
hmsc_tree <- genome_tree %>%
  keep.tip(., tip = m$spNames)

post_estimates <- getPostEstimate(hM = fit_model_16, parName = "Beta")$support %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
  mutate(trend = case_when(
    value >= support_threshold ~ "Positive",
    value <= negsupport_threshold ~ "Negative",
    TRUE ~ "Neutral"
  ))

post_table <- post_estimates %>%
  mutate(genome = factor(genome, levels = rev(hmsc_tree$tip.label))) %>%
  select(-trend) %>%
  mutate(value = case_when(
    value >= support_threshold ~ "Positive",
    value <= negsupport_threshold ~ "Negative",
    TRUE ~ "Neutral"
  )) %>%
  mutate(value = factor(value, levels = c("Positive", "Neutral", "Negative"))) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  select(-`(Intercept)`) %>%
  column_to_rownames(var = "genome")
```

```{r posteriors_16, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE, eval = FALSE}
# Basal ggtree
postestimates_tree <- hmsc_tree %>%
        force.ultrametric(.,method="extend") %>%
        ggtree(., size = 0.3)
        

# Phylums
phylums <- genome_metadata %>%
    filter(genome %in% hmsc_tree$tip.label) %>%
    arrange(match(genome, hmsc_tree$tip.label)) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome") %>%
    select(phylum)


#Add phylum colors next to the tree tips
postestimates_tree <- gheatmap(postestimates_tree, phylums, offset=-0.2, width=0.1, colnames=FALSE) +
      scale_fill_manual(values=phylum_colors)+
      labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
postestimates_tree <- postestimates_tree + new_scale_fill()

# Add posterior significant heatmap

postestimates_tree <- gheatmap(postestimates_tree, post_table, offset=0, width=0.5, colnames=TRUE, colnames_position="top",colnames_angle=90, colnames_offset_y=1, hjust=0) +
        scale_fill_manual(values=c("#be3e2b","#f4f4f4","#b2b530"))+
        labs(fill="Trend") 

postestimates_tree +
        vexpand(.30, 1)  # expand top 

```

```{r congruence_devil_temperature_taxa_16, warning=FALSE, message=FALSE}
# Long table of trends for devil and temperature per genome
devil_temp_trends <- post_table %>%
  rownames_to_column("genome") %>%
  select(genome, devil, temperature) %>%
  left_join(genome_metadata, by = "genome")

# Overall contingency of trends
devil_temp_tab <- devil_temp_trends %>%
  count(devil, temperature) %>%
  arrange(desc(n))

devil_temp_tab %>% tt()

# Phylum-level congruence: how many genomes in each combo
devil_temp_phylum <- devil_temp_trends %>%
  group_by(phylum, devil, temperature) %>%
  summarise(n = n(), .groups = "drop")

# Keep only non-zero combos (for cleaner plots)
devil_temp_phylum %>%
  filter(n > 0) %>%
  ggplot(aes(x = devil, y = temperature, fill = n)) +
  geom_tile() +
  facet_wrap(~ phylum) +
  scale_fill_gradient(low = "lightgrey", high = "blue") +
  labs(
    x = "Devil association",
    y = "Temperature association",
    fill = "Number of genomes",
    title = "Taxonomic congruence of devil vs temperature responses"
  ) +
  theme_minimal(base_size = 8)
```

```{r functional_response_16, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
elements_response <- genome_gifts[rownames(post_table),colnames(genome_gifts) %in% GIFT_db$Code_bundle] %>%
  to.elements(.,GIFT_db=GIFT_db) %>% 
  as.data.frame() %>% 
  rownames_to_column(var="genome") %>% 
  pivot_longer(!genome,names_to="GIFT", values_to="value") %>% 
  left_join(post_table %>% rownames_to_column(var = "genome"), by="genome")
```

```{r congruence_devil_temperature_functional_16, warning=FALSE, message=FALSE}
# elements_response has: genome, GIFT, value, and trend columns (devil, temperature, ...)
# Focus on non-neutral responses for both variables
elem_dev_temp <- elements_response %>%
  filter(devil != "Neutral",
         temperature != "Neutral") %>%
  mutate(
    devil      = droplevels(devil),
    temperature = droplevels(temperature)
  )

# Summarise at the element level: mean value per devil/temperature combo
func_congruence_elem <- elem_dev_temp %>%
  group_by(GIFT, devil, temperature) %>%
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    n        = n(),
    .groups  = "drop"
  )

# Collapse to functional group (Code_function)
func_congruence_group <- func_congruence_elem %>%
  left_join(
    GIFT_db %>% distinct(Code_element, Code_function, Function),
    by = c("GIFT" = "Code_element")
  ) %>%
  group_by(Code_function, devil, temperature) %>%
  summarise(
    mean_val = mean(mean_val, na.rm = TRUE),
    .groups  = "drop"
  )

# For readability, keep only functional groups that show some variation
interesting_funcs <- func_congruence_group %>%
  group_by(Code_function) %>%
  summarise(range_val = max(mean_val) - min(mean_val), .groups = "drop") %>%
  slice_max(order_by = range_val, n = 12) %>%
  pull(Code_function)

func_congruence_group %>%
  filter(Code_function %in% interesting_funcs) %>%
  ggplot(aes(x = devil, y = temperature, fill = mean_val)) +
  geom_tile() +
  facet_wrap(~ Code_function) +
  scale_fill_gradient2() +
  labs(
    x = "Devil association",
    y = "Temperature association",
    fill = "Mean element signal",
    title = "Functional congruence of devil vs temperature responses"
  ) +
  theme_minimal(base_size = 8)
```

## Association with devil

```{r association_devil_16, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
post_table$devil %>% table()
  
devil_phylum_perc <- post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  group_by(phylum) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100)

devil_phylum_perc %>% 
  arrange(-percent_positive)

devil_phylum_perc %>% 
  arrange(-percent_negative)

# Sub-phylum level

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(phylum=="p__Bacillota_A") %>% 
  group_by(order) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(order=="o__TANB77") %>% 
  group_by(family) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(family=="f__CAG-508") %>% 
  group_by(genus) %>%
  summarise(positive = sum(devil == "Positive"),
            percent_positive = mean(devil == "Positive") * 100,
            negative = sum(devil == "Negative"),
            percent_negative = mean(devil == "Negative") * 100) %>% 
  arrange(-percent_positive)
```
```{r beta_devil_16, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
# Mean beta for 'devil' per genome
devil_estimate <- getPostEstimate(hM = fit_model_16, parName = "Beta")$mean %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "devil") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
  dplyr::select(genome, mean)

# Posterior support for 'devil' per genome
devil_support <- getPostEstimate(hM = fit_model_16, parName = "Beta")$support %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "devil") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
  dplyr::select(genome, support)

devil_beta <- inner_join(devil_estimate, devil_support, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

devil_beta %>%
  mutate(
    significant   = support >= support_threshold | support <= negsupport_threshold,
    support_plot  = ifelse(mean < 0, 1 - support, support), # support for the sign of beta
    phylum_sig    = ifelse(significant & support_plot > 0.9, phylum, NA)
  ) %>%
  ggplot(aes(x = mean, y = support_plot, colour = phylum_sig)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = phylum_colors, na.translate = FALSE) +
  geom_vline(xintercept = 0) +
  labs(
    x = "Devil β coefficient",
    y = "Posterior probability (|β| > 0)",
    colour = "Phylum"
  ) +
  theme_minimal()

```
```{r devil_credible_intervals_16, warning=FALSE, comments="", message=FALSE}
# Credible intervals for 'devil'
devil_q <- getPostEstimate(hM = fit_model_16,
                           parName = "Beta",
                           q = c(0.1, 0.9))$q

idx_devil <- which(m$covNames == "devil")
n_genomes <- length(m$spNames)

devil_ci <- map_dfr(1:n_genomes, ~ {
  tibble(
    genome = m$spNames[.x],
    q10 = devil_q[1, idx_devil, .x],
    q90 = devil_q[2, idx_devil, .x]
  )
}) %>%
  right_join(devil_estimate, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

# Genomes with credible non-zero devil effect
devil_ci %>%
  filter(q10 > 0 | q90 < 0) %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal() +
  labs(x = "Devil β coefficient",
       y = "Genomes") +
  theme(legend.position = "none")
```
```{r devil_topgenomes_16, warning=FALSE, comments="", message=FALSE}
# Keep only genomes with credible non-zero effect
devil_ci_sig <- devil_ci %>%
  filter(q10 > 0 | q90 < 0)

# 4. Take top N positive and top N negative genomes
N_top <- 30  

devil_ci_top <- devil_ci_sig %>%
  slice_max(order_by = mean, n = N_top, with_ties = FALSE) %>%
  bind_rows(
    devil_ci_sig %>%
      slice_min(order_by = mean, n = N_top, with_ties = FALSE)
  ) %>%
  distinct(genome, .keep_all = TRUE)

# 5. Plot
devil_ci_top %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal(base_size = 9) +
  labs(
    x = "Devil β coefficient",
    y = "Top genomes with credible devil effect"
  ) +
  theme(legend.position = "right")

```
```{r devil_phylum_ci_16, warning=FALSE, comments="", message=FALSE}
devil_phylum_ci <- devil_ci %>%
  group_by(phylum) %>%
  summarise(
    mean_beta = mean(mean),
    q10 = quantile(mean, 0.1),
    q90 = quantile(mean, 0.9),
    .groups = "drop"
  )

devil_phylum_ci %>%
  ggplot(aes(x = mean_beta,
             y = forcats::fct_reorder(phylum, mean_beta),
             xmin = q10, xmax = q90)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  theme_minimal() +
  labs(x = "Mean devil β (per phylum)",
       y = "Phylum")
```
```{r devil_overview_16, warning=FALSE, comments="", message=FALSE}
devil_positive <- devil_ci %>%
  filter(q10 > 0) %>%
  arrange(-mean) %>%
  dplyr::select(genome, phylum, class, order, family, genus, species, mean, q10, q90)

devil_negative <- devil_ci %>%
  filter(q90 < 0) %>%
  arrange(mean) %>%
  dplyr::select(genome, phylum, class, order, family, genus, species, mean, q10, q90)

devil_positive %>% tt()  # tinytable
devil_negative %>% tt()
```

```{r devil_functions_16, warning=FALSE, comments="", message=FALSE}
functional_differences_devil <- elements_response %>% 
  filter(devil !="Neutral") %>% 
  mutate(devil = droplevels(devil)) %>%
  group_by(GIFT) %>%
  summarise(
    n_pos   = sum(devil == "Positive"),
    n_neg   = sum(devil == "Negative"),
    mean_pos = mean(value[devil == "Positive"], na.rm = TRUE),
    mean_neg = mean(value[devil == "Negative"], na.rm = TRUE),
    diff     = mean_pos - mean_neg,
    p        = if (n_pos > 1 && n_neg > 1 &&
                   stats::sd(value[devil == "Positive"], na.rm = TRUE) > 0 &&
                   stats::sd(value[devil == "Negative"], na.rm = TRUE) > 0) {
                stats::t.test(value ~ devil, var.equal = FALSE)$p.value
              } else NA_real_
  , .groups = "drop") %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)
```


```{r devil_functions_plot_16, warning=FALSE, comments="", message=FALSE}
fc_threshold <- 0.2      # minimum absolute mean difference to call interesting
fdr_threshold <- 0.05    # BH-adjusted p-value cutoff

functional_differences_devil_logfold <- functional_differences_devil %>%
  mutate(func = substr(GIFT, 1, 3)) %>% 
  mutate(
    # guard against p_adj = 0 or NA
    p_adj_plot = ifelse(is.na(p_adj) | p_adj == 0, .Machine$double.xmin, p_adj),
    neglog10_fdr = -log10(p_adj_plot),
    hit = !is.na(p_adj) & p_adj < fdr_threshold & abs(diff) >= fc_threshold
  ) %>% 
  mutate(func = ifelse(hit, func, "NS")) %>% 
  filter(neglog10_fdr<20)

palette_df <- functional_differences_devil_logfold %>%
  dplyr::distinct(func) %>%
  dplyr::left_join(
    gift_colors %>% dplyr::select(func = Code_function, Color),
    by = "func"
  ) %>%
  dplyr::mutate(
    Color = dplyr::case_when(
      func == "NS" & is.na(Color) ~ "grey70",
      is.na(Color)                ~ "#999999",
      TRUE                        ~ Color
    )
  )

color_values <- setNames(palette_df$Color, palette_df$func)

functional_differences_devil_logfold %>% 
  ggplot(aes(x = diff, y = neglog10_fdr, color=func)) +
  geom_point(aes(color = func), alpha = 0.8, size = 2) +
  geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed") +
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed") +
  ylim(0,15)+
  xlim(-0.6,0.6)+
  ggrepel::geom_text_repel(
    data = subset(functional_differences_devil_logfold, hit),
    aes(label = GIFT),
    size = 3, max.overlaps = 20, box.padding = 0.3, min.segment.length = 0
  ) +
  scale_color_manual(values = color_values, breaks = names(color_values)) +
  labs(
    x = "Mean difference (Positive − Negative)",
    y = expression(-log[10]("FDR (BH)")),
    color = "Group",
    subtitle = paste0("Thresholds: |diff| ≥ ", fc_threshold, ", FDR < ", fdr_threshold)
  ) +
  theme_minimal(base_size = 12)
```

```{r devil_function_top_forest_16, warning=FALSE, message=FALSE, fig.height=7}
# Map genomes to functional groups (GIFT 3-letter "func" code)
genome_func <- elements_response %>%
  dplyr::filter(value > 0) %>%
  dplyr::mutate(func = substr(GIFT, 1, 3)) %>%
  dplyr::distinct(genome, func)

# Summarise devil β across genomes for each functional group
devil_func_summary <- genome_func %>%
  dplyr::left_join(devil_ci %>% dplyr::select(genome, mean, q10, q90), by = "genome") %>%
  dplyr::group_by(func) %>%
  dplyr::summarise(
    n_genomes = sum(!is.na(mean)),
    mean_beta = mean(mean, na.rm = TRUE),
    q10_fun   = stats::quantile(mean, 0.10, na.rm = TRUE),
    q90_fun   = stats::quantile(mean, 0.90, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::filter(n_genomes >= 3)

# Take top 15 positive & top 15 negative functions
N_top <- 15

devil_func_top <- dplyr::bind_rows(
  devil_func_summary %>%
    dplyr::slice_max(order_by = mean_beta, n = N_top, with_ties = FALSE),
  devil_func_summary %>%
    dplyr::slice_min(order_by = mean_beta, n = N_top, with_ties = FALSE)
) %>%
  dplyr::distinct(func, .keep_all = TRUE)

# Add nice labels & colours from gift_colors
devil_func_top <- devil_func_top %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::transmute(
        func        = Code_function,
        func_label  = legend,
        color       = Color
      ),
    by = "func"
  ) %>%
  dplyr::mutate(
    func_label = ifelse(is.na(func_label), func, func_label)
  )

func_cols <- setNames(devil_func_top$color, devil_func_top$func_label)

# Forest plot: functions instead of genomes
ggplot(
  devil_func_top,
  aes(x = mean_beta,
      y = forcats::fct_reorder(func_label, mean_beta),
      xmin = q10_fun, xmax = q90_fun,
      colour = func_label)
) +
  geom_point(size = 2) +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = func_cols, guide = "none") +
  theme_minimal(base_size = 10) +
  labs(
    x = "Devil β (mean across genomes carrying the function)",
    y = "Functional trait (GIFT group)"
  )

```


## Association with temperature

```{r association_temperature_16, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10), cache=TRUE}
post_table$temperature %>% table()
  
temp_phylum_perc <- post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  group_by(phylum) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100)

temp_phylum_perc %>% 
  arrange(-percent_positive)

temp_phylum_perc %>% 
  arrange(-percent_negative)

# Sub-phylum level

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(phylum=="p__Bacillota_A") %>% 
  group_by(order) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(order=="o__TANB77") %>% 
  group_by(family) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100) %>% 
  arrange(-percent_positive)

post_table %>%
  rownames_to_column(var = "genome") %>%
  left_join(genome_metadata, by = "genome") %>%
  filter(family=="f__CAG-508") %>% 
  group_by(genus) %>%
  summarise(positive = sum(temperature == "Positive"),
            percent_positive = mean(temperature == "Positive") * 100,
            negative = sum(temperature == "Negative"),
            percent_negative = mean(temperature == "Negative") * 100) %>% 
  arrange(-percent_positive)
```
```{r beta_temperature_16, warning=FALSE, comments="", message=FALSE}
temp_estimate <- getPostEstimate(hM = fit_model_16, parName = "Beta")$mean %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "temperature") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
  dplyr::select(genome, mean)

temp_support <- getPostEstimate(hM = fit_model_16, parName = "Beta")$support %>%
  as.data.frame() %>%
  mutate(variable = m$covNames) %>%
  filter(variable == "temperature") %>%
  pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
  dplyr::select(genome, support)

temp_beta <- inner_join(temp_estimate, temp_support, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

temp_beta %>%
  mutate(significant = support >= support_threshold | support <= negsupport_threshold,
         support_plot = ifelse(mean < 0, 1 - support, support),
         phylum_sig = ifelse(significant & support_plot > 0.9, phylum, NA)) %>%
  ggplot(aes(x = mean, y = support_plot, colour = phylum_sig)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = phylum_colors, na.translate = FALSE) +
  geom_vline(xintercept = 0) +
  labs(x = "Temperature β coefficient",
       y = "Posterior probability (|β| > 0)",
       colour = "Phylum") +
  theme_minimal()
```
```{r temperature_credible_intervals_16, warning=FALSE, comments="", message=FALSE}
temp_q <- getPostEstimate(hM = fit_model_16,
                          parName = "Beta",
                          q = c(0.1, 0.9))$q
idx_temp <- which(m$covNames == "temperature")
n_genomes <- length(m$spNames)

temp_ci <- map_dfr(1:n_genomes, ~ {
  tibble(
    genome = m$spNames[.x],
    q10 = temp_q[1, idx_temp, .x],
    q90 = temp_q[2, idx_temp, .x]
  )
}) %>%
  right_join(temp_estimate, by = "genome") %>%
  left_join(genome_metadata, by = "genome")

temp_ci %>%
  filter(q10 > 0 | q90 < 0) %>%
  ggplot(aes(x = mean,
             y = forcats::fct_reorder(genome, mean),
             xmin = q10, xmax = q90,
             colour = phylum)) +
  geom_point() +
  geom_errorbar() +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = phylum_colors) +
  theme_minimal() +
  labs(x = "Temperature β coefficient",
       y = "Genomes") +
  theme(legend.position = "none")
```

```{r temperature_ci_top30_16, fig.height=10, fig.width=6}
# --- Temperature credible intervals: top 30 genomes ---

# Keep only genomes with credible non-zero temperature effect
temp_ci_sig <- temp_ci %>%
  dplyr::filter(q10 > 0 | q90 < 0)

# 4. Take top 30 positive and top 30 negative genomes
N_top <- 30

temp_ci_top <- temp_ci_sig %>%
  dplyr::slice_max(order_by = mean, n = N_top, with_ties = FALSE) %>%
  dplyr::bind_rows(
    temp_ci_sig %>%
      dplyr::slice_min(order_by = mean, n = N_top, with_ties = FALSE)
  ) %>%
  dplyr::distinct(genome, .keep_all = TRUE) %>%
  dplyr::mutate(
    direction = ifelse(mean > 0, "Positive association", "Negative association"),
    genome    = forcats::fct_reorder(genome, mean)
  )

# 5. Plot credible intervals for these top genomes
temp_ci_top %>%
  ggplot2::ggplot(ggplot2::aes(x = mean,
                               y = genome,
                               xmin = q10, xmax = q90,
                               colour = phylum)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(width = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::scale_color_manual(values = phylum_colors) +
  ggplot2::facet_grid(direction ~ ., scales = "free_y", space = "free_y") +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::labs(
    x = "Temperature β coefficient",
    y = "Top genomes with credible temperature effect"
  ) +
  ggplot2::theme(legend.position = "right")
```

---

### 2) Temperature – phylum-level summary of β

```{r temperature_phylum_ci_16, warning=FALSE, message=FALSE}
temp_phylum_ci <- temp_ci %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarise(
    mean_beta = mean(mean, na.rm = TRUE),
    q10       = stats::quantile(mean, 0.1, na.rm = TRUE),
    q90       = stats::quantile(mean, 0.9, na.rm = TRUE),
    .groups   = "drop"
  )

temp_phylum_ci %>%
  ggplot2::ggplot(ggplot2::aes(x = mean_beta,
                               y = forcats::fct_reorder(phylum, mean_beta),
                               xmin = q10, xmax = q90)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(width = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    x = "Mean temperature β (per phylum)",
    y = "Phylum"
  )

```


```{r functional_differences_temperature_16, warning=FALSE, comments="", message=FALSE}
functional_differences_temperature <- elements_response %>% 
  filter(temperature !="Neutral") %>% 
  mutate(temperature = droplevels(temperature)) %>%
  group_by(GIFT) %>%
  summarise(
    n_pos   = sum(temperature == "Positive"),
    n_neg   = sum(temperature == "Negative"),
    mean_pos = mean(value[temperature == "Positive"], na.rm = TRUE),
    mean_neg = mean(value[temperature == "Negative"], na.rm = TRUE),
    diff     = mean_pos - mean_neg,
    p        = if (n_pos > 1 && n_neg > 1 &&
                   stats::sd(value[temperature == "Positive"], na.rm = TRUE) > 0 &&
                   stats::sd(value[temperature == "Negative"], na.rm = TRUE) > 0) {
                stats::t.test(value ~ temperature, var.equal = FALSE)$p.value
              } else NA_real_
  , .groups = "drop") %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)
```


```{r temperature_functions_plot_16, warning=FALSE, comments="", message=FALSE}
fc_threshold <- 0.2      # minimum absolute mean difference to call interesting
fdr_threshold <- 0.05    # BH-adjusted p-value cutoff

functional_differences_temperature_logfold <- functional_differences_temperature %>%
  mutate(func = substr(GIFT, 1, 3)) %>% 
  mutate(
    # guard against p_adj = 0 or NA
    p_adj_plot   = ifelse(is.na(p_adj) | p_adj == 0, .Machine$double.xmin, p_adj),
    neglog10_fdr = -log10(p_adj_plot),
    hit          = !is.na(p_adj) & p_adj < fdr_threshold & abs(diff) >= fc_threshold
  ) %>% 
  mutate(func = ifelse(hit, func, "NS")) %>%   # NS = non-significant / background
  filter(neglog10_fdr < 20)

# ---- Build a safe colour vector for all func levels ----
func_levels <- sort(unique(functional_differences_temperature_logfold$func))

palette_df <- tibble::tibble(func = func_levels) %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::select(func = Code_function, Color),
    by = "func"
  ) %>%
  dplyr::mutate(
    Color = dplyr::case_when(
      func == "NS" & is.na(Color) ~ "grey70",   # neutral colour for NS
      is.na(Color)                ~ "black",    # fallback for any unknown func
      TRUE                        ~ Color
    )
  )

color_values <- setNames(palette_df$Color, palette_df$func)

functional_differences_temperature_logfold %>% 
  ggplot(aes(x = diff, y = neglog10_fdr, color = func)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed") +
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed") +
  ylim(0, 15) +
  xlim(-0.6, 0.6) +
  ggrepel::geom_text_repel(
    data = subset(functional_differences_temperature_logfold, hit),
    aes(label = GIFT),
    size = 3, max.overlaps = 20, box.padding = 0.3, min.segment.length = 0
  ) +
  scale_color_manual(values = color_values) +
  labs(
    x = "Mean difference (Positive − Negative)",
    y = expression(-log[10]("FDR (BH)")),
    color = "Group",
    subtitle = paste0(
      "Thresholds: |diff| ≥ ", fc_threshold,
      ", FDR < ", fdr_threshold
    )
  ) +
  theme_minimal(base_size = 12)

```

```{r temperature_function_top_forest_16, warning=FALSE, message=FALSE, fig.height=7}
# Map genomes to functional groups (GIFT 3-letter codes)
genome_func_temp <- elements_response %>%
  dplyr::filter(value > 0) %>%
  dplyr::mutate(func = substr(GIFT, 1, 3)) %>%
  dplyr::distinct(genome, func)

# Summarise temperature β across genomes for each functional group
temp_func_summary <- genome_func_temp %>%
  dplyr::left_join(temp_ci %>% dplyr::select(genome, mean, q10, q90), by = "genome") %>%
  dplyr::group_by(func) %>%
  dplyr::summarise(
    n_genomes = sum(!is.na(mean)),
    mean_beta = mean(mean, na.rm = TRUE),
    q10_fun   = stats::quantile(mean, 0.10, na.rm = TRUE),
    q90_fun   = stats::quantile(mean, 0.90, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::filter(n_genomes >= 3)

# Take top 15 positive & top 15 negative functions
N_top <- 15

temp_func_top <- dplyr::bind_rows(
  temp_func_summary %>%
    dplyr::slice_max(order_by = mean_beta, n = N_top, with_ties = FALSE),
  temp_func_summary %>%
    dplyr::slice_min(order_by = mean_beta, n = N_top, with_ties = FALSE)
) %>%
  dplyr::distinct(func, .keep_all = TRUE)

# Add nice labels & colours from gift_colors
temp_func_top <- temp_func_top %>%
  dplyr::left_join(
    gift_colors %>%
      dplyr::transmute(
        func        = Code_function,
        func_label  = legend,
        color       = Color
      ),
    by = "func"
  ) %>%
  dplyr::mutate(
    func_label = ifelse(is.na(func_label), func, func_label)
  )

func_cols_temp <- setNames(temp_func_top$color, temp_func_top$func_label)

# Forest plot: functional groups for temperature

ggplot(
  temp_func_top,
  aes(x = mean_beta,
      y = forcats::fct_reorder(func_label, mean_beta),
      xmin = q10_fun, xmax = q90_fun,
      colour = func_label)
) +
  geom_point(size = 2) +
  geom_errorbar(width = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values = func_cols_temp, guide = "none") +
  theme_minimal(base_size = 10) +
  labs(
    x = "Temperature β (mean across genomes carrying the function)",
    y = "Functional trait (GIFT group)"
  )
```